# Final Data Preprocessing for Visualization

This section creates the final country-level dataset used in the redesigned visualization. It aggregates export value and export quantity separately, joins them by reporting economy, and exports the processed CSV for reproducible plotting.

## Aggregate Export Value and Quantity by Country
```{r}
#| label: data-preprocessing-aggregate

value_w32 <- value_analysis_2024 |>
  group_by(country = reporter_name, geo_region) |>
  summarise(
    export_value = sum(export_value_usd_millions, na.rm = TRUE),
    .groups = "drop"
  )

value_w32

quantity_w32 <- quantity_2024 |>
  filter(country != "World") |>
  group_by(country) |>
  summarise(
    export_quantity = sum(quantity, na.rm = TRUE),
    .groups = "drop"
  )
```

## Join Value and Quantity Data
```{r}
#| label: data-preprocessing-join
combined_country_data <- value_w32 |> 
  left_join(quantity_w32, by = "country") |> 
  select(country, geo_region, export_value, export_quantity)

combined_country_data

sorted_values <- combined_country_data |>
  arrange(desc(export_value))

sorted_values
```

The WTO workbook includes aggregate product rows such as `AoA HS proxy`, `All W32 groups`, and `Other AoA`. These rows are useful as official reference totals, but they should not be summed together with individual product groups because doing so would double-count export activity.

For export value, we aggregate the cleaned `value_analysis_2024` dataset by reporting economy and region. This gives each reporting economy's total export value across the individual W32 product groups.

For export quantity, we use the cleaned `quantity_2024` dataset, where `AoA HS proxy` and product categories measured in USD rather than tonnes have already been removed. In this section, the `World` row is excluded before aggregating quantity by reporting economy.

The value and quantity summaries are then joined by reporting economy to produce the final country-level dataset used in the lollipop visualization.

## Export Processed Visualization Dataset

```{r}
#| label: export-processed-data

if (!dir.exists(export_dir)) {
  dir.create(export_dir, recursive = TRUE)
  message("Created new directory: ", export_dir)
}

csv_filename <- paste0(analysis_year, "_value_and_quantity_export.csv")
export_path <- file.path(export_dir, csv_filename)
write_csv(combined_country_data, export_path)

cat("Success! Processed data successfully exported to:", export_path, "\n")
```

For reproducibility, the preprocessed data has been exported and stored into the `data/processed` directory. 

