# Bubble Visualisation

## Preprocessed Data
```{r}
#| label: fetch-preprocessed-data
#| message: false
preprocessed_data <- read_csv(export_path)
```

## Data Preparation for Bubble Charts

```{r}
#| label: prepare-bubble-data
#| message: false

bubble_country_data <- preprocessed_data |>
  filter(
    str_squish(country) != "World",
    !is.na(export_value),
    !is.na(export_quantity),
    export_value > 0,
    export_quantity > 0,
    !is.na(geo_region)
  ) |>
  mutate(
    # Source units: value in USD millions, quantity in thousand tonnes.
    export_value_usd_billions = export_value / 1000,
    export_quantity_million_tonnes = export_quantity / 1000,
    # Unit price: (USD millions * 1000) / thousand tonnes = USD per tonne.
    usd_per_tonne = (export_value * 1000) / export_quantity
  )

bubble_region_palette <- region_color_palette(bubble_country_data$geo_region)

bubble_region_rest_palette <- c(bubble_region_palette, "Rest of World" = "grey55")

bubble_theme <- theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
    plot.margin = margin(10, 20, 16, 10)
  )

glimpse(bubble_country_data)
```

All of the bubble charts start from this cleaned country table, which keeps only
rows with a positive value and volume. The charts that group the data (by region,
or into a "rest" bubble) build on this table. Shared styling objects keep the
figures looking consistent.

## Regional Overview (Value vs Volume)

```{r}
#| label: bubble-chart-regional-overview
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 12

bubble_region_summary <- bubble_country_data |>
  group_by(geo_region) |>
  summarise(
    reporting_countries = n(),
    export_value = sum(export_value, na.rm = TRUE),
    export_quantity = sum(export_quantity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    export_value_usd_billions = export_value / 1000,
    export_quantity_million_tonnes = export_quantity / 1000,
    usd_per_tonne = (export_value * 1000) / export_quantity
  )

gg_bubble_region <- ggplot(
  bubble_region_summary,
  aes(
    x = export_quantity_million_tonnes,
    y = export_value_usd_billions,
    size = usd_per_tonne,
    color = geo_region
  )
) +
  geom_point(alpha = 0.85) +
  ggrepel::geom_text_repel(
    aes(label = geo_region),
    size = 4,
    color = "grey20",
    show.legend = FALSE,
    box.padding = 0.7,
    min.segment.length = 0
  ) +
  scale_size_area(
    max_size = 24,
    name = "Unit price (US$ / tonne)",
    labels = scales::label_dollar(prefix = "US$")
  ) +
  scale_color_manual(
    values = bubble_region_palette,
    name = "Geographic Region",
    guide = guide_legend(override.aes = list(size = 5))
  ) +
  scale_x_continuous(
    labels = scales::label_number(suffix = " Mt"),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(prefix = "US$", suffix = "B"),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  labs(
    x = "Total Agricultural Export Volume (million tonnes)",
    y = "Total Agricultural Export Value (US$ billions)",
    title = "Agricultural Export Value vs Volume by Region in 2024",
    subtitle = str_wrap(
      "One bubble per region. Bubble size shows average value earned per tonne shipped.",
      width = 95
    ),
    caption = str_wrap(
      "Aggregated across all reporting economies; the seven regions together account for 100% of recorded trade. European Union is treated as a single reporting economy within Europe.",
      width = 110
    )
  ) +
  bubble_theme

gg_bubble_region
```

## Top 3 Economies and Rest per Region

```{r}
#| label: prepare-bubble-top3-rest
#| message: false

# Rank economies within each region by export value (consistent with the
# lollipop ranking) so we can separate the top 3 from the rest.
bubble_ranked <- bubble_country_data |>
  group_by(geo_region) |>
  mutate(value_rank = row_number(desc(export_value))) |>
  ungroup()

# Top 3 economies per region, kept as individual bubbles.
bubble_top3 <- bubble_ranked |>
  filter(value_rank <= 3) |>
  transmute(
    geo_region,
    entity = country,
    entity_type = "Top-3 economy",
    export_value,
    export_quantity
  )

# Everyone ranked 4th or lower in a region, collapsed into one bubble.
bubble_rest <- bubble_ranked |>
  filter(value_rank > 3) |>
  group_by(geo_region) |>
  summarise(
    rest_countries = n(),
    export_value = sum(export_value, na.rm = TRUE),
    export_quantity = sum(export_quantity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  transmute(
    geo_region,
    entity = str_c("Rest of ", geo_region, " (", rest_countries, ")"),
    entity_type = "Rest of region",
    export_value,
    export_quantity
  )

bubble_plot_data <- bind_rows(bubble_top3, bubble_rest) |>
  mutate(
    export_value_usd_billions = export_value / 1000,
    export_quantity_million_tonnes = export_quantity / 1000,
    usd_per_tonne = (export_value * 1000) / export_quantity,
    entity_type = factor(
      entity_type,
      levels = c("Top-3 economy", "Rest of region")
    )
  )
```

```{r}
#| label: bubble-chart-top3-rest
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 14

gg_bubble_top3_rest <- ggplot(
  bubble_plot_data,
  aes(
    x = export_quantity_million_tonnes,
    y = export_value_usd_billions,
    size = usd_per_tonne,
    color = geo_region,
    shape = entity_type
  )
) +
  geom_point(alpha = 0.85, stroke = 1) +
  ggrepel::geom_text_repel(
    aes(label = entity),
    size = 3.2,
    color = "grey20",
    show.legend = FALSE,
    max.overlaps = Inf,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0,
    segment.color = "grey70"
  ) +
  # Solid circles for individual top-3 economies; hollow circles for the
  # aggregated "Rest of region" bubbles.
  scale_shape_manual(
    values = c("Top-3 economy" = 16, "Rest of region" = 1),
    name = "Bubble type"
  ) +
  scale_size_area(
    max_size = 20,
    name = "Unit price (US$ / tonne)",
    labels = scales::label_dollar(prefix = "US$")
  ) +
  scale_color_manual(
    values = bubble_region_palette,
    name = "Geographic Region",
    guide = guide_legend(override.aes = list(size = 5, shape = 16))
  ) +
  scale_x_log10(
    labels = scales::label_number(suffix = " Mt"),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  scale_y_log10(
    labels = scales::label_dollar(prefix = "US$", suffix = "B"),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  labs(
    x = "Agricultural Export Volume (million tonnes, log scale)",
    y = "Agricultural Export Value (US$ billions, log scale)",
    title = "Agricultural Export Value vs Volume in 2024: Top 3 Economies per Region",
    subtitle = str_wrap(
      "Solid bubbles are each region's three largest exporters by value; hollow bubbles combine the remaining economies in that region. Bubble size shows average unit price.",
      width = 100
    ),
    caption = str_wrap(
      "Together the bubbles account for 100% of recorded trade. Both axes are log-scaled because export size spans several orders of magnitude. Aggregate rows such as World are excluded; European Union is treated as a single reporting economy within Europe in the WTO dataset.",
      width = 115
    )
  ) +
  bubble_theme

gg_bubble_top3_rest
```

## Inside Each Region (Small Multiples)

```{r}
#| label: bubble-chart-faceted-region
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

bubble_faceted <- bubble_country_data |>
  group_by(geo_region) |>
  mutate(
    region_value_rank = row_number(desc(export_value)),
    facet_label = if_else(region_value_rank <= 4, country, NA_character_)
  ) |>
  ungroup()

gg_bubble_faceted <- ggplot(
  bubble_faceted,
  aes(
    x = export_quantity_million_tonnes,
    y = export_value_usd_billions,
    color = geo_region
  )
) +
  # Fixed point size: this view is about where economies sit within their
  # region, not unit price, so we avoid the extreme bubble that a few
  # micro-exporters (tiny tonnage, non-trivial value) would otherwise create.
  geom_point(size = 2.3, alpha = 0.7) +
  ggrepel::geom_text_repel(
    aes(label = facet_label),
    size = 2.9,
    color = "grey20",
    show.legend = FALSE,
    max.overlaps = Inf,
    box.padding = 0.5,
    min.segment.length = 0,
    segment.color = "grey75",
    na.rm = TRUE
  ) +
  facet_wrap(vars(geo_region), scales = "free", ncol = 3) +
  scale_color_manual(values = bubble_region_palette, guide = "none") +
  # Power-of-ten ticks keep the labels short so they do not overlap in the
  # narrow panels; the units are stated in the axis titles.
  scale_x_log10(
    breaks = scales::breaks_log(n = 4),
    labels = scales::label_log()
  ) +
  scale_y_log10(
    breaks = scales::breaks_log(n = 4),
    labels = scales::label_log()
  ) +
  labs(
    x = "Agricultural Export Volume (million tonnes, log scale)",
    y = "Agricultural Export Value (US$ billions, log scale)",
    title = "Inside Each Region: Value vs Volume of Every Reporting Economy in 2024",
    subtitle = str_wrap(
      "Each panel has its own axis range so within-region structure is visible; the top four economies per region are labelled. Axis ticks are powers of ten.",
      width = 100
    ),
    caption = str_wrap(
      "Because each panel is scaled independently, bubble positions should be compared within a region, not across panels. European Union is treated as a single reporting economy within Europe.",
      width = 115
    )
  ) +
  bubble_theme +
  theme(
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1.2, "lines"),
    axis.text = element_text(size = 9)
  )

gg_bubble_faceted
```

## Story 4: Top 15 by Value vs Top 15 by Volume

```{r}
#| label: prepare-bubble-value-volume-facets
#| message: false

# Take the top 15 economies by a chosen metric as individual bubbles, and
# collapse every remaining economy into one "Rest of World" bubble.
bubble_top15_with_rest <- function(data, rank_metric, facet_label) {
  ranked <- data |>
    mutate(metric_rank = row_number(desc({{ rank_metric }})))

  top15 <- ranked |>
    filter(metric_rank <= 15) |>
    transmute(
      facet = facet_label,
      entity = country,
      entity_group = geo_region,
      entity_type = "Individual economy",
      export_value,
      export_quantity
    )

  rest <- ranked |>
    filter(metric_rank > 15) |>
    summarise(
      rest_countries = n(),
      export_value = sum(export_value, na.rm = TRUE),
      export_quantity = sum(export_quantity, na.rm = TRUE),
      .groups = "drop"
    ) |>
    transmute(
      facet = facet_label,
      entity = str_c("Rest of World (", rest_countries, ")"),
      entity_group = "Rest of World",
      entity_type = "Rest of World",
      export_value,
      export_quantity
    )

  bind_rows(top15, rest)
}

bubble_value_volume_facets <- bind_rows(
  bubble_top15_with_rest(
    bubble_country_data, export_value, "Top 15 by export value"
  ),
  bubble_top15_with_rest(
    bubble_country_data, export_quantity, "Top 15 by export volume"
  )
) |>
  mutate(
    export_value_usd_billions = export_value / 1000,
    export_quantity_million_tonnes = export_quantity / 1000,
    usd_per_tonne = (export_value * 1000) / export_quantity,
    entity_type = factor(
      entity_type,
      levels = c("Individual economy", "Rest of World")
    ),
    facet = factor(
      facet,
      levels = c("Top 15 by export value", "Top 15 by export volume")
    )
  )
```

```{r}
#| label: bubble-chart-value-volume-facets
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 15

gg_bubble_value_volume <- ggplot(
  bubble_value_volume_facets,
  aes(
    x = export_quantity_million_tonnes,
    y = export_value_usd_billions,
    size = usd_per_tonne,
    color = entity_group,
    shape = entity_type
  )
) +
  geom_point(alpha = 0.85, stroke = 1) +
  ggrepel::geom_text_repel(
    aes(label = entity),
    size = 3,
    color = "grey20",
    show.legend = FALSE,
    max.overlaps = Inf,
    min.segment.length = 0,
    segment.color = "grey70"
  ) +
  facet_wrap(vars(facet)) +
  # Solid circles for individual economies; hollow circles for the pooled
  # Rest of World bubble in each panel.
  scale_shape_manual(
    values = c("Individual economy" = 16, "Rest of World" = 1),
    name = "Bubble type"
  ) +
  scale_size_area(
    max_size = 18,
    name = "Unit price (US$ / tonne)",
    labels = scales::label_dollar(prefix = "US$")
  ) +
  scale_color_manual(
    values = bubble_region_rest_palette,
    name = "Geographic Region",
    guide = guide_legend(override.aes = list(size = 5, shape = 16))
  ) +
  scale_x_log10(labels = scales::label_number(suffix = " Mt")) +
  scale_y_log10(labels = scales::label_dollar(prefix = "US$", suffix = "B")) +
  labs(
    x = "Agricultural Export Volume (million tonnes, log scale)",
    y = "Agricultural Export Value (US$ billions, log scale)",
    title = "Top 15 Agricultural Exporters by Value vs by Volume in 2024",
    subtitle = str_wrap(
      "The left panel ranks the 15 largest economies by export value; the right panel ranks by export volume. Each panel pools all remaining economies into one grey Rest of World bubble. Both panels share the same axes.",
      width = 110
    ),
    caption = str_wrap(
      "The two panels contain different sets of economies because value and volume rank countries differently. Bubble size shows average unit price. European Union is treated as a single reporting economy within Europe.",
      width = 115
    )
  ) +
  bubble_theme +
  theme(strip.text = element_text(face = "bold"))

gg_bubble_value_volume
```

This paired view shows how the list of top exporters changes depending on the
metric. A country can be in the top 15 by value but not by volume, which points
to high-value, lower-tonnage exports, or the other way around for bulk
commodities. The two Rest of World bubbles are worth comparing as well. Because
value and volume put different countries into the top 15, the pooled remainder is
not the same in the two panels, both in which countries it contains and in its
size.

## Story 5: Global Leaders and the Rest of the World

```{r}
#| label: bubble-chart-global-rest
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 13

bubble_global_ranked <- bubble_country_data |>
  mutate(global_rank = row_number(desc(export_value)))

bubble_global_top <- bubble_global_ranked |>
  filter(global_rank <= 15) |>
  transmute(
    entity = country,
    entity_group = geo_region,
    entity_type = "Individual economy",
    export_value,
    export_quantity
  )

bubble_global_rest <- bubble_global_ranked |>
  filter(global_rank > 15) |>
  summarise(
    rest_countries = n(),
    export_value = sum(export_value, na.rm = TRUE),
    export_quantity = sum(export_quantity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  transmute(
    entity = str_c("Rest of World (", rest_countries, ")"),
    entity_group = "Rest of World",
    entity_type = "Rest of World",
    export_value,
    export_quantity
  )

bubble_global <- bind_rows(bubble_global_top, bubble_global_rest) |>
  mutate(
    export_value_usd_billions = export_value / 1000,
    export_quantity_million_tonnes = export_quantity / 1000,
    usd_per_tonne = (export_value * 1000) / export_quantity,
    entity_type = factor(
      entity_type,
      levels = c("Individual economy", "Rest of World")
    )
  )

gg_bubble_global <- ggplot(
  bubble_global,
  aes(
    x = export_quantity_million_tonnes,
    y = export_value_usd_billions,
    size = usd_per_tonne,
    color = entity_group,
    shape = entity_type
  )
) +
  geom_point(alpha = 0.85, stroke = 1) +
  ggrepel::geom_text_repel(
    aes(label = entity),
    size = 3.2,
    color = "grey20",
    show.legend = FALSE,
    max.overlaps = Inf,
    min.segment.length = 0,
    segment.color = "grey70"
  ) +
  scale_shape_manual(
    values = c("Individual economy" = 16, "Rest of World" = 1),
    name = "Bubble type"
  ) +
  scale_size_area(
    max_size = 20,
    name = "Unit price (US$ / tonne)",
    labels = scales::label_dollar(prefix = "US$")
  ) +
  scale_color_manual(
    values = bubble_region_rest_palette,
    name = "Geographic Region",
    guide = guide_legend(override.aes = list(size = 5, shape = 16))
  ) +
  scale_x_log10(labels = scales::label_number(suffix = " Mt")) +
  scale_y_log10(labels = scales::label_dollar(prefix = "US$", suffix = "B")) +
  labs(
    x = "Agricultural Export Volume (million tonnes, log scale)",
    y = "Agricultural Export Value (US$ billions, log scale)",
    title = "The 15 Largest Agricultural Exporters and the Rest of the World in 2024",
    subtitle = str_wrap(
      "The 15 highest-value economies are shown individually; every remaining economy is combined into one grey bubble. Bubble size shows average unit price.",
      width = 100
    ),
    caption = str_wrap(
      "The Rest of World bubble aggregates all economies outside the top 15 by value. European Union is treated as a single reporting economy within Europe in the WTO dataset.",
      width = 115
    )
  ) +
  bubble_theme

gg_bubble_global
```

This last view looks at how concentrated trade is, similar to the infographic
that the assignment is based on. A small group of economies takes up most of
global agricultural trade. At the same time, the single "Rest of World" bubble,
which pools well over a hundred economies, still holds a large share of both
value and volume. Putting that pooled bubble next to the named leaders makes it
easy to see how concentrated trade is and where that concentration stops.
