## Load Processed Data
```{r}
#| label: fetch-preprocessed-data
#| message: false
preprocessed_data <- read_csv(export_path)
```

Fetching the preprocessed data from the `data/processed` directory, which contains the combined export values and quantities for each reporting country.

## Prepare Labels, Rankings, and Formatters
```{r}
#| label: data-formats-functions

value_formatter <- label_number(
  scale = 1 / 1000,
  accuracy = 0.1,
)

quantity_formatter <- label_number(
  scale = 1 / 1000,
  accuracy = 0.1,
)

price_formatter <- label_dollar(
  accuracy = 1,
  suffix = "/t"
)
```

`value_formatter` and `quantity_formatter` are functions that format the export value and quantity for display in the lollipop chart. The export value is converted from **millions to billions**, while the export quantity is converted from **thousand tonnes to million tonnes**. 

`price_formatter` formats unit values as currency rounded to the nearest dollar with a **"/t"** suffix (e.g., **"$1,250/t"**), providing a clean display for the color bar legend.

```{r}
#| label: prepare-lollipop-data
lollipop_country_data <- preprocessed_data |>
  mutate(
    usd_per_tonne = export_value * 1000 / export_quantity,
    point_label = str_c(value_formatter(export_value), " B"),
    value_rank = min_rank(desc(export_value)),
    volume_rank = min_rank(desc(export_quantity))
  )

lollipop_top20 <- lollipop_country_data |>
  mutate(country = fct_reorder(country, export_value)) |>
  slice_max(
    order_by = export_value,
    n = 20,
    with_ties = FALSE
  )

price_range <- range(lollipop_top20$usd_per_tonne, na.rm = TRUE)

price_breaks <- scales::breaks_log(n = 6)(price_range)

lollipop_top20
```

The above code prepares the preprocessed dataset for subsequent visualization. It calculates the unit value (`usd_per_tonne`), constructs formatted display labels for export values (`point_label`), and calculates rank positions for both trade value and volume. Finally, it filters the top 20 reporting economies by value and establishes log-transformed scale breaks (`price_breaks`) for the color legend, ensuring readability and visual consistency in the chart.

## Calculate Rest of World Context
```{r}
#| label: rest-of-world-data

lollipop_rest_of_world <- lollipop_country_data |>
  filter(
    !country %in% lollipop_top20$country
  ) |>
  summarise(
    country = "Rest of the World",
    geo_region = "Rest of the World",
    export_value = sum(export_value),
    export_quantity = sum(export_quantity),
    .groups = "drop"
  ) |>
  mutate(
    annotation_label = str_c(
      "All other reporting economies\n",
      value_formatter(export_value), " B",
      " | ",
      quantity_formatter(export_quantity), " million tonnes"
    )
  )

lollipop_rest_of_world <- lollipop_rest_of_world |>
  mutate(country = fct_reorder(country, export_value))

lollipop_rest_of_world
```


## Export Value with Export Volume Annotation
```{r}
#| label: lollipop-chart-breaks
#| message: false
#| warning: false

x_max <- max(lollipop_top20$export_value, na.rm = TRUE)

break_step <- 20 * 1000
minor_break_step <- 10 * 1000

x_breaks <- seq(
  0,
  ceiling(x_max / break_step) * break_step,
  by = break_step
)

x_minor_breaks <- seq(
  0,
  ceiling(x_max / minor_break_step) * minor_break_step,
  by = minor_break_step
)

annotation_data <- lollipop_rest_of_world |>
  transmute(
    x = x_max * 0.80,
    label = annotation_label
  )
```

This section configures the dynamic scale positioning and annotation placement for the chart. It calculates appropriate major (`x_breaks`) and minor (`x_minor_breaks`) x-axis intervals based on the maximum export value, ensuring consistent axis spacing across updates. Additionally, it constructs `annotation_data` to position the aggregate rest-of-world metrics dynamically near the right side of the visualization.

## Build the Lollipop Chart
The final visualization keeps **export value** as the primary ranking measure and uses **export quantity** as supporting context. This avoids comparing USD and tonnes on the same axis while still revealing countries whose economic export value differs from their physical export volume.

```{r}
#| label: lollipop-chart-value-with-volume-annotation
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 15

gg_lollipop_value_quantity <- ggplot(
  lollipop_top20,
  aes(
    x = export_value,
    y = country
  )
) +
  geom_segment(
    aes(
      x = 0,
      xend = export_value,
      yend = country
    ),
    linewidth = 1,
    alpha = 0.7
  ) +
  geom_point(
    aes(fill = usd_per_tonne),
    shape = 21,
    size = 4,
    stroke = 0.5,
    color = "grey40"
  ) +
  geom_text(
    aes(label = point_label),
    hjust = -0.35,
    size = 5.5,
    color = "grey25",
    show.legend = FALSE
  ) +
  geom_label(
    data = annotation_data,
    aes(
      x = x,
      y = lollipop_top20$country[nrow(lollipop_top20) - 2],
      label = label
    ),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 0.5,
    size = 5,
    lineheight = 1,
    colour = "grey20",
    fill = "white",
    linewidth = 0.35,
    label.padding = grid::unit(0.4, "lines")
  ) +
  scale_x_continuous(
    limits = c(0, x_max * 1.12),
    labels = value_formatter,
    breaks = x_breaks,
    minor_breaks = x_minor_breaks,
    expand = expansion(mult = c(0, 0.04))
  ) +
  scale_fill_fermenter(
    palette = "Blues",
    direction = 1,
    trans = "log10",
    breaks = price_breaks,
    labels = price_formatter,
    limits = range(price_breaks),
    oob = scales::squish,
    name = "Export value per tonne\n[USD/tonne, log scale]",
    guide = guide_coloursteps(
      title.position = "top",
      ticks.colour = "grey30",
      frame.colour = "grey65",
      theme = theme(
        legend.key.height = grid::unit(160, "pt"),
        legend.key.width = grid::unit(35, "pt"),
        legend.title = element_text(size = 16, face="bold"),
        legend.text = element_text(size = 14)
      )
    )
  ) +
  labs(
    x = "Agricultural Export Value (USD Billions)",
    y = "Reporting Economy",
    title = "Top 20 Agricultural Exporters including Rest of World by Export Value in 2024",
    subtitle = str_wrap(
    "Point color reflects export value per tonne, highlighting economies whose trade is concentrated in higher or lower-value goods.",
    width = 150
  ),
  caption = str_wrap(
      "Source: WTO dataset. The European Union is treated as a reporting economy.",
      width = 125
    ) 
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.y = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.title.x = element_text(size = 18, margin = margin(t = 10)),
    axis.title.y = element_text(size = 18, margin = margin(r = 10)),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey86", linewidth = 0.5),
    panel.grid.minor.x = element_line(color = "grey93", linewidth = 0.35),
    panel.grid.minor.y = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(size = 20, hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0),
    plot.margin = margin(10, 0, 16, 10),   # reduced right margin, was 24
    legend.box.spacing = grid::unit(5, "pt"),  # pulls legend closer to plot
    legend.margin = margin(0, 0, 0, 0)          # trims internal legend padding
  )

gg_lollipop_value_quantity
```

## Interpretation of Final Visualization

The redesigned chart demonstrates that agricultural export leadership is driven by both overall volume and product unit value. The log-scale color gradient clearly distinguishes high-unit-value exporters (darker points, such as Mexico, New Zealand, and Chile) from bulk commodity exporters (lighter points, such as Brazil, Ukraine, and the Russian Federation). By pairing explicit dollar labels at each point with unit-value color encoding, the visualization reveals structural differences in trade profiles without cluttering the primary ranking.