# Lollipop Visualisation

## Preprocessed Data
```{r}
#| label: fetch-preprocessed-data
#| message: false
preprocessed_data <- read_csv(export_path)
```

Fetching the preprocessed data from the `data/processed` directory, which contains the combined export values and quantities for each reporting country.

## Data Preparation for Lollipop Chart
```{r}
#| label: data-formats-functions

value_formatter <- label_number(
  scale = 1 / 1000,
  accuracy = 0.1,
  prefix = "US$",
  suffix = "B"
)

quantity_formatter <- label_number(
  scale = 1 / 1000,
  accuracy = 0.1,
  suffix = " Mt"
)
```

`value_formatter` and `quantity_formatter` are functions that format the export value and quantity for display in the lollipop chart. The export value is converted from **millions to billions**, while the export quantity is converted from **thousand tonnes to million tonnes**.

```{r}
#| label: prepare-lollipop-data
lollipop_country_data <- preprocessed_data |>
  mutate(
    point_label = str_c(
      value_formatter(export_value),
      " | ",
      quantity_formatter(export_quantity)
    ),
    value_rank = min_rank(desc(export_value)),
    volume_rank = min_rank(desc(export_quantity))
  )

lollipop_top30 <- lollipop_country_data |>
  mutate(country = fct_reorder(country, export_value)) |>
  slice_max(
    order_by = export_value,
    n = 30,
    with_ties = FALSE
  )

lollipop_top30
```

The above codes prepare the preprocessed dataset for subsequent visualisation. It creates labels for both export values and export quantity, and ranks the countries based on export value. This transformation are made such that the data and its labels are ready for subsquent visualisation ensuring readability and consistency.

```{r}
#| label: rest-of-world-data

lollipop_rest_of_world <- lollipop_country_data |>
  filter(
    !country %in% lollipop_top30$country
  ) |>
  summarise(
    country = "Rest of World",
    geo_region = "Rest of World",
    export_value = sum(export_value),
    export_quantity = sum(export_quantity),
    .groups = "drop"
  ) |>
  mutate(
    point_label = str_c(
      value_formatter(export_value),
      " | ",
      quantity_formatter(export_quantity)
    ),
    value_rank = min_rank(desc(export_value)),
    volume_rank = min_rank(desc(export_quantity))
  )

lollipop_rest_of_world <- lollipop_rest_of_world |>
  mutate(country = fct_reorder(country, export_value))

lollipop_rest_of_world

lollipop_top30_with_rest_of_world <- bind_rows(
  lollipop_top30,
  lollipop_rest_of_world
) |>
  mutate(country = fct_reorder(country, export_value))

head(lollipop_top30_with_rest_of_world)
tail(lollipop_top30_with_rest_of_world)
```

## Lollipop Chart Standardise Values
```{r}
lollipop_okabe_ito_palette <- c(
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#CC79A7", # reddish purple
  "#999999"  # neutral grey, used instead of black for softer contrast
)
lollipop_region_levels <- sort(unique(lollipop_country_data$geo_region))

if (length(lollipop_region_levels) > length(lollipop_okabe_ito_palette)) {
  stop("There are more geographic regions than available Okabe-Ito colours.")
}

lollipop_region_palette <- set_names(
  lollipop_okabe_ito_palette[seq_along(lollipop_region_levels)],
  lollipop_region_levels
)

lollipop_region_palette["Rest of World"] <- "#999999"
```

To ensure that the visualisation is color-blind friendly, the team decided to use [Okabe-Ito colour-blind-safe palette]( https://jfly.uni-koeln.de/color/#pallet) as our color palette. 

## Export Value with Export Volume Annotation

```{r}
#| label: lollipop-chart-value-with-volume-annotation
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 15

x_max <- max(lollipop_top30_with_rest_of_world$export_value, na.rm = TRUE)

gg_lollipop_value_quantity <- ggplot(
  lollipop_top30_with_rest_of_world,
  aes(
    x = export_value,
    y = country,
    color = geo_region
  )
) +
  geom_segment(
    aes(
      x = 0,
      xend = export_value,
      yend = country
    ),
    linewidth = 1,
    alpha = 0.7
  ) +
  geom_point(
    size = 3.2
  ) +
  geom_text(
    aes(label = point_label),
    hjust = -0.15,
    size = 3.5,
    color = "grey25",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = lollipop_region_palette,
    name = "Geographic Region"
  ) +
  scale_x_continuous(
    limits = c(0, x_max * 1.22),
    labels = value_formatter,
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Agricultural Export Value (USD Billions)",
    y = "Country / Reporting Economy",
    title = "Top 30 Agricultural Exporters including Rest of World by Export Value in 2024",
    caption = str_wrap(
      "Export volume is shown in million tonnes (Mt) for product groups measured by physical quantity. European Union is treated as a reporting economy in the WTO dataset.",
      width = 125
    )
  ) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    plot.margin = margin(10, 24, 16, 10)
  )

gg_lollipop_value_quantity
```

This is the main redesign view. It keeps the primary comparison on one visual scale: export value. Export volume is included as a direct label so readers can distinguish high-value exporters from high-volume exporters without placing USD and tonnes on the same axis.