# Stacked Bar: Regional Composition (Value and Quantity, Faceted)

```{r}
#| label: stacked-bar-region-composition-prep

# Named economies per region. Everything below the cut collapses into one
# "Rest of <region>" segment so each bar still represents 100% of that region.
n_top <- 3

# Rank WITHIN each region by export value so the SAME three economies stay named in both
# the value and the quantity panel. The guard drops any pre-aggregated
# "Rest of X" rows so the residual is never double-counted.
ranked <- combined_country_data |>
  filter(!str_detect(country, "^Rest of")) |>
  group_by(geo_region) |>
  mutate(value_rank = row_number(desc(export_value))) |>
  ungroup() |>
  mutate(is_rest = value_rank > n_top)

# Top-3 economies kept individually.
named_segments <- ranked |>
  filter(!is_rest) |>
  transmute(geo_region, segment = country, is_rest,
            export_value, export_quantity)

# Everyone ranked 4th or lower per region, pooled into a counted residual.
rest_segments <- ranked |>
  filter(is_rest) |>
  group_by(geo_region) |>
  summarise(rest_countries = n(),
            export_value = sum(export_value, na.rm = TRUE),
            export_quantity = sum(export_quantity, na.rm = TRUE),
            .groups = "drop") |>
  mutate(segment = str_c("Rest of ", geo_region, " (", rest_countries, ")"),
         is_rest = TRUE) |>
  transmute(geo_region, segment, is_rest, export_value, export_quantity)

# One long table with a single `measure` column so a single facet can split it.
stacked_data <- bind_rows(named_segments, rest_segments) |>
  pivot_longer(
    cols = c(export_value, export_quantity),
    names_to = "measure",
    values_to = "amount"
  ) |>
  filter(!is.na(amount), amount > 0) |>
  mutate(measure = recode(measure,
    export_value    = "Export value (USD millions)",
    export_quantity = "Export quantity (tonnes)"))

# Stable region ordering (by total export value) shared across both panels.
region_order <- stacked_data |>
  filter(measure == "Export value (USD millions)") |>
  group_by(geo_region) |>
  summarise(total = sum(amount), .groups = "drop") |>
  arrange(total) |>
  pull(geo_region)

stacked_plot_df <- stacked_data |>
  mutate(geo_region = factor(geo_region, levels = region_order),
         stack_key  = if_else(is_rest, 0, amount)) |>
  arrange(measure, geo_region, desc(is_rest), desc(amount))
```

```{r}
#| label: stacked-bar-region-composition
#| fig-width: 12
#| fig-height: 6
#| fig-dpi: 300
#| warning: false

ggplot(
  stacked_plot_df,
  aes(x = amount, y = geo_region,
      fill = is_rest,
      group = interaction(is_rest, -stack_key))
) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  geom_text(
    aes(label = if_else(!is_rest, segment, "")),
    position = position_stack(vjust = 0.5),
    size = 2.3, color = "white"
  ) +
  facet_wrap(~ measure, scales = "free_x") +
  scale_x_continuous(labels = scales::label_comma(),
                     expand = expansion(mult = c(0, 0.04))) +
  scale_fill_manual(
    values = c(`FALSE` = "#2f7d6b", `TRUE` = "#c9ccd1"),
    labels = c(`FALSE` = "Top 3 by value (named)", `TRUE` = "Rest of region"),
    name = NULL
  ) +
  labs(
    title = "Agricultural Exports by Region: Value vs. Quantity, 2024",
    subtitle = "Small multiples of one stacked bar; the same three value-leaders are named in both panels, and each bar covers 100% of its region",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "top",
        panel.grid.major.y = element_blank())
```

This chart is presented as **small multiples of a single stacked bar** rather
than two separate charts: one `ggplot()` faceted by `measure` produces a value
panel and a quantity panel from the same table.


Because the top three are fixed by value, a value-leader with modest tonnage
appears as a large segment in the value panel but a small one in the quantity
panel, with the difference absorbed into that region's residual — making the
value-versus-volume contrast directly readable. The chart retains the inherent
limitation of stacked bars: only the segment anchored at the zero baseline can be
compared by position, while the rest must be judged by length from a shifting
baseline, which is less precise than position on a common scale
(Cleveland & McGill, 1985). Value covers all 32 W32 product groups whereas
quantity covers only the tonne-measured groups, so the panels are read as a
structural comparison rather than a unit-for-unit conversion, and the chart shows
a single year (2024) without temporal comparison.
```
