# Stacked Bar Chart Visualisation

# Fetch the Preprocessed data
```{r}
#| label: fetch-preprocessed-data-stacked
#| message: false

library(tidyverse)
library(readr)
library(forcats)
library(scales)
library(stringr)
library(tibble)

preprocessed_data <- read_csv(export_path)
```

# Filter Rows
```{r}
#| label: filter-stacked-country-data
#| message: false

stacked_country_data <- preprocessed_data |>
  filter(
    str_squish(country) != "World",
    !is.na(export_value),
    !is.na(export_quantity),
    !is.na(geo_region)
  )
```

# Select top exporters
```{r}
#| label: prepare-top15-stacked-data
#| message: false

# n = 15 matches `bubble_top15_with_rest()` in the bubble visualisation. The bar
# chart and scatter use n = 30; at 30, Asia contributes 11 named economies to a
# single bar and South & Central America contributes 7 (Brazil at 221.5 Mt beside
# Peru at 0.7 Mt), which cannot be direct-labelled without overlap. n = 15 keeps
# every segment individually labelled.
n_global <- 15

stacked_top15 <- stacked_country_data |>
  slice_max(
    order_by = export_value,
    n = n_global,
    with_ties = FALSE
  )
```

# ISO3 Codes
```{r}
#| label: define-iso3-codes
#| message: false

# Named economies are labelled with ISO3 codes so a label fits above its own
# segment. The pooled remainder keeps a written label, since an abbreviation
# there would not be interpretable.
iso3_codes <- c(
  "European Union" = "EU", "Brazil" = "BRA",
  "United States" = "USA", "Canada" = "CAN",
  "Mexico" = "MEX", "China" = "CHN",
  "Australia" = "AUS", "Argentina" = "ARG",
  "India" = "IND", "Indonesia" = "IDN",
  "Thailand" = "THA", "Ukraine" = "UKR",
  "Turkey" = "TUR", "Russian Federation" = "RUS",
  "Malaysia" = "MYS", "New Zealand" = "NZL",
  "United Kingdom" = "GBR", "Vietnam" = "VNM",
  "Chile" = "CHL", "Peru" = "PER",
  "South Africa" = "ZAF", "Colombia" = "COL",
  "Cote d'Ivoire" = "CIV", "Ecuador" = "ECU",
  "Egypt, Arab Rep." = "EGY", "Paraguay" = "PRY",
  "Pakistan" = "PAK", "Hong Kong, China" = "HKG",
  "Philippines" = "PHL", "Switzerland" = "CHE"
)

to_iso3 <- function(country) {
  code <- unname(iso3_codes[country])
  if (any(is.na(code))) {
    stop(
      "Missing ISO3 code for: ",
      str_c(country[is.na(code)], collapse = ", ")
    )
  }
  code
}
```

# Okabe Ito Palette
```{r}
#| label: define-stacked-region-palette
#| message: false

okabe_ito_palette <- c(
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#CC79A7", # reddish purple
  "#999999" # grey
)

region_levels <- sort(unique(stacked_country_data$geo_region))

if (length(region_levels) > length(okabe_ito_palette)) {
  stop("There are more geographic regions than available Okabe-Ito colours.")
}

region_palette <- set_names(
  okabe_ito_palette[seq_along(region_levels)],
  region_levels
)

# Hue encodes region (qualitative). Lightness encodes rank within the panel
# (sequential): the largest economy takes the region's full-strength colour and
# smaller ones are progressively lightened towards white.
lighten_hex <- function(hex, factor) {
  channels <- col2rgb(hex) / 255
  channels <- channels + (1 - channels) * factor
  rgb(channels[1], channels[2], channels[3])
}
```

# Set Styling Constants
```{r}
#| label: set-stacked-chart-styling
#| message: false

stacked_base_text_size <- 12
stacked_strip_text_size <- 10
stacked_axis_text_size <- 8
stacked_label_size <- 2.6
stacked_rest_grey <- "#D9D9D9"
stacked_max_lighten <- 0.62
stacked_bar_width <- 0.30
stacked_tier_gap <- 0.30
stacked_fig_width <- 16
stacked_fig_height <- 13
stacked_n_cols <- 2
```

# Build the segments
```{r}
#| label: prepare-stacked-segments
#| message: false

# Each region's members of the global top 15, named individually.
named_segments <- stacked_top15 |>
  transmute(
    geo_region,
    segment = to_iso3(country),
    is_rest = FALSE,
    rest_n = NA_integer_,
    export_value,
    export_quantity
  )

# Every remaining economy in that region, pooled into one counted residual, so
# each bar still accounts for 100% of its region's recorded trade.
rest_segments <- stacked_country_data |>
  anti_join(stacked_top15 |> select(country), by = "country") |>
  group_by(geo_region) |>
  summarise(
    rest_n = n(),
    export_value = sum(export_value, na.rm = TRUE),
    export_quantity = sum(export_quantity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  transmute(
    geo_region,
    segment = "Rest",
    is_rest = TRUE,
    rest_n,
    export_value,
    export_quantity
  )

stacked_segments <- bind_rows(named_segments, rest_segments) |>
  pivot_longer(
    cols = c(export_value, export_quantity),
    names_to = "measure",
    values_to = "raw_amount"
  ) |>
  mutate(
    measure = recode(
      measure,
      export_value = "Export value (US$ billions)",
      export_quantity = "Export quantity (million tonnes)"
    ),
    # Source units: value in USD millions, quantity in thousand tonnes.
    amount = raw_amount / 1000
  ) |>
  filter(amount > 0)
```

```{r}
#| label: prepare-stacked-geometry
#| message: false

stacked_plot_df <- stacked_segments |>
  group_by(measure, geo_region) |>
  # Draw order: the pooled remainder sits at the ORIGIN, then the named
  # economies ascend, so the largest lands at the far right of each bar.
  arrange(desc(is_rest), amount, .by_group = TRUE) |>
  mutate(
    # Shading rank is computed independently of draw order: 1 = largest in this
    # panel = darkest. The residual must not consume a rank slot, or a country's
    # shade would fall outside the intended ramp.
    panel_rank = if_else(
      is_rest,
      NA_integer_,
      as.integer(rank(-if_else(is_rest, -Inf, amount), ties.method = "first"))
    ),
    n_named = sum(!is_rest),
    # Rectangles are drawn from these coordinates directly, so bar geometry and
    # label position cannot disagree (no reliance on position_stack()).
    x_max = cumsum(amount),
    x_min = x_max - amount,
    x_mid = x_max - amount / 2,
    panel_max = max(x_max) * 1.03
  ) |>
  ungroup() |>
  mutate(
    fill_hex = pmap_chr(
      list(geo_region, panel_rank, n_named, is_rest),
      \(region, rank, n, rest) {
        if (rest || is.na(rank)) {
          return(stacked_rest_grey)
        }
        lighten_hex(
          region_palette[[region]],
          if (n <= 1) 0 else (rank - 1) / (n - 1) * stacked_max_lighten
        )
      }
    ),
    # Two-line label for named economies, three-line for the residual. Stacking
    # the value UNDER the name rather than beside it roughly halves the label's
    # WIDTH, and width is the binding constraint on placement.
    label = if_else(
      is_rest,
      str_c("Rest\n(", rest_n, ")\n", number(amount, accuracy = 0.1)),
      str_c(segment, "\n", number(amount, accuracy = 0.1))
    ),
    panel = str_c(geo_region, " \u2014 ", measure)
  )

# Panels: regions ordered by total export value, each followed by its own
# quantity panel.
region_order <- stacked_plot_df |>
  filter(measure == "Export value (US$ billions)") |>
  group_by(geo_region) |>
  summarise(total = sum(amount), .groups = "drop") |>
  arrange(desc(total)) |>
  pull(geo_region)

panel_levels <- as.vector(t(outer(
  region_order,
  c("Export value (US$ billions)", "Export quantity (million tonnes)"),
  \(region, measure) str_c(region, " \u2014 ", measure)
)))

stacked_plot_df <- stacked_plot_df |>
  mutate(panel = factor(panel, levels = panel_levels))
```

# Place the direct labels
```{r}
#| label: place-stacked-labels
#| message: false

# Measure each label in data units. For a multi-line label the binding width is
# the WIDEST of its lines.
panel_inches <- (stacked_fig_width / stacked_n_cols) - 1.15

grDevices::pdf(NULL)
label_inches <- vapply(
  stacked_plot_df$label,
  \(text) {
    lines <- strsplit(text, "\n", fixed = TRUE)[[1]]
    max(vapply(
      lines,
      \(line) graphics::strwidth(line, units = "inches", cex = 0.80, font = 2),
      numeric(1)
    ))
  },
  numeric(1)
)
grDevices::dev.off()

stacked_plot_df <- stacked_plot_df |>
  mutate(
    label_width = (label_inches / panel_inches) * panel_max,
    # Clamp: a label centred on a segment at the panel edge must not be clipped.
    x_label = pmin(
      pmax(x_mid, label_width / 2),
      panel_max - label_width / 2
    )
  )

# Place every label on a single shared line where possible. A label may SLIDE
# horizontally within its own segment before it is bumped to a second tier:
# sliding keeps it over the segment it names, whereas a tier bump lifts it away
# from the bar.
place_labels <- function(panel_df) {
  panel_df <- panel_df |> arrange(x_label)
  occupied <- list()
  tier_of <- integer(nrow(panel_df))
  x_of <- numeric(nrow(panel_df))

  is_clear <- function(taken, lo, hi, pad) {
    is.null(taken) || all(lo >= taken$right + pad | hi <= taken$left - pad)
  }

  for (i in seq_len(nrow(panel_df))) {
    pad <- 0.004 * panel_df$panel_max[i]
    half <- panel_df$label_width[i] / 2
    lower <- half
    upper <- panel_df$panel_max[i] - half

    seg_lo <- panel_df$x_min[i] + half
    seg_hi <- panel_df$x_max[i] - half
    candidates <- panel_df$x_label[i]
    if (seg_hi > seg_lo) {
      steps <- seq(0, 1, by = 0.1)
      candidates <- unique(c(
        panel_df$x_label[i],
        pmin(pmax(panel_df$x_mid[i] + steps * (seg_hi - panel_df$x_mid[i]),
          seg_lo
        ), seg_hi),
        pmin(pmax(panel_df$x_mid[i] - steps * (panel_df$x_mid[i] - seg_lo),
          seg_lo
        ), seg_hi)
      ))
    }
    candidates <- pmin(pmax(candidates, lower), upper)
    candidates <- candidates[order(abs(candidates - panel_df$x_mid[i]))]

    placed <- FALSE
    tier <- 1L
    while (!placed) {
      taken <- occupied[[as.character(tier)]]
      for (candidate in candidates) {
        if (is_clear(taken, candidate - half, candidate + half, pad)) {
          occupied[[as.character(tier)]] <- bind_rows(
            taken,
            tibble(left = candidate - half, right = candidate + half)
          )
          tier_of[i] <- tier
          x_of[i] <- candidate
          placed <- TRUE
          break
        }
      }
      if (!placed) tier <- tier + 1L
    }
  }

  panel_df$tier <- tier_of
  panel_df$x_label <- x_of
  panel_df
}

stacked_plot_df <- stacked_plot_df |>
  group_split(panel) |>
  map_dfr(place_labels)

stacked_max_tier <- max(stacked_plot_df$tier)

stacked_plot_df <- stacked_plot_df |>
  mutate(
    y_min = -stacked_bar_width / 2,
    y_max = stacked_bar_width / 2,
    y_label = stacked_bar_width / 2 + 0.10 + (tier - 1) * stacked_tier_gap
  )
```

# Build the stacked bar chart
```{r}
#| label: build-stacked-bar-chart
#| message: false
#| warning: false
#| fig-height: 13
#| fig-width: 16

gg_stacked_region <- ggplot(stacked_plot_df) +
  geom_rect(
    aes(
      xmin = x_min,
      xmax = x_max,
      ymin = y_min,
      ymax = y_max,
      fill = fill_hex
    ),
    colour = "white",
    linewidth = 0.3
  ) +
  scale_fill_identity() +
  geom_text(
    aes(
      x = x_label,
      y = y_label,
      label = label,
      colour = if_else(is_rest, "rest", "named")
    ),
    size = stacked_label_size,
    fontface = "bold",
    lineheight = 0.92,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = c(named = "grey15", rest = "grey40")) +
  facet_wrap(
    vars(panel),
    ncol = stacked_n_cols,
    scales = "free_x"
  ) +
  # Each panel carries its own LINEAR x-range, so tick density adapts to the
  # region while segment length stays proportional to value.
  scale_x_continuous(
    breaks = scales::pretty_breaks(8),
    minor_breaks = scales::pretty_breaks(16),
    labels = scales::label_number(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  scale_y_continuous(
    limits = c(
      -stacked_bar_width / 2 - 0.06,
      stacked_bar_width / 2 + 0.14 + stacked_max_tier * stacked_tier_gap
    ),
    breaks = NULL,
    expand = expansion(0)
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "The World's 15 Largest Agricultural Exporters, Grouped by Region in 2024",
    subtitle = str_wrap(
      "Each bar is one region's full recorded trade: its members of the global top 15 are named individually by ISO3 code, and every remaining economy is pooled into a single grey segment. Segments ascend from the pooled remainder to the region's largest exporter, and the darkest shade marks the largest economy in that panel.",
      width = 150
    ),
    caption = str_wrap(
      "Each panel has its own x-range, so segments are compared within a region and not across panels. Africa and the Middle East contain no top-15 exporter, so their bars are entirely the pooled remainder. Export value is shown in US$ billions and export volume in million tonnes (Mt) for product groups measured by physical quantity. Aggregate rows such as World are excluded. European Union is treated as a reporting economy in the WTO dataset.",
      width = 165
    )
  ) +
  theme_minimal(base_size = stacked_base_text_size) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey85", linewidth = 0.3),
    panel.grid.minor.x = element_line(colour = "grey92", linewidth = 0.22),
    strip.text = element_text(face = "bold", size = stacked_strip_text_size),
    axis.text.x = element_text(size = stacked_axis_text_size),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
    panel.spacing.y = unit(0.5, "lines"),
    plot.margin = margin(10, 30, 16, 10)
  )

gg_stacked_region
```

Each regional bar accounts for 100% of that region's recorded trade, so the seven
bars together cover the whole of world agricultural trade. Hue encodes region
using the Okabe-Ito colour-blind-safe palette and lightness encodes rank within
the panel, which avoids a fifteen-entry country legend while keeping every
segment identifiable from its direct label.

Faceting each region onto its own x-range is what makes the direct labels
possible. On a shared axis, Brazil's 221.5 Mt sets the scale and the smaller
economies collapse into slivers too narrow to hold text. The cost is that
magnitude can no longer be compared across panels, so the panels are read within
a region rather than against one another.

The chart retains the inherent limitation of stacked bars: only the segment
anchored at the zero baseline can be compared by position, while the others must
be judged by length from a shifting baseline, which is less precise than position
along a common scale (Cleveland & McGill, 1985). It also shows a single year
(2024) and does not support temporal comparison.
