---
title: "2024-Focused Agriculture Export EDA"
author: "Team Orchid"
date: ""
format:
  html:
    toc: true
    number-sections: true
---

# Purpose

This document refocuses the exploratory data analysis on **2024** only and prepares cleaned datasets for the team's redesigned static infographic. The source workbook contains annual columns from 2014 to 2024, so the preprocessing workflow first converts the export value and export quantity sheets into tidy long format and then filters both datasets to the selected year.

Keeping the long-format transformation before filtering makes the workflow reproducible: if the team needs to compare another year later, only `analysis_year` needs to change. The final export step writes cleaned 2024 CSV files that can be submitted with the assignment and reused in the infographic design workflow.

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
```

```{r}
#| label: analysis-settings
analysis_year <- 2024
source_file <- "data/export_of_agriculture_products_data.xlsx"
export_dir <- "data/processed"
```

# Data Source

```{r}
#| label: workbook-sheets
excel_sheets(source_file)
```

The workbook contains overview tables, product definitions, export quantity data, and export value data. This EDA uses both the `Export values` and `Exports qties` worksheets because the planned infographic redesign needs to communicate both monetary export value and reported export quantity.

The `Export values` worksheet includes monetary export values in USD millions and a `GeoRegion` field for regional comparisons. The `Exports qties` worksheet provides reported export quantities, but it does not contain `GeoRegion`, so region-level analysis comes from the value dataset.

# Export Value Preprocessing

## Import Raw Value Data

```{r}
#| label: value-import
value_raw <- read_excel(
  source_file,
  sheet = "Export values",
  range = "A4:N4839"
)

glimpse(value_raw)
```

The first three rows of the worksheet contain workbook-level title and unit notes, so the import starts at row 4 where the data table begins. The selected range keeps `ProductDescription`, `ReporterName`, `GeoRegion`, and annual export value columns from 2014 to 2024.

## Clean and Reshape Value Data

```{r}
#| label: value-clean-long
value_clean <- value_raw |>
  rename(
    product_description = ProductDescription,
    reporter_name = ReporterName,
    geo_region = GeoRegion
  ) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "export_value_usd_millions"
  ) |>
  mutate(
    year = as.integer(year),
    product_description = str_squish(product_description),
    reporter_name = str_squish(reporter_name),
    geo_region = str_squish(geo_region),
    export_value_usd_millions = as.numeric(export_value_usd_millions)
  )

head(value_clean)
```

The value data preprocessing performs five steps:

1. Standardises column names into snake case.
2. Converts the dataset from wide format to long format, with one row per product, reporter, region, and year.
3. Converts `year` into an integer so it can be filtered and sorted correctly.
4. Converts export values into numeric form.
5. Removes extra whitespace from product, reporter, and region labels using `str_squish()`.

## Filter to 2024

```{r}
#| label: value-filter-2024
value_2024 <- value_clean |>
  filter(year == analysis_year)

head(value_2024)
```

All subsequent value-based EDA uses `value_2024`. This keeps the analysis focused on the latest available year in the workbook instead of mixing 2024 patterns with earlier years.

## Value Data Validation

```{r}
#| label: value-validation
value_validation <- tibble(
  check = c(
    "Rows in raw value data",
    "Columns in raw value data",
    "Rows after long transformation",
    "Available years",
    "Rows retained for 2024",
    "Missing 2024 export values",
    "Negative 2024 export values",
    "Duplicate 2024 product-reporter rows"
  ),
  result = c(
    nrow(value_raw),
    ncol(value_raw),
    nrow(value_clean),
    paste(range(value_clean$year), collapse = "-"),
    nrow(value_2024),
    sum(is.na(value_2024$export_value_usd_millions)),
    sum(value_2024$export_value_usd_millions < 0, na.rm = TRUE),
    value_2024 |>
      count(product_description, reporter_name) |>
      filter(n > 1) |>
      nrow()
  )
)

value_validation
```

The validation checks confirm whether the 2024 filter retained the expected one-year slice and whether any values require special handling before visualisation. Missing values would indicate incomplete source coverage, negative values require interpretation, and duplicate product-reporter rows would affect aggregation.

## Negative Value Check

```{r}
#| label: value-negative-2024
negative_values_2024 <- value_2024 |>
  filter(export_value_usd_millions < 0) |>
  arrange(export_value_usd_millions)

negative_values_2024
```

Negative values are not automatically removed. In this workbook, negative export values can appear in derived or residual product categories, especially where a category is calculated by subtraction. They are retained unless the team chooses to exclude residual categories for a specific visualisation.

## Construct 2024 Analysis Dataset

```{r}
#| label: value-analysis-dataset
aggregate_products <- c(
  "All W32 groups",
  "AoA HS proxy",
  "Other AoA"
)

value_analysis_2024 <- value_2024 |>
  filter(
    reporter_name != "World",
    !product_description %in% aggregate_products
  )

glimpse(value_analysis_2024)
```

Aggregate reporter and product rows are excluded from the general 2024 analysis dataset to avoid double-counting totals that are already represented by individual reporters and product groups. The aggregate rows remain available in `value_2024` for checks that specifically need official totals.

# 2024 Export Value EDA

## Missing and Zero Values

```{r}
#| label: value-zero-summary-2024
value_zero_summary_2024 <- value_analysis_2024 |>
  summarise(
    observations = n(),
    missing_values = sum(is.na(export_value_usd_millions)),
    zero_values = sum(export_value_usd_millions == 0, na.rm = TRUE),
    zero_percent = zero_values / observations * 100
  )

value_zero_summary_2024
```

Zero values are retained because they likely represent no recorded export value for a product-reporter combination in 2024. They are different from missing values, which would indicate absent or incomplete data.

## 2024 Summary Statistics

```{r}
#| label: value-summary-2024
value_summary_2024 <- value_analysis_2024 |>
  summarise(
    total_export_value = sum(export_value_usd_millions, na.rm = TRUE),
    mean = mean(export_value_usd_millions, na.rm = TRUE),
    median = median(export_value_usd_millions, na.rm = TRUE),
    q1 = quantile(export_value_usd_millions, 0.25, na.rm = TRUE),
    q3 = quantile(export_value_usd_millions, 0.75, na.rm = TRUE),
    max = max(export_value_usd_millions, na.rm = TRUE),
    sd = sd(export_value_usd_millions, na.rm = TRUE)
  )

value_summary_2024
```

The mean, median, quartiles, maximum, and standard deviation describe the spread of 2024 export values. Because export data is usually highly skewed, the median and quartiles are more reliable indicators of a typical product-reporter observation than the mean alone.

## 2024 Categorical Coverage

```{r}
#| label: value-categorical-coverage-2024
value_categorical_coverage_2024 <- value_analysis_2024 |>
  summarise(
    product_groups = n_distinct(product_description),
    reporters = n_distinct(reporter_name),
    regions = n_distinct(geo_region)
  )

value_categorical_coverage_2024
```

Coverage is checked after filtering to 2024 and removing aggregate rows. These counts define the valid comparison base for 2024 visualisations.

## Reporter Coverage by Region

```{r}
#| label: value-regional-coverage-2024
regional_coverage_2024 <- value_analysis_2024 |>
  distinct(reporter_name, geo_region) |>
  count(geo_region, name = "number_of_reporters") |>
  arrange(desc(number_of_reporters))

regional_coverage_2024
```

This table shows how many reporting economies are assigned to each geographical region in the 2024 analysis dataset. Regional export totals should be interpreted alongside this coverage because regions with more reporters naturally have more observations.

## Product Coverage Across Reporters

```{r}
#| label: value-product-coverage-2024
total_reporters_2024 <- n_distinct(value_analysis_2024$reporter_name)

product_coverage_2024 <- value_analysis_2024 |>
  distinct(product_description, reporter_name) |>
  count(product_description, name = "number_of_reporters") |>
  mutate(
    coverage_percent = round(
      number_of_reporters / total_reporters_2024 * 100,
      1
    )
  ) |>
  arrange(number_of_reporters, product_description)

product_coverage_2024
```

Product coverage is not assumed to be complete. If a product-reporter pair is absent from the 2024 data, it should not automatically be interpreted as zero exports unless the source documentation confirms that missing combinations represent zero trade.

## Top 2024 Export Regions

```{r}
#| label: value-top-regions-2024
top_regions_2024 <- value_analysis_2024 |>
  group_by(geo_region) |>
  summarise(
    total_export_value = sum(export_value_usd_millions, na.rm = TRUE),
    reporters = n_distinct(reporter_name),
    .groups = "drop"
  ) |>
  arrange(desc(total_export_value))

top_regions_2024
```

```{r}
#| label: value-top-regions-plot-2024
ggplot(
  top_regions_2024,
  aes(x = reorder(geo_region, total_export_value), y = total_export_value)
) +
  geom_col(fill = "#2f7d6b") +
  coord_flip() +
  labs(
    title = "2024 Export Value by Region",
    x = NULL,
    y = "Export value (USD millions)"
  ) +
  theme_minimal()
```

This regional view identifies where 2024 export value is concentrated geographically. The reporter count is included to help distinguish high regional totals from broad regional coverage.

## Top 2024 Export Reporters

```{r}
#| label: value-top-reporters-2024
top_reporters_2024 <- value_analysis_2024 |>
  group_by(reporter_name, geo_region) |>
  summarise(
    total_export_value = sum(export_value_usd_millions, na.rm = TRUE),
    product_groups = n_distinct(product_description),
    .groups = "drop"
  ) |>
  arrange(desc(total_export_value)) |>
  slice_head(n = 15)

top_reporters_2024
```

```{r}
#| label: value-top-reporters-plot-2024
ggplot(
  top_reporters_2024,
  aes(x = reorder(reporter_name, total_export_value), y = total_export_value)
) +
  geom_col(fill = "#6c5ce7") +
  coord_flip() +
  labs(
    title = "Top 15 Reporting Economies by 2024 Export Value",
    x = NULL,
    y = "Export value (USD millions)"
  ) +
  theme_minimal()
```

The top reporter table and chart identify the economies contributing the largest export values in 2024. Product group counts show whether each reporter's total comes from broad product coverage or from fewer reported categories.

## Top 2024 Product Groups

```{r}
#| label: value-top-products-2024
top_products_2024 <- value_analysis_2024 |>
  group_by(product_description) |>
  summarise(
    total_export_value = sum(export_value_usd_millions, na.rm = TRUE),
    reporters = n_distinct(reporter_name),
    .groups = "drop"
  ) |>
  arrange(desc(total_export_value)) |>
  slice_head(n = 15)

top_products_2024
```

```{r}
#| label: value-top-products-plot-2024
ggplot(
  top_products_2024,
  aes(
    x = reorder(product_description, total_export_value),
    y = total_export_value
  )
) +
  geom_col(fill = "#b85c38") +
  coord_flip() +
  labs(
    title = "Top 15 Product Groups by 2024 Export Value",
    x = NULL,
    y = "Export value (USD millions)"
  ) +
  theme_minimal()
```

This product-level summary identifies which product groups dominate 2024 export value. Reporter counts help flag product groups that are large because they are widely reported versus product groups dominated by a small number of major exporters.

## 2024 Export Value Distribution

```{r}
#| label: value-distribution-2024
positive_value_2024 <- value_analysis_2024 |>
  filter(export_value_usd_millions > 0)

ggplot(positive_value_2024, aes(x = export_value_usd_millions)) +
  geom_histogram(bins = 40, fill = "#4c78a8", color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of Positive 2024 Export Values",
    x = "Export value (USD millions, log scale)",
    y = "Number of observations"
  ) +
  theme_minimal()
```

Only positive values are used in the log-scale distribution because zero and negative values cannot be plotted on a logarithmic axis. The chart helps check whether the 2024 data is dominated by many small values and a small number of very large observations.

## 2024 Concentration Check

```{r}
#| label: value-concentration-2024
total_value_2024 <- sum(
  value_analysis_2024$export_value_usd_millions,
  na.rm = TRUE
)

reporter_concentration_2024 <- value_analysis_2024 |>
  group_by(reporter_name) |>
  summarise(
    total_export_value = sum(export_value_usd_millions, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_export_value)) |>
  mutate(
    rank = row_number(),
    cumulative_value = cumsum(total_export_value),
    cumulative_share_percent = cumulative_value / total_value_2024 * 100
  ) |>
  filter(rank <= 10)

reporter_concentration_2024
```

The concentration check shows how much of the 2024 export value is captured by the largest reporting economies. This is useful for deciding whether later visualisations should show all reporters, focus on the top reporters, or use log scales.

# 2024 Export Quantity Preprocessing

## Import Raw Quantity Data

```{r}
#| label: quantity-import
quantity_raw <- read_excel(
  source_file,
  sheet = "Exports qties",
  range = "A4:M4559"
)

glimpse(quantity_raw)
```

The quantity worksheet has product and reporter fields plus annual columns from 2014 to 2024. It does not include `GeoRegion`, so it is joined to the value dataset by product, reporter, and year when a combined infographic-ready dataset is needed.

## Clean, Reshape, and Filter Quantity Data

```{r}
#| label: quantity-clean-2024
quantity_clean <- quantity_raw |>
  rename(
    product_description = ProductDescription,
    reporter_name = ReporterName
  ) |>
  pivot_longer(
    cols = matches("^\\d{4}$"),
    names_to = "year",
    values_to = "export_quantity"
  ) |>
  mutate(
    year = as.integer(year),
    product_description = str_squish(product_description),
    reporter_name = str_squish(reporter_name),
    export_quantity = as.numeric(export_quantity)
  )

quantity_2024 <- quantity_clean |>
  filter(year == analysis_year)

head(quantity_2024)
```

The quantity preprocessing mirrors the value workflow: names are standardised, year columns are pivoted into long format, `year` and `export_quantity` are converted to numeric types, labels are cleaned, and then the dataset is filtered to 2024.

## Quantity Validation

```{r}
#| label: quantity-validation-2024
quantity_validation_2024 <- tibble(
  check = c(
    "Rows in raw quantity data",
    "Columns in raw quantity data",
    "Rows after long transformation",
    "Rows retained for 2024",
    "Missing 2024 quantity values",
    "Negative 2024 quantity values",
    "Duplicate 2024 product-reporter rows"
  ),
  result = c(
    nrow(quantity_raw),
    ncol(quantity_raw),
    nrow(quantity_clean),
    nrow(quantity_2024),
    sum(is.na(quantity_2024$export_quantity)),
    sum(quantity_2024$export_quantity < 0, na.rm = TRUE),
    quantity_2024 |>
      count(product_description, reporter_name) |>
      filter(n > 1) |>
      nrow()
  )
)

quantity_validation_2024
```

This confirms whether the 2024 quantity slice is complete enough for infographic use. Because the worksheet mixes physical quantities and some USD-based product groups, quantity values should not be aggregated across all products as if they share a single unit.

## Construct 2024 Quantity Analysis Dataset

```{r}
#| label: quantity-analysis-dataset
quantity_analysis_2024 <- quantity_2024 |>
  filter(
    reporter_name != "World",
    !product_description %in% aggregate_products
  )

glimpse(quantity_analysis_2024)
```

The same aggregate reporter and product rows are removed from the quantity analysis dataset. This keeps the quantity and value datasets aligned for the infographic workflow.

# 2024 Export Quantity EDA

## Quantity Missing and Zero Values

```{r}
#| label: quantity-zero-summary-2024
quantity_zero_summary_2024 <- quantity_analysis_2024 |>
  summarise(
    observations = n(),
    missing_values = sum(is.na(export_quantity)),
    zero_values = sum(export_quantity == 0, na.rm = TRUE),
    zero_percent = zero_values / observations * 100
  )

quantity_zero_summary_2024
```

Zero quantity values are retained as valid reported values. Missing values would be handled separately, but the validation table checks whether they occur in the selected 2024 slice.

## Quantity Summary Statistics

```{r}
#| label: quantity-summary-2024
quantity_summary_2024 <- quantity_analysis_2024 |>
  summarise(
    mean = mean(export_quantity, na.rm = TRUE),
    median = median(export_quantity, na.rm = TRUE),
    q1 = quantile(export_quantity, 0.25, na.rm = TRUE),
    q3 = quantile(export_quantity, 0.75, na.rm = TRUE),
    max = max(export_quantity, na.rm = TRUE),
    sd = sd(export_quantity, na.rm = TRUE)
  )

quantity_summary_2024
```

These statistics describe the spread of the reported 2024 quantity measure. They are useful for quality checks, but they should not be interpreted as a single global physical total because product units may differ.

## Quantity Categorical Coverage

```{r}
#| label: quantity-coverage-2024
quantity_coverage_2024 <- quantity_analysis_2024 |>
  summarise(
    product_groups = n_distinct(product_description),
    reporters = n_distinct(reporter_name)
  )

quantity_coverage_2024
```

The quantity coverage table shows how many products and reporters are available in the supplementary 2024 quantity data.

## Top Quantity Product Groups

```{r}
#| label: quantity-top-products-2024
quantity_top_products_2024 <- quantity_analysis_2024 |>
  group_by(product_description) |>
  summarise(
    total_reported_quantity = sum(export_quantity, na.rm = TRUE),
    reporters = n_distinct(reporter_name),
    .groups = "drop"
  ) |>
  arrange(desc(total_reported_quantity)) |>
  slice_head(n = 15)

quantity_top_products_2024
```

```{r}
#| label: quantity-top-products-plot-2024
ggplot(
  quantity_top_products_2024,
  aes(
    x = reorder(product_description, total_reported_quantity),
    y = total_reported_quantity
  )
) +
  geom_col(fill = "#5c7c99") +
  coord_flip() +
  labs(
    title = "Top 15 Product Groups by Reported 2024 Export Quantity",
    x = NULL,
    y = "Reported export quantity"
  ) +
  theme_minimal()
```

This chart helps identify large reported quantity categories in 2024. It should be interpreted alongside the value view rather than treated as a direct replacement, because some product groups may use different reporting units.

## Value vs Quantity Coverage Check

```{r}
#| label: value-quantity-coverage-comparison-2024
value_pairs_2024 <- value_2024 |>
  distinct(product_description, reporter_name)

quantity_pairs_2024 <- quantity_2024 |>
  distinct(product_description, reporter_name)

coverage_comparison_2024 <- tibble(
  check = c(
    "Product-reporter pairs in value data",
    "Product-reporter pairs in quantity data",
    "Pairs available in both datasets",
    "Pairs only in value data",
    "Pairs only in quantity data"
  ),
  result = c(
    nrow(value_pairs_2024),
    nrow(quantity_pairs_2024),
    inner_join(
      value_pairs_2024,
      quantity_pairs_2024,
      by = c("product_description", "reporter_name")
    ) |>
      nrow(),
    anti_join(
      value_pairs_2024,
      quantity_pairs_2024,
      by = c("product_description", "reporter_name")
    ) |>
      nrow(),
    anti_join(
      quantity_pairs_2024,
      value_pairs_2024,
      by = c("product_description", "reporter_name")
    ) |>
      nrow()
  )
)

coverage_comparison_2024
```

The value and quantity worksheets do not need to have identical coverage. This comparison makes the difference explicit before the team decides whether to combine or compare measures in later visualisations.

# Export Cleaned Datasets

```{r}
#| label: export-cleaned-datasets
dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)

infographic_2024 <- full_join(
  value_analysis_2024,
  quantity_analysis_2024,
  by = c("product_description", "reporter_name", "year")
) |>
  mutate(
    has_export_value = !is.na(export_value_usd_millions),
    has_export_quantity = !is.na(export_quantity)
  ) |>
  select(
    year,
    geo_region,
    reporter_name,
    product_description,
    export_value_usd_millions,
    export_quantity,
    has_export_value,
    has_export_quantity
  ) |>
  arrange(reporter_name, product_description)

value_output_file <- file.path(
  export_dir,
  "agriculture_export_values_2024_clean.csv"
)
quantity_output_file <- file.path(
  export_dir,
  "agriculture_export_quantities_2024_clean.csv"
)
infographic_output_file <- file.path(
  export_dir,
  "agriculture_exports_2024_infographic_ready.csv"
)

write_csv(value_analysis_2024, value_output_file)
write_csv(quantity_analysis_2024, quantity_output_file)
write_csv(infographic_2024, infographic_output_file)

export_manifest <- tibble(
  file = c(value_output_file, quantity_output_file, infographic_output_file),
  rows = c(
    nrow(value_analysis_2024),
    nrow(quantity_analysis_2024),
    nrow(infographic_2024)
  ),
  description = c(
    "Cleaned 2024 export value data with aggregate rows removed.",
    "Cleaned 2024 export quantity data with aggregate rows removed.",
    "Joined 2024 value and quantity data for infographic design."
  )
)

export_manifest
```

The exported CSV files are saved in `data/processed`. The value and quantity files preserve the cleaned single-measure datasets, while the infographic-ready file combines both measures into one table. The boolean fields `has_export_value` and `has_export_quantity` show whether each product-reporter pair is available in each source worksheet.

# 2024 EDA Notes

The 2024-focused workflow makes the analysis year explicit, preserves reproducible preprocessing, and avoids accidental mixing of historical and current-year patterns. The main infographic dataset is `infographic_2024`, which combines cleaned export values and reported export quantities after removing aggregate rows. Regional comparisons should use the value fields because only the value worksheet provides `GeoRegion`. Quantity-based comparisons should be framed carefully because product groups may use different reporting units.
