---
title: "Assignment Report"
author: "Team Orchid"
date: ""
format:
  html:
    toc: true
    number-sections: true
---

# Roles and Contribution

| Member | Student ID | Contribution |
| ------ | ---------- | -------------|
| Quan Yi | 2401058 | Performed EDA for the "Exports qties" sheet |
| Shi Min | 2401793 | - imported data and did basic exploratory data analysis to determine if the dataset needs to be cleaned before visualisation |
| Aiman | <add your student id> | <add your contribution> |
| Subhashini | <add your student id> | <add your contribution>|
| Jabar  | <add your student id> | <add your contribution> |
| Min Xuan | <add your student id> | <add your contribution> |

# Setup
```{r}
#| label: qmd-setup
#| message: false
library(tidyverse)
library(readxl) # for excel files
```

```{r}
#| label: raw-data-imports
analysis_year <- 2024
source_file <- "data/export_of_agriculture_products_data.xlsx"
export_dir <- "data/processed"
```

The code above defines the necessary libraries, source file path, output directory, and target year for the analysis.

<!-- Include Export Values EDA qmd -->
{{< include _export-values-eda.qmd >}}

<!-- Include Export Quantity EDA qmd -->
{{< include _quantity-eda-2024.qmd >}}

<!-- Include Data Preprocessing qmd -->
{{< include _2024_data_preprocessing.qmd >}}

<!-- Include Bar Chart qmd -->
{{< include _bar-chart-visualisation.qmd >}}