---
title: "Assignment Report"
author: "Team Orchid"
date: ""
format:
  html:
    toc: true
    number-sections: true
---

# Roles and Contribution

| Member | Student ID | Contribution |
| ------ | ---------- | -------------|
| Quan Yi | 2401058 | Performed EDA for the "Exports qties" sheet |
| Shi Min | 2401793 | - imported data and did basic exploratory data analysis to determine if the dataset needs to be cleaned before visualisation |
| Aiman | <add your student id> | <add your contribution> |
| Subhashini | <add your student id> | <add your contribution>|
| Jabar  | <add your student id> | <add your contribution> |
| Min Xuan | <add your student id> | <add your contribution> |

# Setup
```{r}
#| label: qmd-setup
#| message: false
library(tidyverse)
library(readxl) # for excel files
```

```{r}
#| label: raw-data-imports
analysis_year <- 2024
source_file <- "data/export_of_agriculture_products_data.xlsx"
export_dir <- "data/processed"
```

The code above defines the necessary libraries, source file path, output directory, and target year for the analysis.

<!-- Include Export Values EDA qmd -->
{{< include _export-values-eda.qmd >}}

<!-- Include Export Quantity EDA qmd -->
{{< include _quantity-eda-2024.qmd >}}

# Data Preprocessing

```{r}
#| label: data-preprocessing

north_america_infographic_countries <- c(
  "United States", "Canada", "Mexico"
)

south_america_infographic_countries <- c(
  "Brazil", "Argentina", "Chile", "Peru"
)

asia_infographic_countries <- c(
  "China", "Indonesia", "India", "Thailand", "Malaysia", "Vietnam", "Singapore"
)

europe_infographic_countries <- c(
  "European Union", "United Kingdom", "Russian Federation", "Ukraine", "Switzerland"
) # Missing: France, Netherlands, Italy, Germany, Spain, Poland, Belgium, Ireland

oceania_infographic_countries <- c(
  "Australia", "New Zealand"
)

africa_infographic_countries <- c(
  "South Africa"
)

middle_east_infographic_countries <- c(
  "Turkiye"
)

all_target_countries <- c(
  north_america_infographic_countries,
  south_america_infographic_countries,
  asia_infographic_countries,
  europe_infographic_countries,
  oceania_infographic_countries,
  africa_infographic_countries,
  middle_east_infographic_countries
)

value_w32 <- value_2024 |> 
  filter(
    reporter_name != "World",
    product_description != "AoA HS proxy"
  ) |> 
  group_by(country = reporter_name, geo_region) |> 
  summarise(export_value = sum(export_value_usd_millions, na.rm = TRUE), .groups = "drop")

quantity_w32 <- quantity_2024 |> 
  group_by(country) |> 
  summarise(export_quantity = sum(quantity, na.rm = TRUE), .groups = "drop")

unified_infographic_data <- value_w32 |> 
  left_join(quantity_w32, by = "country") |> 
  mutate(
    display_entity = ifelse(
      country %in% all_target_countries, 
      country, 
      paste("Rest of", geo_region)
    )
  ) |> 
  group_by(display_entity, geo_region) |> 
  summarise(
    Total_Value_2024 = sum(export_value, na.rm = TRUE),
    Total_Quantity_2024 = sum(export_quantity, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  select(display_entity, geo_region, Total_Value_2024, Total_Quantity_2024)

sorted_infographic_data <- unified_infographic_data |>
  arrange(desc(Total_Value_2024))

sorted_infographic_data
```

# Exporting data


```{r}
#| label: export-processed-data

if (!dir.exists(export_dir)) {
  dir.create(export_dir, recursive = TRUE)
  message("Created new directory: ", export_dir)
}

csv_filename <- paste0("2024_value_and_quantity_export", analysis_year, ".csv")
export_path <- file.path(export_dir, csv_filename)
write_csv(sorted_infographic_data, export_path)

cat("Success! Processed data successfully exported to:", export_path, "\n")


```