---
title: "Assignment Report"
author: "Team Orchid"
date: ""
format:
  html:
    toc: true
  # pptx:
---

# Roles and Contribution

| Member | Student ID | Contribution |
| ------ | ---------- | -------------|
| Quan Yi | <add your student id> | <add your contribution> |
| Shi Min | 2401793 | - imported data and did basic exploratory data analysis to determine if the dataset needs to be cleaned before visualisation |
| Aiman | <add your student id> | <add your contribution> |
| Subhashini | <add your student id> | <add your contribution>|
| Jabar  | <add your student id> | <add your contribution> |
| Min Xuan | <add your student id> | <add your contribution> |

# Setup
```{r}
#| label: qmd-setup
#| message: false
library(tidyverse)
library(readxl) # for excel files
```

# Data Import
```{r}
#| label: data-source-metadata
source_file <- "data/export_of_agriculture_products_data.xlsx"

excel_sheets(source_file)
```

```{r}
#| label: source-attribution
source_notes <- read_excel(
  source_file,
  sheet = "Export values",
  range = "A4841:A4841",
  col_names = FALSE
)

source_notes
```

The dataset was imported from `export_of_agriculture_products_data.xlsx`, specifically the `Export values` worksheet. The workbook also contains overview, product definition, and export quantity sheets, but only export value data was used for this analysis. The first three rows contain workbook title/unit metadata, so the data table begins at row 4. The selected range `A4:N4839` keeps only `ProductDescription`, `ReporterName`, `GeoRegion`, and annual export value fields from 2014 to 2024.

All annual values in the `Export values` worksheet are measured in USD millions. Therefore, the data represents monetary export value rather than physical export quantity, allowing comparisons across products, reporters, regions, and years using a common monetary unit.

# Data Cleaning and Transformation

The raw dataset is stored in wide format, where each year is represented as a separate column. For reproducible analysis and visualisation, the data was transformed into long format so that each row represents one product-reporter-region-year observation.

```{r}
#| label: clean-transform-data
data_raw <- read_excel(
  source_file,
  sheet = "Export values",
  range = "A4:N4839" # Limit import to only relevant data range
)

data_clean <- data_raw |>
  rename(
    product_description = ProductDescription,
    reporter_name = ReporterName,
    geo_region = GeoRegion
  ) |>
  pivot_longer(
    cols = matches("^\\d{4}$"), # Select year columns named with exactly four digits, e.g. 2014
    names_to = "year",
    values_to = "export_value_usd_millions"
  ) |>
  mutate(
    year = as.integer(year),
    product_description = str_squish(product_description),
    reporter_name = str_squish(reporter_name),
    geo_region = str_squish(geo_region),
    export_value_usd_millions = as.numeric(export_value_usd_millions)
  )

head(data_clean)
```

The cleaning process performed the following actions:

1. renamed variables into a consistent snake case format
2. converted the dataset from wide to long format
3. converted `year` into an integer
4. ensured `export_value_usd_millions` is numeric
5. Standardised product, reporter and region labels by removing unnecessary whitespace using `str_squish()`

## Validation Checks
```{r}
#| label: data-validation-checks
validation_checks <- tibble(
  check = c(
    "Rows in raw data",
    "Columns in raw data",
    "Rows after transformation",
    "Missing export values",
    "Negative export values",
    "Duplicate product-reporter-year rows"
  ),
  result = c(
    nrow(data_raw),
    ncol(data_raw),
    nrow(data_clean),
    sum(is.na(data_clean$export_value_usd_millions)),
    sum(data_clean$export_value_usd_millions < 0),
    data_clean |>
      count(product_description, reporter_name, year) |>
      filter(n > 1) |>
      nrow()
  )
)
validation_checks
```

The validation checks confirms that the raw dataset contains **4,835 observations and 14 variables**. After transformation, the cleaned dataset contains **53,185 observations** product-reporter-region-year observations covering 2014–2024. No missing values or duplicate product-reporter-year records were detected. However, **57 negative values** were identified and investigated separately.

## Negative Values
```{r}
#| label: negative-values
negative_values <- data_clean |>
  filter(export_value_usd_millions < 0) |>
  arrange(export_value_usd_millions)

negative_values
unique(negative_values$product_description)
unique(negative_values$reporter_name)
unique(negative_values$geo_region)
```

The dataset contains 57 negative values affecting 21 reporters across 6 region. All occur under the derived `Other AoA` product category. The workbook defines `Other AoA` as a residual category calculated by subtraction, so negative results may occur when its component estimates differ. These values were retained because they are derived source values rather than data-entry errors.

## Removal of Aggregate Rows
```{r}
#| label: aggregate-products
aggregate_products <- c(
  "All W32 groups",
  "AoA HS proxy",
  "Other AoA"
)

data_analysis <- data_clean |>
  filter(
    reporter_name != "World",
    !product_description %in% aggregate_products
  )
```

Aggregate reporter and product rows were excluded from general summary calculations to avoid double-counting values already represented by individual reporters and product groups. They remain available in `data_clean` for analyses specifically focused on official aggregate totals.


# Data Exploration

After cleaning and validation, exploratory checks were performed on the cleaned dataset to understand

1. Data Types
2. Zero Values
3. Numerical Distributions
4. Categorical Coverage
5. Region Coverage
6. Reporter Coverage

## Data Types
```{r}
#| label: eda-data-types
glimpse(data_analysis)
```

The structure of the final analytical dataset was examined using `glimpse()`. It contains **48,037 observations and five variables**. `product_description`, `reporter_name`, and `geo_region` are character variables, `year` is an integer, and `export_value_usd_millions` is numeric. These data types are appropriate for grouping observations and analysing export values over time.

## Zero Values

```{r}
#| label: zero-value-checks
zero_summary_by_year <- data_analysis |>
  group_by(year) |>
  summarise(
    observations = n(),
    zero_values = sum(export_value_usd_millions == 0, na.rm = TRUE),
    zero_percent = zero_values / observations * 100,
    .groups = "drop"
  )

zero_summary_by_year
```

No missing values were detected in the imported range, so no imputation was applied. However, zero export values were retained and treated as valid observations because they likely represent no recorded exports for a product-reporter-year combination rather than missing data.

## Summary Statistics
```{r}
#| label: eda-numerical-data-summary
summary_by_year <- data_analysis |>
  group_by(year) |>
  summarise(
    total_export_value = sum(export_value_usd_millions),
    mean = mean(export_value_usd_millions),
    median = median(export_value_usd_millions),
    q1 = quantile(export_value_usd_millions, 0.25),
    q3 = quantile(export_value_usd_millions, 0.75),
    max = max(export_value_usd_millions),
    sd = sd(export_value_usd_millions),
    .groups = "drop"
  )

summary_by_year
```

This summary provides insights into the central tendency and variability of export value across different years. Since export value are likely skewed, median and quartiles are more informative than the mean alone. The maximum value is also included to identify the scale of dominant exporters or product groups.

## Categorical Coverage

```{r}
#| label: categorical-coverage
categorical_summary <- data_analysis |>
  summarise(
    product_groups = n_distinct(product_description),
    reporters = n_distinct(reporter_name),
    regions = n_distinct(geo_region)
  )

categorical_summary
```

## Region Coverage
```{r}
#| label: regional-coverage
regional_coverage <- data_analysis |>
  distinct(reporter_name, geo_region) |>
  count(geo_region, name = "number_of_reporters") |>
  arrange(desc(number_of_reporters))

regional_coverage
```

## Reporter Coverage
```{r}
#| label: product-coverage
total_reporters <- n_distinct(data_analysis$reporter_name)

product_coverage <- data_analysis |>
  distinct(product_description, reporter_name) |>
  count(
    product_description,
    name = "number_of_reporters"
  ) |>
  mutate(
    coverage_percent = round(number_of_reporters / total_reporters * 100, 1)
  ) |>
  arrange(number_of_reporters)

product_coverage
```

After aggregate rows were removed, the analytical dataset contained 32 product groups, 155 reporters, and seven geographical regions. Each reporter was assigned consistently to one region. However, product coverage was not uniform: the dataset contained 4,367 of the 4,960 possible product-reporter combinations. Therefore, the absence of a product-reporter combination should not automatically be interpreted as zero exports. Regional and product comparisons should account for these differences in source coverage.