---
title: "Preston Curve Lab"
author: "Team Orchid"
date: "2026-06-23"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

# Data Preparation

## Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(ggrepel)
library(scales)
```

For this lab, the new R library packages are `readxl` and `scales`.

## 3.1 Data Import
```{r}
#| label: import-data

gdp <-
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
    read_xls(skip = 3)

life_exp <-
    "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
    read_xls(skip = 3)

pop <-
    "API_SP.POP.TOTL_DS2_en_excel_v2_417154.xls" |>
    read_xls(skip = 3)

gdp
life_exp
pop

```

We import the three World Bank indicator files using `read_xls()`. 

The argument `skip = 3` ignores the first three rows of the `Data` sheet. This is necessary as the workbooks place metadata in those first three rows.

Each of the resulting tibbles - `gdp`, `life_exp`, and `pop`contains the actual column headers such as `Country Name`, `Country Code` and etc.

### 3.1 Verifier Task
```{r}
#| label: verifier-3.1

# Check all three tibbles have required columns
stopifnot("Country Name" %in% names(gdp), "Country Code" %in% names(gdp), "2024" %in% names(gdp))
stopifnot("Country Name" %in% names(life_exp), "Country Code" %in% names(life_exp), "2024" %in% names(life_exp))
stopifnot("Country Name" %in% names(pop), "Country Code" %in% names(pop), "2024" %in% names(pop))

# Check all three tibbles have the same number of rows
stopifnot(nrow(gdp) == nrow(life_exp), nrow(life_exp) == nrow(pop))

# Check 2024 column is not entirely empty in any tibble
stopifnot(any(!is.na(gdp[["2024"]])), any(!is.na(life_exp[["2024"]])), any(!is.na(pop[["2024"]])))

cat("3.1 checks passed\n")
```

We run three sets of checks to determine the success of import.

First we ensure that each tibble contains the three columns subsequent steps rely on: `Country Name`, `Country Code` and `2024`. The `stopifnot` will catch error immediately if any column is missing.

Second, we verify that all three tibbles have the same number of rows. This is crucial as we will later merge them by matching `name` and `code`.

Third, we confirm that the `2024` column is not empty in any tibble.

## 3.2 Select Columns

```{r}
#| label: selecting-data

gdp <- gdp |>
    select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024")

life_exp <- life_exp |>
    select(name = "Country Name", code = "Country Code", life_exp = "2024")

pop <- pop |>
    select(name = "Country Name", code = "Country Code", pop = "2024")

gdp
life_exp
pop
```

`select()` to retain those three columns and rename them to shorter, consistent names.

The 2024 column is explicitly renamed to match each tibble: `gdp_per_cap` for the gdp tibble and etc.

After this step, each tibble has exactly three columns.

## 3.3 Merge Indicator Data

```{r}
#| label: merged-tibbles

countries <- gdp |>
    left_join(life_exp, join_by(name, code)) |>
    left_join(pop, join_by(name, code))

countries
```

`left_join()` argument `join_by()` tells R to match rows where both the country name and country code are identical.

We used `left_join()` rather than `inner_join()` in retaining countries even though it has missing values. 

After both joins, `countries` contains five columns - `name`, `code`, `gdp_per_cap`, `life_exp` and `pop`.

## 3.4 Import the World Bank Regions

```{r}
#| label: import-metadata

metadata <- 
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
    read_xls(sheet = "Metadata - Countries") |>
    select(code = "Country Code", region = "Region")

metadata
```

We import this sheet by passing `sheet = "Metadata - Countries"`, mapping to each country code.

Unlike the `Data` sheet, this sheet does not have metadata rows at the top so there is no need for `skip` argument.

We then pipe the result through `select` and retain only two columns: `Country Code` renamed to `code` and `Region` renamed to `region`.

### 3.4 Verifier Task

```{r}
#| label: verifier-3.4

stopifnot(all(metadata$code %in% countries$code))

# Check converse
missing_from_metadata <- countries$code[!countries$code %in% metadata$code]
cat("Codes in countries but NOT in metadata:", length(missing_from_metadata), "\n")

cat("3.4 checks passed\n")
```

`stopifnot()` to perform checks on the relationship between `metadata` and `countries`.

The first check confirms that every country code in `metadata` also presents in `countries`. Line uses the `%in%` operator to test whether it returns `TRUE` for each code in `metadata` founded somewhere in `countries$code`. `all()` reaffirms that every single one passes.

The second check test the converse: whether every code in `countries` also appears in `metadata`. We compute `missing_from_data` as the subset of `countries$code` values not present in `metadata$code`. It reveals that there is one code appearing in `countries` but absent from `metadata`.

## 3.5 Merge Regions with Indicators

```{r}
#| label: join-metadata

countries <- countries |>
    left_join(metadata, join_by(code))

countries
```

`left_join()` attaches the `region` column from `metadata` to `countries`.

After the join, `countries` gains a sixth column - `region`.

### 3.5 Verifier Task

```{r}
#| label: verifier-3.5

# Row count must be the same before and after the join
rows_after_33 <- nrow(gdp |> left_join(life_exp, join_by(name, code)) |> left_join(pop, join_by(name, code)))
stopifnot(nrow(countries) == rows_after_33)

cat("3.5 checks passed ")
```

We confirm that the join function left the row count of `countries` unchanged.

In recomputing the expected row count independently of the current state of `countries`, replaying the three-way merge and compare its row count. The `stopifnot()` throws an error if the counts differ.

## 3.6 Remove Missing Data

```{r}
#| label: remove-missing-data

countries <- countries |>
    drop_na()

countries
```

Remove every row from `countries` using `drop_na()`.

In the World Bank data sheets, terms like `Arab World` and `Sub-Saharan Africa` seem as if they were countries but they are actually a group statistics rather than specific countries.

## 3.7 Prepare Country Labels

```{r}
#| label: prepare-labels

labeled_countries <- c(
  countries |> slice_max(gdp_per_cap,      n = 10) |> pull(code),
  countries |> slice_min(gdp_per_cap,      n = 10) |> pull(code),
  countries |> slice_max(life_exp, n = 10) |> pull(code),
  countries |> slice_min(life_exp, n = 10) |> pull(code),
  countries |> slice_max(pop,      n = 10) |> pull(code)
)

countries <- countries |>
  mutate(label = if_else(code %in% labeled_countries, code, ""))

countries
```

Compute `labeled_countries` by pulling the `code` columsn from each of the five `slice_max()` or `slice_min()` and concatenating the results with `c()`.

We then use `mutate()` with `if_else()` in adding a `label` column to `countries`. For every row whose `code()` are in `labelled_countries`, the label is set to the three-letter code itself. While the all the other rows with empty strings is set to `""`. 

### 3.7 Verifier Task

```{r}
#| label: verifier-3.7

# Confirm exact set of 7 columns
expected_cols <- c("name", "code", "gdp_per_cap", "life_exp", "pop", "region", "label")
stopifnot(all(expected_cols %in% names(countries)))
stopifnot(ncol(countries) == 7)

# No missing values anywhere
stopifnot(all(complete.cases(countries)))

# Labels are either empty string or a 3-letter country code
stopifnot(all(countries$label == "" | nchar(countries$label) == 3))

cat("3.7 checks passed\n")
```

Run three checks to ensure that `countries` is in the correct format before visualization begins.

First check confirms no extra columns have been accidentally introduced and that all seven expected columns are present via total column count is exactly seven.

Second check confirms there are no missing values using `complete.cases()`. It returns `TRUE` for every row having `NA` in any column. `all(complete.cases())` is `TRUE` only if the whole tibble is clean.

Third check confirms every value in `label` column is either an empty string or a three-letter code.  

# Visualizing the Data
## 4.1 Scale Audit

|  Aesthetic | Variable | Scale decision | Justification |
|-----------|----------|----------------|---------------|
| x | GDP per capita | continuous, log-transformed | GDP per capita is right-skewed; a log₁₀ scale spreads the points more evenly and makes multiplicative differences appear as equal visual distances. |
| y | Life expectancy | continuous, linear | Life expectancy is roughly symmetric and bounded (~50–90 years), so a linear scale faithfully represents differences without distortion. |
| size | Population | continuous, area-proportional | Using area to encode population follows Tufte's area principle: equal data ratios should map to equal area ratios, which the default radius-based scale does not guarantee. |
| color | Region | discrete, qualitative palette | Region is a nominal, unordered categorical variable, so a qualitative palette avoids implying any ordering or magnitude differences between groups. |
| label | Selected country code | direct annotation, no legend | Direct labels on selected points are more efficient than a lookup legend; labelling every country would cause severe overplotting. |

## 4.2 Set Up Basic Plot

```{r}
#| label: basic-plot

basic_plot <-
    ggplot(countries, aes(x = gdp_per_cap, y = life_exp, color=region))+
    geom_point(aes(size = pop), alpha = 0.5)

basic_plot
```

Using `ggplot()` to create a bubble chart with four aesthetics mapped from `countries`. 

On the x-axis is GDP per capita and life expectancy on the y-axis, with region mapped to colour and population mapped to bubble size within `geom_point()`.

Setting `alpha = 0.5` makes the bubbles semi-translucent so that overlapping points in dense regions of the chart are still visible.

## 4.3 Add Country Labels

```{r}
#| label: add-label

basic_plot <- basic_plot + geom_text_repel(
    aes(label = label),
    show.legend = FALSE,
    size = 3,
    max.overlaps = 20,
  )

basic_plot
```

`geom_texxt_repel()` automatically nudges labels away from each other avoiding overlap. 

Configure `show.legend = FALSE` ensures that legends is protected from redundant entry.

Only countries with `label` column not empty receive a visible label whereas others pass an empty string and silently skipped.

## 4.4 Add Trend Curve

```{r}
#| label: add-curve

basic_plot <- basic_plot + 
    geom_smooth(aes(weight = pop), method = "loess", color = "gray")

basic_plot

```

`geom_smooth()` with `aes(weight = pop)` produce a LOESS smooth curve.

Highly populous countries like China and India have a greater influence on the shape of the fitted curve than small nations with similar income levels.

### 4.4 Verifier Task

```{r}
#| label: verifier-4.4

#basic_plot +
  #geom_smooth(method = "loess", color = "red", linetype = "dashed", se = FALSE)

```

I visually confirmed that the two curves differ, specifically noticing how the population-weighted curve shifts to account for highly populous countries like China and India.

The weighted curve noticeably shifts in the middle-income range reaffirming that `weight = pop` is active and resourceful.

## 4.5 Add Annotations

```{r}
#| label: add-annotation

basic_plot <- basic_plot + labs(
        x = "GDP per capita ($ Purchasing Power Parity 2021)",
        y = "Life Expectancy (years)",
        title = "Life Expectancy vs. GDP per Capita (2024)",
        caption = "Source: World Bank (2024)",
    )

basic_plot
```

According to the World Bank (under Metadata - Indicators), the GDP indicator is adjusted using purchasing power parities (PPPs) and expressed in constant 2021 international dollars. This adjustment controls for differences in local price levels, allowing for an accurate cross-country comparison of true economic output and standard of living. 

`labs()` to add axis labels, title and a data source caption. 

## 4.6 Change x-Axis Scale

```{r}
#| label: change-x-axis

basic_plot <- basic_plot +
    scale_x_log10( 
        breaks = breaks_log(n = 6),
        minor_breaks = NULL,
        labels = label_comma()
    )

basic_plot
```

Applying `scale_x_log10()`, compresses and spread the points more evenly across the plot.

`breaks_log()` places breaks that are reader-friendly.

`minor_breaks = NULL` suppreses minor gridlines reducing visual clutter.

`label_comma()` are used for the ticks so as to be readable rather than scientific notation.

b. Using `label_currency()` (which prints $ 1,000) would be misleading here because the unit is a "common currency" adjusted for PPP, not nominal US dollars. A reader could incorrectly interpret the values as market-rate USD. Since our x-axis title already explicitly mentions the unit ($ Purchasing Power Parity 2021), using `label_comma()` is the better choice as it presents the numerical values cleanly without introducing confusing, repetitive currency symbols.

## 4.7 Reason about the Choice of Logarithmic Axis Ticks

For the `breaks_log(n=6)` approach, the ticks use familiar base-10 multiples (1 and 3), making it highly intuitive to estimate the GDP figure because each tick is a simple multiple of a power of ten. 

For the successive doublings approach, moving one tick to the right always means the country is exactly twice as wealthy. This makes it immediately intuitive to compare relative income ratios between countries.

## 4.8 Change Size Scale

```{r}
#| label: change-size-scale

basic_plot <- basic_plot +
    scale_size_area(
        breaks = 10^(6:9),
        labels = label_number(scale_cut = cut_short_scale()),
        max_size = 15
    )


basic_plot
```

`scale_size_area` maps bubble area directly to the data value so if a country is ten times more populous appears exactly ten times larger than area.

breaks are configured at 1M, 10M, 100M and 1B

`label_number()` with `short_cut_scale` formats into readable suffixed readings. 

### 4.8 Verifier Task

```{r}
#| label: verifier-4.8
# Check the area principle by computing expected diameter ratios between legend keys.
# Expected diameter ratio between successive keys = sqrt(10) ≈ 3.16, NOT 10.

breaks <- 10^(6:9)
expected_ratios <- sqrt(breaks[-1] / breaks[-length(breaks)])
cat("Expected diameter ratios between successive legend keys (should all be ~3.16):\n")
print(round(expected_ratios, 3))

stopifnot(all(abs(expected_ratios - sqrt(10)) < 0.01))
cat("Area principle verified: diameter ratios ≈ sqrt(10) ≈ 3.16 between each successive key.\n")
```

I physically checked the size legend on the generated plot. The largest bubble is approximately 3 times wider than the next size down, which aligns with the expected ~3.16 ratio and confirms the area principle is functioning correctly.

## 4.9 Change Color Scale

```{r}
#| label: change-color-scale

basic_plot <- basic_plot +
    scale_color_brewer(palette = "Dark2")

basic_plot

```

`scale_color_brewer()` makes distinct hues straightforward to visually segregate the seven world regions across the plot.

### 4.9 Verifier Task

```{r}
#| label: verifier-4.9
# Confirm the number of color legend keys matches the number of distinct regions in countries.
n_regions <- n_distinct(countries$region)
cat("Distinct regions in countries:", n_regions, "\n")
cat("Check that the color legend in the plot has exactly", n_regions, "keys.\n")

# Confirm Dark2 has enough colors 
if (n_regions > 8) {
  warning("Dark2 only supports 8 colors but there are ", n_regions, " regions — some may be recycled or NA.")
} else {
  cat("Dark2 palette has sufficient colors for", n_regions, "regions.\n")
}
```

I confirmed that the number of colour legend keys matches the number of distinct regions in `countries`. No colour collision as the `Dark2` palette supports up to eight colours with the dataset containing seven distinct regions. Shows exactly seven keys, one per region thru the rendered colour legend.

## 4.10 Refine Guides

```{r}
#| label: refine-guides

basic_plot <- basic_plot +
    guides(
        color = guide_legend(
            order = 1,
            override.aes = list(alpha = 1, size = 4),
            title = "Region"
        ),
        size = guide_legend(
            order = 2,
            reverse = TRUE,
            title = "Population"
        )
    )

basic_plot
```

`guides()` is to arrange both legends in decree of order. 

color legend is placed first, `order = 1` because readers use region to interpret the plot. 

`alpha = 1` meaning its symbols are rendered at full capacity so each colour is clearly visible.

size legend is placed second, `order = 2` with reverse = TRUE` making the largest bubble to be at the top.

a.  The color legend should appear at the top over the size legend. When we look at a chart, we notice the distinct color groupings instantly before we are able to calculate the relative sizes. Therefore, placing the color legend at the top aligns with this visual hierarchy

b. Reversing the size legend makes the population mapping easier to read because we tend to associate the 'top' with larger quantities, similar to how the y-axis increases upwards. However, `ggplot2` defaults to placing the smallest values at the top of the legend. By reversing it, the largest population sizes are placed at the top, making the visual scale much more intuitive to interpret at a glance.

c. When we plotted the points with `alpha = 0.5` to reduce overplotting, the legend automatically inherits this transparency. This makes the legend keys faint and difficult to see clearly. Overriding the legend to be fully opaque with `alpha = 1` ensures distinct color contrast in the key, making it much easier for the reader to identify and map the regions at a glance.

## 4.11 Compare Axis Guides

```{r}
#| label: move-x-axis

temp_plot <- basic_plot +
    guides(x = guide_axis(title = NULL, position = "top"))

temp_plot
```

`guides_axis()` creates a temporary copy of the plot while the x-axis moved to the top.

Moving the x-axis to the top would worsen the presentation of the Preston curve. Currently, our title is "Life Expectancy vs. GDP per Capita (2024)", while the x-axis is "GDP per capita ($ Purchasing Power Parity 2021)". The title in its current state cannot serve as a substitute for the x-axis title, as removing the unit of measurement ($ PPP 2021) obscures vital context and alters the clarity of the visualization.

Additionally, people typically expect the independent variable (GDP per Capita) to be positioned at the bottom. Moving the x-axis to the top forces the viewer's eyes to constantly track upward to check economic values while looking at data points concentrated lower down, breaking conventional reading habits.

Therefore, we reject this change.


### 4.1 Verifier Task

```{r}
#| label: verifier-4.1

# 1. x-axis is log-transformed
has_log_x <- any(sapply(basic_plot$scales$scales, function(s) {
  "x" %in% s$aesthetics &&
    inherits(s, "ScaleContinuousPosition") &&
    !is.null(s$trans) &&
    grepl("log", s$trans$name)
}))
stopifnot(has_log_x)
cat("x-axis: log scale confirmed\n")

has_custom_y <- any(sapply(basic_plot$scales$scales, function(s) "y" %in% s$aesthetics))
if (has_custom_y) warning("y-axis has a custom scale — confirm it is still linear")
cat("y-axis: linear (default) confirmed\n")

has_size_area <- any(sapply(basic_plot$scales$scales, function(s) "size" %in% s$aesthetics))
stopifnot(has_size_area)
cat("size: area-proportional scale present\n")

has_color_scale <- any(sapply(basic_plot$scales$scales, function(s) "colour" %in% s$aesthetics))
stopifnot(has_color_scale)
cat("color: discrete qualitative scale present\n")
 
has_repel <- any(sapply(basic_plot$layers, function(l) inherits(l$geom, "GeomTextRepel")))
stopifnot(has_repel)
repel_layer <- Filter(function(l) inherits(l$geom, "GeomTextRepel"), basic_plot$layers)[[1]]
stopifnot(isFALSE(repel_layer$show.legend))
cat("label: geom_text_repel with no legend confirmed\n")

cat("4.1 scale audit fully implemented\n")
```

`basic_plot$scales$scales` is used by the x-axis to inspect log transformation. 

`sapply()` is to loop over each element in the list of scale and apply a checking function which returns a `TRUE`/`FALSE`. 

`any()` collapses the vector to a single `TRUE` if and only if at least one element passes the check.

`grepl()` searches the transformation name string of each scale for the subset string `log`.

`inherits()` verifies that one of the plot layers is a `geom_text_repel()` call and checks that its `show.legend` is `FALSE` meaning no redundant label legend entry appears.

Each finding is wrapped with `stopifnot()` confirming log x-axis, linear y-axis, area-proportional size scale, qualitative color scale and direct text labels are implemented.

## 4.12 Reason about the Data

From the plot, we are able to observe an sigmodial curve, illustrating a general trend where a higher GDP per capita leads to a higher life expectancy. However, we observe that Nigeria (NGA) sits drastically below the curve, and has one of the lowest life expectancy despite maintaining a mid-range GDP per capita. 

According to Rédaction Africanews (2025), we are able to relate the low life expectancy to persistent health system challenges, which includes a high burden of infectious diseases, low access to quality care, and regular health worker strikes. Therefore, while Nigeria possesses a relatively strong economic output, they seem to have limited medical infrastructure to effectively combat preventable illnesses, leading to a low life expectancy.

Source: https://www.africanews.com/2025/10/07/nigerias-life-expectancy-ranked-lowest-globally/

# AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Useful AI Response

**In the response :**
```r
#| eval: false
The Core Approach
The task requires you to union five sets of countries (top/bottom 10 by GDP, top/bottom 10 by life expectancy, top 10 by population), then label only those countries with their code, and everyone else with "".

r
#| label: add-labels

labeled_countries <- c(
  countries |> slice_max(gdp,      n = 10) |> pull(code),
  countries |> slice_min(gdp,      n = 10) |> pull(code),
  countries |> slice_max(life_exp, n = 10) |> pull(code),
  countries |> slice_min(life_exp, n = 10) |> pull(code),
  countries |> slice_max(pop,      n = 10) |> pull(code)
)

countries <- countries |>
  mutate(label = if_else(code %in% labeled_countries, code, ""))

countries
```

In multiple instances where the team did not know how to approach a question, a simple prompt regarding a question was able to give accurate and usable code snippets that was directly usable within the lab sheet despite the lack of context given. This greatly helped the team understand what the question is asking for and even help us completing the task at hand.


### Suboptimal AI Response


**From the prompt :**
```r
#| eval: false
I tried a similar version to what you suggested but got the following output:
Warning: Using `size` aesthetic for lines was deprecated in ggplot2 3.4.0.
ℹ Please use `linewidth` instead.
`geom_smooth()` using formula = 'y ~ x'
Warning: The following aesthetics were dropped during statistical transformation: size.
ℹ This can happen when ggplot fails to infer the correct grouping structure in
the data.
ℹ Did you forget to specify a `group` aesthetic or to convert a numerical
variable into a factor?
```

The AI was not useful in this certain question and was not every helpful in diagnosing the issue. The team did not know how to further question the AI and was able to solve it on our own.

### Reflection

Our team has consistently used Perplexity (Claude Sonnet 4.6) throughout the past lab sessions till present. In several instances, with gardual prompts that help build the understanding of the project has been able to give more accurate answers compared to uploading the entire lab instruction brief for answers. Simple prompts produced accurate, directly usable code snippets that helped us understand what a task was asking for and complete it efficiently. As always, the assistance in understanding functions, formulas, reasoning and more makes it much easier to for the team to complete this lab but, it is only complimentable due to the fact that certain team members are knowledgable in their code and is able to verify the work of what AI suggests.

There will be chances (although rare) that we get answers from the AI that are not very usable nor applicable as mentioned. Therefore, re-reading the task requirements is always a must to ensure the team does not go off-course. This reminded us to always evaluate AI output critically rather than applying it blindly.
