---
title: "Preston Curve Lab"
author: "Team Orchid"
date: "2026-06-23"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

# Data Preparation

## Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(ggrepel)
library(scales)
```

## 3.1 Data Import
```{r}
#| label: import-data

gdp <-
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
    read_xls(skip = 3)

life_exp <-
    "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
    read_xls(skip = 3)

pop <-
    "API_SP.POP.TOTL_DS2_en_excel_v2_417154.xls" |>
    read_xls(skip = 3)

gdp
life_exp
pop

```

### 3.1 Verifier Task
```{r}
#| label: verifier-3.1

# Check all three tibbles have required columns
stopifnot("Country Name" %in% names(gdp), "Country Code" %in% names(gdp), "2024" %in% names(gdp))
stopifnot("Country Name" %in% names(life_exp), "Country Code" %in% names(life_exp), "2024" %in% names(life_exp))
stopifnot("Country Name" %in% names(pop), "Country Code" %in% names(pop), "2024" %in% names(pop))

# Check all three tibbles have the same number of rows
stopifnot(nrow(gdp) == nrow(life_exp), nrow(life_exp) == nrow(pop))

# Check 2024 column is not entirely empty in any tibble
stopifnot(any(!is.na(gdp[["2024"]])), any(!is.na(life_exp[["2024"]])), any(!is.na(pop[["2024"]])))

cat("3.1 checks passed\n")
```

## 3.2 Select Columns

```{r}
#| label: selecting-data

gdp <- gdp |>
    select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024")

life_exp <- life_exp |>
    select(name = "Country Name", code = "Country Code", life_exp = "2024")

pop <- pop |>
    select(name = "Country Name", code = "Country Code", pop = "2024")

gdp
life_exp
pop
```

## 3.3 Merge Indicator Data

```{r}
#| label: merged-tibbles

countries <- gdp |>
    left_join(life_exp, join_by(name, code)) |>
    left_join(pop, join_by(name, code))

countries
```

## 3.4 Import the World Bank Regions

```{r}
#| label: import-metadata

metadata <- 
    "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
    read_xls(sheet = "Metadata - Countries") |>
    select(code = "Country Code", region = "Region")

metadata
```

### 3.4 Verifier Task

```{r}
#| label: verifier-3.4

stopifnot(all(metadata$code %in% countries$code))

# Check converse
missing_from_metadata <- countries$code[!countries$code %in% metadata$code]
cat("Codes in countries but NOT in metadata:", length(missing_from_metadata), "\n")

cat("3.4 checks passed\n")
```

## 3.5 Merge Regions with Indicators

```{r}
#| label: join-metadata

countries <- countries |>
    left_join(metadata, join_by(code))

countries
```

### 3.5 Verifier Task

```{r}
#| label: verifier-3.5

# Row count must be the same before and after the join
rows_after_33 <- nrow(gdp |> left_join(life_exp, join_by(name, code)) |> left_join(pop, join_by(name, code)))
stopifnot(nrow(countries) == rows_after_33)

cat("3.5 checks passed ")
```

## 3.6 Remove Missing Data

```{r}
#| label: remove-missing-data

countries <- countries |>
    drop_na()

countries
```

## 3.7 Prepare Country Labels

```{r}
#| label: prepare-labels

labeled_countries <- c(
  countries |> slice_max(gdp_per_cap,      n = 10) |> pull(code),
  countries |> slice_min(gdp_per_cap,      n = 10) |> pull(code),
  countries |> slice_max(life_exp, n = 10) |> pull(code),
  countries |> slice_min(life_exp, n = 10) |> pull(code),
  countries |> slice_max(pop,      n = 10) |> pull(code)
)

countries <- countries |>
  mutate(label = if_else(code %in% labeled_countries, code, ""))

countries
```

### 3.7 Verifier Task

```{r}
#| label: verifier-3.7

# Confirm exact set of 7 columns
expected_cols <- c("name", "code", "gdp_per_cap", "life_exp", "pop", "region", "label")
stopifnot(all(expected_cols %in% names(countries)))
stopifnot(ncol(countries) == 7)

# No missing values anywhere
stopifnot(all(complete.cases(countries)))

# Labels are either empty string or a 3-letter country code
stopifnot(all(countries$label == "" | nchar(countries$label) == 3))

cat("3.7 checks passed\n")
```

# Visualizing the Data
## 4.1 Scale Audit

|  Aesthetic | Variable | Scale decision | Justification |
|-----------|----------|----------------|---------------|
| x | GDP per capita | continuous, log-transformed | GDP per capita is right-skewed; a log₁₀ scale spreads the points more evenly and makes multiplicative differences appear as equal visual distances. |
| y | Life expectancy | continuous, linear | Life expectancy is roughly symmetric and bounded (~50–90 years), so a linear scale faithfully represents differences without distortion. |
| size | Population | continuous, area-proportional | Using area to encode population follows Tufte's area principle: equal data ratios should map to equal area ratios, which the default radius-based scale does not guarantee. |
| color | Region | discrete, qualitative palette | Region is a nominal, unordered categorical variable, so a qualitative palette avoids implying any ordering or magnitude differences between groups. |
| label | Selected country code | direct annotation, no legend | Direct labels on selected points are more efficient than a lookup legend; labelling every country would cause severe overplotting. |

## 4.2 Set Up Basic Plot

```{r}
#| label: basic-plot

basic_plot <-
    ggplot(countries, aes(x = gdp_per_cap, y = life_exp, color=region))+
    geom_point(aes(size = pop), alpha = 0.5)

basic_plot
```

## 4.3 Add Country Labels

```{r}
#| label: add-label

basic_plot <- basic_plot + geom_text_repel(
    aes(label = label),
    show.legend = FALSE,
    size = 3,
    max.overlaps = 20,
  )

basic_plot
```

## 4.4 Add Trend Curve

```{r}
#| label: add-curve

basic_plot <- basic_plot + 
    geom_smooth(aes(weight = pop), method = "loess", color = "gray")

basic_plot

```

### 4.4 Verifier Task

```{r}
#| label: verifier-4.4

#basic_plot +
  #geom_smooth(method = "loess", color = "red", linetype = "dashed", se = FALSE)

```

I visually confirmed that the two curves differ, specifically noticing how the population-weighted curve shifts to account for highly populous countries like China and India.

## 4.5 Add Annotations

```{r}
#| label: add-annotation

basic_plot <- basic_plot + labs(
        x = "GDP per capita ($ Purchasing Power Parity 2021)",
        y = "Life Expectancy (years)",
        title = "Life Expectancy vs. GDP per Capita (2024)",
        caption = "Source: World Bank (2024)",
    )

basic_plot
```

According to the World Bank (under Metadata - Indicators), the GDP indicator is adjusted using purchasing power parities (PPPs) and expressed in constant 2021 international dollars. This adjustment controls for differences in local price levels, allowing for an accurate cross-country comparison of true economic output and standard of living. 

## 4.6 Change x-Axis Scale

```{r}
#| label: change-x-axis

basic_plot <- basic_plot +
    scale_x_log10( 
        breaks = breaks_log(n = 6),
        minor_breaks = NULL,
        labels = label_comma()
    )

basic_plot
```

b. Using `label_currency()` (which prints $ 1,000) would be misleading here because the unit is a "common currency" adjusted for PPP, not nominal US dollars. A reader could incorrectly interpret the values as market-rate USD. Since our x-axis title already explicitly mentions the unit ($ Purchasing Power Parity 2021), using `label_comma()` is the better choice as it presents the numerical values cleanly without introducing confusing, repetitive currency symbols.

## 4.7 Reason about the Choice of Logarithmic Axis Ticks

For the `breaks_log(n=6)` approach, the ticks use familiar base-10 multiples (1 and 3), making it highly intuitive to estimate the GDP figure because each tick is a simple multiple of a power of ten. 

For the successive doublings approach, moving one tick to the right always means the country is exactly twice as wealthy. This makes it immediately intuitive to compare relative income ratios between countries.

## 4.8 Change Size Scale

```{r}
#| label: change-size-scale

basic_plot <- basic_plot +
    scale_size_area(
        breaks = 10^(6:9),
        labels = label_number(scale_cut = cut_short_scale()),
        max_size = 15
    )


basic_plot
```

### 4.8 Verifier Task

```{r}
#| label: verifier-4.8
# Check the area principle by computing expected diameter ratios between legend keys.
# Expected diameter ratio between successive keys = sqrt(10) ≈ 3.16, NOT 10.

breaks <- 10^(6:9)
expected_ratios <- sqrt(breaks[-1] / breaks[-length(breaks)])
cat("Expected diameter ratios between successive legend keys (should all be ~3.16):\n")
print(round(expected_ratios, 3))

stopifnot(all(abs(expected_ratios - sqrt(10)) < 0.01))
cat("Area principle verified: diameter ratios ≈ sqrt(10) ≈ 3.16 between each successive key.\n")
```

I physically checked the size legend on the generated plot. The largest bubble is approximately 3 times wider than the next size down, which aligns with the expected ~3.16 ratio and confirms the area principle is functioning correctly.

## 4.9 Change Color Scale

```{r}
#| label: change-color-scale

basic_plot <- basic_plot +
    scale_color_brewer(palette = "Dark2")

basic_plot

```

### 4.9 Verifier Task

```{r}
#| label: verifier-4.9
# Confirm the number of color legend keys matches the number of distinct regions in countries.
n_regions <- n_distinct(countries$region)
cat("Distinct regions in countries:", n_regions, "\n")
cat("Check that the color legend in the plot has exactly", n_regions, "keys.\n")

# Confirm Dark2 has enough colors 
if (n_regions > 8) {
  warning("Dark2 only supports 8 colors but there are ", n_regions, " regions — some may be recycled or NA.")
} else {
  cat("Dark2 palette has sufficient colors for", n_regions, "regions.\n")
}
```

## 4.10 Refine Guides

```{r}
#| label: refine-guides

basic_plot <- basic_plot +
    guides(
        color = guide_legend(
            order = 1,
            override.aes = list(alpha = 1, size = 4),
            title = "Region"
        ),
        size = guide_legend(
            order = 2,
            reverse = TRUE,
            title = "Population"
        )
    )

basic_plot
```

a.  The color legend should appear at the top over the size legend. When we look at a chart, we notice the distinct color groupings instantly before we are able to calculate the relative sizes. Therefore, placing the color legend at the top aligns with this visual hierarchy

b. Reversing the size legend makes the population mapping easier to read because we tend to associate the 'top' with larger quantities, similar to how the y-axis increases upwards. However, `ggplot2` defaults to placing the smallest values at the top of the legend. By reversing it, the largest population sizes are placed at the top, making the visual scale much more intuitive to interpret at a glance.

c. When we plotted the points with `alpha = 0.5` to reduce overplotting, the legend automatically inherits this transparency. This makes the legend keys faint and difficult to see clearly. Overriding the legend to be fully opaque with `alpha = 1` ensures distinct color contrast in the key, making it much easier for the reader to identify and map the regions at a glance.

## 4.11 Compare Axis Guides

```{r}
#| label: move-x-axis

temp_plot <- basic_plot +
    guides(x = guide_axis(title = NULL, position = "top"))

temp_plot
```

Moving the x-axis to the top would worsen the presentation of the Preston curve. Currently, our title is "Life Expectancy vs. GDP per Capita (2024)", while the x-axis is "GDP per capita ($ Purchasing Power Parity 2021)". The title in its current state cannot serve as a substitute for the x-axis title, as removing the unit of measurement ($ PPP 2021) obscures vital context and alters the clarity of the visualization.

Additionally, people typically expect the independent variable (GDP per Capita) to be positioned at the bottom. Moving the x-axis to the top forces the viewer's eyes to constantly track upward to check economic values while looking at data points concentrated lower down, breaking conventional reading habits.

Therefore, we reject this change.


### 4.1 Verifier Task

```{r}
#| label: verifier-4.1

# 1. x-axis is log-transformed
has_log_x <- any(sapply(basic_plot$scales$scales, function(s) {
  "x" %in% s$aesthetics &&
    inherits(s, "ScaleContinuousPosition") &&
    !is.null(s$trans) &&
    grepl("log", s$trans$name)
}))
stopifnot(has_log_x)
cat("x-axis: log scale confirmed\n")

has_custom_y <- any(sapply(basic_plot$scales$scales, function(s) "y" %in% s$aesthetics))
if (has_custom_y) warning("y-axis has a custom scale — confirm it is still linear")
cat("y-axis: linear (default) confirmed\n")

has_size_area <- any(sapply(basic_plot$scales$scales, function(s) "size" %in% s$aesthetics))
stopifnot(has_size_area)
cat("size: area-proportional scale present\n")

has_color_scale <- any(sapply(basic_plot$scales$scales, function(s) "colour" %in% s$aesthetics))
stopifnot(has_color_scale)
cat("color: discrete qualitative scale present\n")
 
has_repel <- any(sapply(basic_plot$layers, function(l) inherits(l$geom, "GeomTextRepel")))
stopifnot(has_repel)
repel_layer <- Filter(function(l) inherits(l$geom, "GeomTextRepel"), basic_plot$layers)[[1]]
stopifnot(isFALSE(repel_layer$show.legend))
cat("label: geom_text_repel with no legend confirmed\n")

cat("4.1 scale audit fully implemented\n")
```

## 4.12 Reason about the Data

From the plot, we are able to observe an sigmodial curve, illustrating a general trend where a higher GDP per capita leads to a higher life expectancy. However, we observe that Nigeria (NGA) sits drastically below the curve, and has one of the lowest life expectancy despite maintaining a mid-range GDP per capita. 

According to Rédaction Africanews (2025), we are able to relate the low life expectancy to persistent health system challenges, which includes a high burden of infectious diseases, low access to quality care, and regular health worker strikes. Therefore, while Nigeria possesses a relatively strong economic output, they seem to have limited medical infrastructure to effectively combat preventable illnesses, leading to a low life expectancy.

Source: https://www.africanews.com/2025/10/07/nigerias-life-expectancy-ranked-lowest-globally/

# AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Useful AI Response

**In the response :**



### Suboptimal AI Response