---
title: "Preston Curve Lab"
author: "Team Orchid"
date: "2026-06-23"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

# Data Preparation

## Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(ggrepel)
library(scales)
```

## 3.1 Data Import
```{r}
#| label: data-import
gdp_raw <- read_xls(
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls",
  sheet = "Data",
  skip = 3
)

gdp_raw

life_raw <- read_xls(
  "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls",
  sheet = "Data",
  skip = 3
)

life_raw

pop_raw <- read_xls(
  "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls",
  sheet = "Data",
  skip = 3
)

pop_raw
```

### 3.1 Verifier Task
```{r}
#| label: 3.1-verifier-task-col
gdp_col_verify <- select(gdp_raw, "Country Name", "Country Code", "2024")
gdp_col_verify

life_col_verify <- select(life_raw, "Country Name", "Country Code", "2024")
life_col_verify

pop_col_verify <- select(pop_raw, "Country Name", "Country Code", "2024")
pop_col_verify
```



```{r}
#| label: 3.1-verifier-task-rows
nrow(gdp_raw)
ncol(gdp_raw)

nrow(life_raw)
ncol(life_raw)

nrow(pop_raw)
ncol(pop_raw)
```

```{r}
#| label: 3.1-verifier-task-na
gdp_nas <- sum(is.na(gdp_raw["2024"]))
gdp_nas

life_nas <- sum(is.na(life_raw["2024"]))
life_nas

pop_nas <- sum(is.na(pop_raw["2024"]))
pop_nas
```

## 3.2 Select Columns
```{r}
#| label: select-columns
gdp_2024 <- rename(
  gdp_col_verify,
  name = "Country Name",
  code = "Country Code",
  gdp_per_cap = "2024"
)
gdp_2024

life_2024 <- rename(
  life_col_verify,
  name = "Country Name",
  code = "Country Code",
  life_exp = "2024"
)
life_2024


pop_2024 <- rename(
  pop_col_verify,
  name = "Country Name",
  code = "Country Code",
  pop = "2024"
)
pop_2024
```

## 3.3 Merge Indicator Data
```{r}
#| label: merge-data
countries <- inner_join(gdp_2024, life_2024, join_by(name, code)) |>
  inner_join(pop_2024, join_by(name, code))

countries
```

## 3.4 Import the World Bank Regions
```{r}
#| label: metadata-countries
metadata <- read_xls(
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls",
  sheet = "Metadata - Countries"
) |>
  select(code = "Country Code", region = "Region")

metadata
```

### 3.4 Verifier Task
```{r}
#| label: 3.4-verifier
tibble(
  check = c(
    "All metadata country codes are present in countries",
    "All countries country codes are present in metadata"
  ),
  result = c(
    all(metadata$code %in% countries$code),
    all(countries$code %in% metadata$code)
  )
)
```

## 3.5 Merge Regions with Indicators
```{r}
#| label: merge-region
countries <- left_join(countries, metadata, join_by(code))
countries
```

### 3.5 Verifier Task
Looking at [initial `countries`](#merge-indicator-data), it has 266 rows and 5 columns. The newly merged `countries` shown above, produced a 266 rows and 6 columns.

## 3.6 Remove Missing Data
```{r}
#| label: remove-missing-data
countries_dropped <- countries |> drop_na()
countries_dropped
```

## 3.7 Prepare Country Labels

### 3.7 Verifier Task


# Visualizing the Data
## 4.1 Scale Audit


## 4.2 Set Up Basic Plot


## 4.3 Add Country Labels


## 4.4 Add Trend Curve


### 4.4 Verifier Task


## 4.5 Add Annotations


## 4.6 Change x-Axis Scale

Harder to read and relate to as GapMinder uses double scaling, e.g. 64,000 vs 100,000. There is no right or wrong. Log scale's distance between each ticks may be slightly difference


## 4.7 Reason about the Choice of Logarithmic Axis Ticks


## 4.8 Change Size Scale


### 4.8 Verifier Task


## 4.9 Change Color Scale


### 4.9 Verifier Task


## 4.10 Refine Guides


## 4.11 Compare Axis Guides


## 4.12 Reason about the Data


# AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Useful AI Response

**In the response :**



### Suboptimal AI Response