---
title: "Penguins-Lab"
author: "Team Orchid"
date: "2026-05-26"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | 
|---|---|
| Quan Yi | Coder |
| Shi Min | Navigator |
| Aiman | AI Liaison |
| Subhashini | Verifier |
| Jabar | Reporter |
| Min Xuan | Constructive Critic |

## Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(ggbeeswarm)
```


## 3.1 Data Preparation

```{r}
#| label: data-preparation

penguins <- palmerpenguins::penguins

nrow(penguins)
ncol(penguins)

head(penguins)
tail(penguins)
glimpse(penguins)

penguins_bill <- penguins |>
    select(species, bill_len = bill_length_mm , bill_dep = bill_depth_mm) |>
    drop_na()

nrow(penguins_bill)
```

a. Size of the dataset: 344 rows and 8 columns

b. Preliminary Exploration: We noticed that there were a few data with missing data points when we use `glimpse()`

c. Data Cleaning: After removing all the missing values using `drop_na()`, it removed 2 rows and now we are left with 342 rows.

```{r}
#| label: verifier-3.1

```

## 3.2 Distribution of Bill Length

### 3.2a Ungrouped Bill-Length Histogram
```{r}
#| label: bill-length-histogram

caption_source <- "Palmer Station Antarctica LTER and K. Gorman, 2020"

gg_hist_bins <- function(w) {
    ggplot(penguins_bill, aes(x = .data$bill_len)) +
        geom_histogram(
            binwidth = w,
            fill = "lightyellow",
            color = "darkblue"
        ) +
        geom_rug(
            aes(y = 0),
            position = position_jitter( height = 0 ),
            alpha = 0.5,
            sides = "b"
        ) +
        labs(
            x = "Bill Length (mm)",
            y = "Count",
            title = "Bill Length Histogram",
            caption = caption_source
        )
}

gg_hist_bins(1)

```


```{r}
#| label: verifier-3.2a


```

### 3.2b Bin Width

```{r}
#| label: histogram-bin-width

gg_hist_bins(0.8)

gg_hist_bins(1.5)

```

Comparing the binwidth of `0.8` and `1.5`, more of the data points are "joined together", which leads it to have a higher count at the same part of the histogram. 

We would use a binwidth of 0.8 in the report as it helps to show a clearer distinction as of how many penguins are of `x` bill length.

### 3.2c Histogram Interpretation

There are multiple peaks in different parts of the histogram, where there are 3 peaks when we use a binwidth of `0.8`. This helps support that the distribution is not unimodal otherwise, there would only be one clean peak

### 3.2d Raw Point Plot by Species

```{r}
#| label: raw-point-plot

labs_len_by_species <- labs(
    x = "Bill Length (mm)",
    y = "Species",
    caption = caption_source
)

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$species)) +
    geom_point() + 
    labs_len_by_species +
    labs(title = "Raw Point Plot")

```

It was noted that the Adelie penguins have a shorter bill length amongst the 3 penguin species, Gentoo has a entry with a bill length of almost 60mm and Chinstrap seems to have the fewest data point among the 3 species.

### 3.2e Jitter Plot

```{r}
#| label: jitter-point-plot

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$species)) +
    geom_jitter(
        alpha = 0.5,
        width = 0,
        height = 0.25
    ) + 
    labs_len_by_species +
    labs(title = "Jitter Point Plot")

```

It helps to show the density of each penguin at their respective bill length vertically so that we are able to observe that there are mutliple penguins at the same bill length which we could not see previously

### 3.2f Beeswarm Plot

```{r}
#| label: beeswarm-plot

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$species)) +
    geom_beeswarm( 
        cex = 1.5
    ) +
    labs_len_by_species +
    labs(title = "Beeswarm Plot")


```

The beeswarm plot seems to not use a random variable like jitter and stacks each instance of the penguin of `x` bill length instead.

### 3.2g Pros and Cons

One advantage of using beeswarm plot is for showing the exact shape of the distribution as it makes it structurally visible where most points exist

One disadvantage of Beeswarm plots is that it can tend to give a misleading density from a viewer's perception.The width of the swarm depend on point size and spacing settings, not just data density.

### 3.2h Faceted Bill-Length Histogram

```{r}
#| label: faceted-histogram

ggplot(penguins_bill, aes(x = .data$bill_len)) +
    geom_histogram(
        bins = 20,
        fill = "lightyellow",
        color = "darkblue"
    ) +
    geom_rug(
        aes(y = 0),
        position = position_jitter( height = 0 ),
        alpha = 0.5,
        sides = "b"
    ) +
    facet_wrap(vars(species), ncol = 1) +
    geom_vline(aes(xintercept = mean(bill_len)), color = "red") +
    labs_len_by_species +
    labs(title = "Faceted Bill Length Histogram")

```

### 3.2i Violin-with-Box Plot
```{r}
#| label: violin-with-box-plot

ggplot(penguins_bill, aes(bill_len, species)) +
    geom_violin() +
    geom_boxplot(aes(fill = species), width = 0.2) +
    labs_len_by_species +
    labs(title = "Violin and Box Plot")

```

### 3.2j Why Combine Violin and Box Plot

With only box plot, we know whether there are any outliers, and we know the median and mean of the data, but we do not know the "shape" of the data. 

With only violin plot, we know approximately how dense the data is across the penguin species but we do not know any more than that

Using both, we know how dense the data is across the penguin species, where the mean and median are, and whether there are any outliers, being the Gentoo species.

## 3.3 Distribution of Bill Depth

### 3.3a Bill-Depth Histogram

```{r}
#| label: bill-depth-histogram

gg_depth_bins <- function(w) {
    ggplot(penguins_bill, aes(x = .data$bill_dep)) +
        geom_histogram(
            binwidth = w,
            fill = "lightyellow",
            color = "darkblue"
        ) +
        geom_rug(
            aes(y = 0),
            position = position_jitter( height = 0 ),
            alpha = 0.5,
            sides = "b"
        ) +
        labs(
            x = "Bill Depth (mm)",
            y = "Count",
            title = str_c("Bill Depth Histogram"),
            caption = caption_source
        )
}

gg_depth_bins(1)
```

There seems to be lesser variation when it comes to Bill depth where the range is only from 12.5 to ~ 22mm while Bill length range is much larger, ranging from `30~` to `~60`.

### 3.3b Faceted Bill-Depth Histogram
```{r}
#| label: faceted-bill-depth-histogram

global_mean_dep <- mean(penguins_bill$bill_dep)

ggplot(penguins_bill, aes(x = .data$bill_dep)) +
    geom_histogram(
        binwidth = 0.5,
        fill = "lightyellow",
        color = "darkblue"
    ) +
    facet_wrap(vars(species), ncol = 1) +
    geom_rug(
        aes(y = 0),
        position = position_jitter( height = 0 ),
        alpha = 0.5,
        sides = "b"
    ) +
    geom_vline(aes(xintercept = global_mean_dep), color = "red") +
    labs_len_by_species +
    labs(
        x = "Bill Depth (mm)",
        title = "Faceted Bill Depth Histogram"
    )
```

### 3.3c Comparison of Bill Length and Bill Depth

Using the Bill Length, we are able to differentiate the Adelie penguins from the Gentoo and Chinstrap penguins as the Adelie penguins have a shorter bill length compared to the other two penguin species. However, we are unable to differentiate the Chinstrap and Gentoo penguins based on Bill Length alone

Using Bill Depth, we are able to differentiate the Gentoo penguins as they have a shorter bill depth compared to the Adelie and Chinstrap species. 

Therefore the two measurements help to separate different species 

```{r}
#| label: verifier-3.3c


```

## 3.4 Bivariate View: Bill Depth vs Bill length

### 3.4a Scatter plot
```{r}
#| label: scatter-plot-with-linear-model

labs_len_by_bills <- labs(
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    caption = caption_source
)

lm_plot <-
    ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$bill_dep)) +
        geom_point() +
        geom_smooth(
            method = "lm"
        ) +
        labs_len_by_bills

lm_plot +
    labs( title = "Scatter plot with linear model")

```

### 3.4b Interpretation of Pooled trend

When species information is ignored and all penguins are pooled together, the regression line shows a negative relationship between bill length and bill depth which is; penguins with longer bills have shallower bills. 

The confidence band has a "hourglass" shape and never crosses or approaches a horizontal slope, this indicates that the negative trend is statistically significant.

### 3.4c Exact Overplotting

```{r}
#| label: overplotting

coord_counts <- penguins_bill |>
  count(bill_len, bill_dep)

hidden_points <- sum(coord_counts$n) - nrow(coord_counts)
max_stack <- max(coord_counts$n)

hidden_points
max_stack

```

No, exactly coincident coordinates are not the only cause of overplotting. While perfectly identical coordinates cause points to sit directly on top of one another, there are other causes such as:

- Data Rounding: Even if data points are not strictly identical, they may be rounded to the same precision (e.g., both measuring 45.1 mm).

- Large Dataset Size: In very large datasets, even if data points have distinct values, the sheer number of points relative to the resolution of the screen (the "drawing space") ensures that points will inevitably overlap.

### 3.4d Mitigation of Overplotting

```{r}
#| label: jitter-overplot

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$bill_dep)) +
    geom_jitter(
    alpha = 0.5,
    width = 0,
    height = 0.1
    ) + 
    geom_smooth(
        method = "lm"
    ) +
    labs_len_by_bills +
    labs( title = "Jittered Scatter plot")

```

```{r}
#| label: count-overplot

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$bill_dep)) +
    geom_smooth(method = "lm") +
    geom_count(alpha = 0.5) +
    scale_size_area() +
    labs_len_by_bills +
    labs(title = "Count-symbol Plot") 

```

### 3.4e Assessment of Scatter Plot Variants

For a journal submission, we would submit (d.ii) the count-symbol plot. Jittering uses a random variable into the data positions to prevent overlap, this causes a different plot to generate each time it runs. 

geom_count() provides an accurate and reproducible visual representation of the data density. It allows the user to clearly see both the underlying trend and the exact frequency of overlapping data coordinates.

### 3.4f Grouped Scatter Plot

```{r}
#| label: color-scatter-plot

ggplot(penguins_bill, aes(x = .data$bill_len, y = .data$bill_dep, color = .data$species)) +
    geom_smooth(method = "lm") +
    geom_count(alpha = 0.5) +
    scale_size_area() +
    labs_len_by_bills +
    labs(
        title = "Colored Count-symbol Plot",
        color = "Species"
    ) 

```

After mapping the `species` to colour, it reveals that the negative trend observed previously was misleading.

It originally showed that there was a negative correlation, where as bill length increases, the bill depth decreases. However, for each individual species, there is actually a positive correlation between the bill length and the bill depth.

The statistical phenomenon observed is called **Simpson's Paradox**.

This occurs when a trend appears in different groups of data but disappears or reverses when these groups are combined.

## 3.5 Summarizing the Insights gained

Yes, bill length and bill depth are highly effective at distinguishing the three Pygoscelis species. As shown in the colored count-symbol plot (`3.4f`), the species cluster into distinct regions, Gentoo penguins occupy a unique space characterized by lower bill depths, while Adelie and Chinstrap penguins show clear separation based on bill length, with Chinstrap penguins tending toward longer bills and Adelie penguins exhibiting shorter bills.

Ignoring species leads to a misleading view because it obscures the true relationship between these two traits. When the data was combined, as seen in `3.4a`, the regression line suggests a negative correlation, implying that as bill length increases, bill depth decreases. However, the colored plot (`3.4f`) reveals that within each species, the relationship is actually positive.

This illustrates Simpson's Paradox, a statistical phenomenon where a trend observed within individual groups reverses when the data is aggregated. By ignoring the species as a grouping variable, the pooled analysis creates a trend that reflects the differences in mean bill dimensions between the species rather than the actual relationship between the measurements within those populations.