---
title: "Week 10 Lab: Choropleth Maps of Country-Level Statistics"
author: "Team Orchid"
date: "2026-07-07"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

## 3.1 Identify Suitable Fill Variables for Choropleth Maps
| Variable | Variable Type | Suitable as fill color? Why or why not?  | Would the result be a choropleth map? Why or why not?  |
| ------ | ---- | -----------------------| -----------------------|
| GDP per capita | Quantitative - Intensive  | Yes - it's a rate, doesn't distort with merged polygons  | Yes as it's a rate and not raw total Eg. Computing the weighted average of GDP per capita rather than just adding all GDP per-capita values together |
| Life expectancy | Quantitative - Intensive  | Yes - same reasoning as GDP  | Yes and also the same reasoning with GDP |
| Region | Categorical - Nominal | No not really - No ordering and just labels  | No as it is more of a categorical map  |
| Population | Quantitative - Extensive | No - misleads by area size when summing up  | No as it is suited best for bubble size  |

## 3.2 Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(mapview)
library(patchwork)
library(readxl)
library(rmapshaper)
library(scales)
library(sf)
library(shadowtext)
```

Utilized new R packages `mapview` for interactive geometry inspection, `patchwork` for mixing the two final maps, `rmapshaper` for polygon simplification, `sf` for handling spatial data and `shadowtext` for halo-outlined map labels. 

## 3.3 Import Statistics
```{r}
#| label: import-data

gdp <-
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
  read_xls(skip = 3)

life_exp <-
  "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
  read_xls(skip = 3)

gdp
life_exp
```

`gdp` imports the GDP workbook plus skipping the 3 header rows of metadata.

`life_exp` does the same but instead imports the life expectancy workbooks.

```{r}
#| label: selecting-and-merging-data

gdp <- gdp |>
  select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024")

life_exp <- life_exp |>
  select(name = "Country Name", code = "Country Code", life_exp = "2024")

country_stats <- gdp |>
  left_join(life_exp, join_by(name, code))

country_stats
```

`select(...)` keeps only `Country Name`, `Country Code`, and the `2024` column from each sheet plus renaming them to `name`, `code` and `gdp_per_cap`/`life_exp` respectively.

`left_join(...)` merges the GDP and life expectancy tables taking into account missing values on the matching name/code pair.

```{r}
#| label: import-metadata

metadata <-
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
  read_xls(sheet = "Metadata - Countries") |>
  select(code = "Country Code", region = "Region")
```

`metadata` tibble stores `Metadata-Countries` sheet by calling `read_xls()`.

`select(...)` keeps only `code` and `region`.

```{r}
#| label: join-metadata

country_stats <- country_stats |>
  left_join(metadata, join_by(code))

country_stats
```

`left_join(...)` attaches the region column to `country_stats`.

```{r}
#| label: handle-missing-data

country_stats <- country_stats |>
  drop_na(region) |>
  select(name, code, gdp_per_cap, life_exp)

country_stats
```

`drop_na()` discards rows with empty region.

`select(...)` drops the `region` column leaving `name`, `code`, `gdp_per_cap` and `life_exp`.


### 3.3 Verifier Task
```{r}
#| label: verifier-task-3.3
names(country_stats)
nrow(country_stats)
```

Verifier Comment: The country_states do show the 4 column names and nrow(countries) do return 217.

`names()` lists the column names of `country_stats` to acknowledge that it holds exactly `name`, `code`, `gdp_per_cap` and `life_exp`.

`nrow()` returns the number of rows.

## 3.4 Import Boundaries

```{r}
#| label: import-boundaries
#| cache: true

country_boundaries <-
  "WB_GAD_ADM0_complete.geojson" |>
  read_sf() |>
  select(code = ISO_A3, status = WB_STATUS)

country_boundaries
```

`read_sf()` reads GeoJSON boundary file into an `sf` object.

`select(...)` renames and keeps only the ISO country code and legal status columns.

### 3.4 Verifier Task
```{r}
#| label: verifier-task-3.4
inherits(country_boundaries, "sf")
unique(st_geometry_type(country_boundaries))
st_is_longlat(country_boundaries)
```

Verifier Comment: Countries_boundaries is an `sf` object, and all geometries are of type `MULTIPOLYGON`, and the coordinate reference system is longitude/latitude.

`inherits(..., sf)` inspects whether `country_boundaries` belongs to `sf` class.

`st_geometry_type()` returns every feature's geometry type. Wrapping it in `unique()` establishes every feature is exclusively multipolygon.

`st_is_longlat()` inspects whether the longitude/latitude is used in the coordinate reference system which returns `TRUE`.

## 3.5 Validate and Repair Geometry

```{r}
#| label: counting-geometry

sum(!st_is_valid(country_boundaries))
```

There are 5 invalid geometries within country_boundaries.

`!st_is_valid()` returns a logical vector flagging incomplete geometries. Wrapping it with `sum()` counts how many are incomplete.

```{r}
#| label: preparing-geometry

country_boundaries <- st_make_valid(country_boundaries)
sum(!st_is_valid(country_boundaries))
```

`st_make_valid()` managed to repair the invalid geometries and this is confirmed by counting the number of invalid geometries with `!st_is_valid()` again, which returns 0.

### 3.5 Verifier Task
```{r}
#| label: verifier-task-3.5
sum(!st_is_valid(country_boundaries))
nrow(country_boundaries)
unique(st_geometry_type(country_boundaries))
```

Verifier Comment: Geometries did repair perfectly. There is 0 invalid geometries remain and the row count is unchanged.

`!st_is_valid()` re-checks every geometry and summing the negated result with `sum()`.

`nrow()` verifies that the repair process will not change the row count.

## 3.6 Simplify Polygons

### 3.6a Determining number of boundary points

```{r}
#| label: count-points

npts(country_boundaries)
```

`npts()` is used to count across all geometries the total number of boundary points.

### 3.6b Determine number of polygons (Singapore)
```{r}
#| label: count-polygons-singapore

singapore <- country_boundaries |>
  filter(code == "SGP")

singapore

no_of_polygons <- st_cast(singapore, to = 'POLYGON')
no_of_polygons
```

There are a total of 20 polygons in Singapore.

`filter()` isolates the specific country's feature.

`st_cast(...)` splits multipolygon into individual component polygons to be counted.

### 3.6c Simplify boundary

```{r}
#| label: simplify-boundaries

country_boundaries <- ms_simplify(
  country_boundaries,
  keep_shapes = TRUE,
  keep = 0.0015
)

npts(country_boundaries)
country_boundaries
```

Using `keep_shapes = TRUE` allows us to retain at least one polygon for every country when using `ms_simplify`.

To retain approximately 10,000 boundary points, we reduce the argument `keep` from a default value of `0.05` to `0.0015`.

`ms_simplify()` minimizes the boundary point count to approximately 10,000 meanwhile `keep_shapes()` guarantees at least one polygon with every feature.

### 3.6d Determine number of polygons #2 (Singapore)

```{r}
#| label: simplified-polygons-singapore

singapore <- country_boundaries |>
  filter(code == "SGP")

no_of_polygons <- st_cast(singapore, to = 'POLYGON')
no_of_polygons
```

After the simplification, Singapore is left with only one polygon.

`filter()` + `st_cast(...)` confirms the intended country has just 1 polygon. 

### 3.6 Verifier Task
```{r}
#| label: verifier-task-3.6
nrow(country_boundaries)
any(st_is_empty(country_boundaries))
sum(!st_is_valid(country_boundaries))
mapview::npts(country_boundaries)
```

Verifier Comment: Polygons successfully simplified to ~9,500 points with no empty or invalid geometries introduced.

`nrow()` reaffirms all features survived `ms_simplify()`.

`st_is_empty()` flags any feature whose geometry got simplified to nothing. Wrapping it in `any()` confirms no feature was voided out.

`!st_is_valid()` reaffirms no invalid geometries were introduced during simplification.

`mapview::npts()` re-counts boundary points.

## 3.7 Cure Antimeridian

```{r}
#| label: plot-sf

ggplot(country_boundaries) + geom_sf()
```

`ggplot()` + `geom_sf()` plots the raw world map.

### 3.7b Fixed antimeridian

```{r}
#| label: cure-antimeridian

country_boundaries <- st_wrap_dateline(country_boundaries)
country_boundaries <- country_boundaries |>
  st_cast(country_boundaries, to = 'MULTIPOLYGON')

ggplot(country_boundaries) + geom_sf()
```

`country_boundaries` tibble uses `st_wrap_dateline()` to split polygons which cross the meridian of longitude.

`st_cast(...)` restores a consistent geometry type, re-plotting confirming that the streak is gone.

### 3.7 Verifier Task
```{r}
#| label: verifier-task-3.7
nrow(country_boundaries)

unique(st_geometry_type(country_boundaries))

sum(!st_is_valid(country_boundaries))
```

Verifier Comment: Antimeridian cured safely. All features cast back to MULTIPOLYGON with no invalid shapes.

`nrow()` verifies the antimeridian fix `st_wrap_dateline()` + `st_cast()` left the feature count unchanged.

`unique(st_geometry_type())` confirms every feature is exclusively multipolygon.

`!st_is_valid()` wrapping in `sum()` re-counts that no invalid geometries were introduced by the fix.

## 3.8 Merge features

### 3.8a Find shared non-missing code
```{r}
#| label: find-shared-code

count_code <- count(country_boundaries, code) |>
  filter(n > 1)

count_code
```

`count(...)` with pipe operator `filter()` finds country codes shared by more than one feature.

### 3.8b Verify Information

The code FRA is shared by France and Clipperton Island, an uninhabited coral atoll in the eastern Pacific Ocean, roughly 1,300 miles (2,090 km) southwest of Mexico. It carries France's country code because Clipperton is a French dependency administered directly by the Minister of Overseas France, rather than a separate territory with its own ISO code. Background information: [https://www.britannica.com/place/Clipperton-Island](https://www.britannica.com/place/Clipperton-Island)

### 3.8c Merge features

```{r}
#| label: merge-features

duplicate_countries <- count_code |>
  select(code) |>
  filter(code != "NA") |>
  st_drop_geometry()

dup_codes <- duplicate_countries$code

duplicates_only <- country_boundaries |>
  filter(code %in% dup_codes)

non_duplicates <- country_boundaries |>
  filter(!code %in% dup_codes)

merged_features <- duplicates_only |>
  group_by(code) |>
  summarize(
    status = ifelse(
      any(status == "Member State"),
      "Contains a Member State",
      "Does Not Contain a Member State"
    )
  ) |>
  st_cast("MULTIPOLYGON")

country_boundaries <- bind_rows(non_duplicates, merged_features)

country_boundaries
```

`select()` + `filter()` extracts only the shared codes from `count_code` with `st_drop_geometry()` dropping the geometry column to leave a non-spatial vector of duplicate codes.

`duplicates_only` and `non_duplicates` tibbles splits the data into duplicated and non-duplicated via function call `filter(!code %in% dup_codes)`.

`merge_features` tibble uses `duplicates_only` tibble to merge each duplicate groups into a single feature per code via `group_by()` + `summarize(...)` and setting `status` to `Contains a Member State` or `Does Not Contains a Member State` plus an `st_cast()` function converting the combined geometries back to a consistent type.

`country_boundaries` then deployed `bind_rows(...)` function recombining the merged and untouched rows.

### 3.8 Verifier Task
```{r}
#| label: verifier-task-3.8
nrow(country_boundaries)

codes <- country_boundaries$code[!is.na(country_boundaries$code)]
any(duplicated(codes))

sum(
  country_boundaries$status == "Non-determined legal status area",
  na.rm = TRUE
)

unique(st_geometry_type(country_boundaries))
```

Verifier Comment: Features merged correctly into 264 rows. No duplicate codes exist and all 24 undetermined areas are preserved.

`nrow()` confirms that there are 264 features after merging.

`country_boundaries$code[!is.na(...)]` extracts only the non-missing codes.

`duplicated()` flags any repetition and wrapping it in `any()` which verifies no non-missing code appears more than once.

`sum(...)` counts the undetermined features.

`unique(st_geometry_type(...))` confirms every feature is exclusively multipolygon. 

## 3.9 Join Statistics

```{r}
#| label: join-statistics

countries <- country_boundaries |>
  left_join(country_stats, join_by(code))

countries
```

`countries` tibble uses `left_join()` + `join_by()` attaches `country_stats` to `country_boundaries` keeping every polygon even those with no matching statistics.

### 3.9 Verifier Task
```{r}
#| label: verifier-task-3.9
mismatches <- anti_join(
  country_stats,
  sf::st_drop_geometry(country_boundaries),
  by = "code"
)
print(mismatches)

inherits(countries, "sf")
nrow(countries)
all(c("name", "gdp_per_cap", "life_exp") %in% names(countries))
```

Verifier Comment: Yes , only the Channel Islands mismatch was found and the final sf object contains all the required columns. 

`sf::st_drop_geometry(...)` strips the geometry column from `country_boundaries` tibble. `anti_join(...)` returns rows in `country_stats` whose code has no match and stored in `mismatch()`

`inherits(...)` confirms `countries` is still of `sf` object after the join.

`nrows()` counts the number of rows.

`all(...) %in% ...` confirms the original boundary columns were appended with the three statistics columns.

## 3.10 Compare Projections

### 3.10a Equirectangular Projection

```{r}
#| label: projection-name

st_crs(country_boundaries, parameters = TRUE)[c("Name", "srid")]
```

The name is "WGS 84", which stands for World Geodetic System 84. This is the default projection which maps the longitude to the x-axis and latitude to the y-axis. However, a trait of this projection is that it is a "equirectangular" which do not plot the areas in the correct proportion.

`str_cs(...)` reports the name and code of default coordinate reference system (WGS 84).

### 3.10b Equal Earth Projection

```{r}
#| label: plot-equal-earth

ggplot(country_boundaries) +
  geom_sf() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45)) +
  scale_y_continuous(
    breaks = c(-89.9, -45, 0, 45, 89.9)
  ) +
  coord_sf(crs = "EPSG:8857")
```

`gg_plot()` + `geom_sf()` + `coord_sf()` replots the map with `scale_x_continuos(...)`/`scale_y_continuos(...)` configure graticule breaking at 45 degrees major latitude lines.

A second `coord_sf()` replots the same map for comparison. 

### 3.10c Mollweide Projection
```{r}
#| label: plot-other-projection

ggplot(country_boundaries) +
  geom_sf() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45)) +
  scale_y_continuous(
    breaks = c(-89.9, -45, 0, 45, 89.9)
  ) +
  coord_sf(crs = "ESRI:54009")
```

The Equirectangular projection while straightforward, it significantly distorts areas and shapes. It drastically stretches areas as you move away from the equator toward the poles. As an example, Greenland appears to be the same size as Australia but is only 28% of Australia's land. However the graticule interpretability is high as The grid lines form straight, evenly spaced horizontal and vertical lines, making it the easiest projection for reading raw latitude and longitude coordinates.

The Mollweide projection is an equal-area projection that depicts the pole as single points and the antimeridians merge into each other without any sharp corners at the poles. The area distortion is zero as it preserves relative land areas across the entire globe, correctly depicting e.g. the area of Greenland to Australia. This projection undergoes some shape distortion as continents near the boundary appear slanted or stretched. Lastly, the graticule interpretability isn't as high as the latitude are straight horizontal lines but the longitude curve significantly outwards towards the edge.

The Equal Earth projection like the MollWeide projection preserves the relative area well, leading it to have zero distortion in terms of Area. The shape distortion is relatively low as the map has a flat top and bottom, rather than a cylindrical shape like the Mollweide projection. The graticule interpretability is high as the latitude are straight and equally spaced, and the curvature of the meridians are predictable and balanced.

Overall, the Equal Earth projection is the most suitable for a global choropleth map as while the Mollweide projection also achieves zero distortion in terms of area (Equirectangular completely fails in this regard), Mollweide's shape distortion is relatively worse to the Equal Earth projection, and the graticule interpretability of the Equal Earth projection is better than the Mollweide projection. 

## 3.11 Choropleth Draft

```{r}
#| label: basic-choropleth-map

choropleth_map <- ggplot(countries) +
  geom_sf(aes(fill = gdp_per_cap)) +
  coord_sf(crs = "EPSG:8857")

choropleth_map
```

While there are gray patches to indicate missing values, it is unclear whether it is because of missing gdp_per_cap values or undetermined status areas. The default continuous scale is also unsuitable as due to the outlier values, the midpoint is around ~60,000. This causes a lot of countries to be in the dark blue part of the scale and most of the plot looks like the same dark blue shade. Additionally, a darker color is usually used to represent a higher value, which is not represented by default.

`ggplot()` + `geom_sf(...)` + `coord_sf()` produces a choropleth map which maps GDP per capita to fill color under Equal Earth projection.

## 3.12 Fixing Flaws

### 3.12a Fix scale and color palette

```{r}
#| label: fix-scale-and-colour

choropleth_map <- choropleth_map +
  scale_fill_fermenter(
    palette = "Reds",
    direction = -1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale())
  )

choropleth_map
```

`scale_fill_fermenter(...)` substitutes continuos fill scale with ColorBrewer scale using a persistent break ladder and compact axis labels. Note `direction = 1` reverses the default palette order so darker reds map to lower GDP values. 

### 3.12b Map missing countries

```{r}
#| label: fill-missing-countries

ggplot(countries) +
  geom_sf(aes(fill = gdp_per_cap)) +
  scale_fill_fermenter(
    palette = "Reds",
    direction = 1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  geom_sf(
    data = filter(countries, is.na(gdp_per_cap)),
    aes(color = "Missing"),
    fill = "lightgray"
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    aes(color = "Undetermined"),
    fill = "darkgray"
  ) +
  scale_color_manual(
    values = c("Missing" = "black", "Undetermined" = "black")
  ) +
  guides(
    color = guide_legend(override.aes = list(fill = c("lightgray", "darkgray")))
  ) +
  coord_sf(crs = "EPSG:8857")
```

`ggplot()` + `geom_sf(...)` replots the map from start and this round with the direction of `scale_fill_fermenter(...)` equals 1 so darker reds now correctly map to higher GDP values.

Two additional `geom_sf(...)` layers plot the missing-data countries and undermined status areas as separate layers via `filter(...)` with them mapped to a constant `color` string ("Missing"/"Undetermined").

`scale_color_manual(...)` assigns both constants the same black color and `guides(...)` overrides the legend fill so they display light gray and dark grey swatches respectively.

## 3.13 Two-Panel Choropleth Map
```{r}
#| label: two-panel-choropleth-prep

earth <- st_polygon(
  x = list(
    cbind(c(rep(-180, 181), rep(180, 181), -180), c(-90:90, 90:-90, -90))
  )
) |>
  st_sfc() |>
  st_set_crs(4326) |>
  st_as_sf()

top_40_largest <- countries |>
  mutate(area = as.numeric(st_area(geometry))) |>
  slice_max(order_by = area, n = 40)
```

`st_polygon(...)` + `st_sfc()` + `st_set_crs()` + `st_as_sf()` constructs a rectangular polygon pertaining "earth" and used as a background layer.

`mutate(...)` + `slice_max(...)` calculates each country's area and keeps the 40 largest for map labels.


```{r}
#| label: life-exp-plot

map_life_exp <- ggplot(countries) +
  geom_sf(data = earth, fill = "aliceblue", color = "black", linewidth = 0.5) +
  geom_sf(
    aes(fill = life_exp),
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_fermenter(
    name = "Life expectancy at birth \n(years)",
    palette = "Greens",
    direction = 1,
    breaks = c(60, 65, 70, 75, 80),
    guide = guide_colorsteps(
      position = "top",
      barwidth = unit(15, "lines"),
      barheight = unit(0.8, "lines"),
      title.hjust = 0.5
    )
  ) +
  geom_sf(
    data = filter(countries, is.na(life_exp)),
    fill = "lightgray",
    color = "black",
    linewidth = 0.1
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    fill = "darkgray",
    color = "black",
    linewidth = 0.1
  ) +
  geom_shadowtext(
    data = top_40_largest,
    aes(label = code, geometry = geometry),
    stat = "sf_coordinates",
    bg.color = "white",
    color = "black",
    check_overlap = TRUE
  ) +
  theme_void() +
  theme(
    legend.margin = margin(5, 0, 0, 0),
    legend.frame = element_rect(color = "black"),
    legend.position = "top"
  ) +
  coord_sf(
    crs = "EPSG:8857",
    ylim = c(-56, 84),
    default_crs = 4326
  ) +
  scale_y_continuous(expand = c(0, 0))
```

`map_life_exp` layers the background `earth` polygon while the main `geom_sf(...)` choropleth layer with thin black borders and undertermined status grey layers and `geom_shadowtext(...)` to place outlined country-code labels at each of the 40 largest countries' centroids.

`scale_fill_fermenter(...)` places a horizontal, top-centered binned legend which doubles as the map title.

`scale_color_manual(...)` on the GDP map places the shared "Missing"/"Undetermined" legend keys at the bottom of the figure.

`theme_void()` strips the background grid lines and axis text to minimize non-data ink.

`coord_sf(...)` clips both the maps to populated latitude range while keeping the Equal Earth projection.

`scale_y_continuous(expand = c(0, 0))` helps to keep the crop exactly at 84°N and 56°S

```{r}
#| label: gdp-plot

map_gdp <- ggplot(countries) +
  geom_sf(data = earth, fill = "aliceblue", color = "black", linewidth = 0.5) +
  geom_sf(
    aes(fill = gdp_per_cap),
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_fermenter(
    name = "GDP per capita \n($ Purchasing Power Parity 2021)",
    palette = "Reds",
    direction = 1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale()),
    guide = guide_colorsteps(
      position = "top",
      barwidth = unit(15, "lines"),
      barheight = unit(0.8, "lines"),
      title.hjust = 0.5
    )
  ) +
  geom_sf(
    data = filter(countries, is.na(gdp_per_cap)),
    aes(color = "Missing"),
    fill = "lightgray",
    linewidth = 0.1
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    aes(color = "Undetermined"),
    fill = "darkgray",
    linewidth = 0.1
  ) +
  scale_color_manual(
    name = "",
    values = c("Missing" = "black", "Undetermined" = "black"),
    guide = guide_legend(
      position = "bottom",
      override.aes = list(
        fill = c("Missing" = "lightgray", "Undetermined" = "darkgray"),
        color = "black",
        linewidth = 0.1
      )
    )
  ) +
  geom_shadowtext(
    data = top_40_largest,
    aes(label = code, geometry = geometry),
    stat = "sf_coordinates",
    bg.color = "white",
    color = "black",
    check_overlap = TRUE
  ) +
  theme_void() +
  theme(
    legend.margin = margin(5, 0, 0, 0),
    legend.frame = element_rect(color = "black"),
  ) +
  coord_sf(
    crs = "EPSG:8857",
    ylim = c(-56, 84),
    default_crs = 4326
  ) +
  scale_y_continuous(expand = c(0, 0))
```


`map_gdp` layers the background `earth` polygon while the main `geom_sf(...)` choropleth layer with thin black borders and undertermined status grey layers and `geom_shadowtext(...)` to place outlined country-code labels at each of the 40 largest countries' centroids.

`scale_fill_fermenter(...)` places a horizontal, top-centered binned legend which doubles as the map title.

`scale_color_manual(...)` on the GDP map places the shared "Missing"/"Undetermined" legend keys at the bottom of the figure.

`theme_void()` strips the background grid lines and axis text to minimize non-data ink.

`coord_sf(...)` clips both the maps to populated latitude range while keeping the Equal Earth projection.

`scale_y_continuous(expand = c(0, 0))` helps to keep the crop exactly at 84°N and 56°S

```{r}
#| label: patchwork-plot
#| fig-width: 10
#| fig-height: 12

map_life_exp /
  map_gdp +
  plot_annotation(
    title = "Comparing wealth and longevity across nations (2024)",
    caption = "\"Undetermined\" refers to an area's legal source \nSource: World Bank",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.caption = element_text(hjust = 0.5),
    )
  )
```

`map_life_exp`/`map_gdp` stacks the two maps vertically and `plot_annotation(...)` adds a centered boldface title plus a centered caption crediting to the World Bank.

### 3.13 Verifier Task
1. **Layout:** Two choropleth maps are stacked vertically, with life expectancy on the top panel and GDP per capita on the bottom panel.
2. **Map projection:** Both maps use the same equal-area projection (EPSG:8857).
3. **Clipping:** Both maps are clipped at 56°S and 84°N to focus on populated regions.
4. **Color palettes:** Sequential ColorBrewer palettes are used (Greens for life expectancy and Reds for GDP per capita).
5. **Gray fills for missing values and undetermined areas:** Missing values are filled with light gray, while legally undetermined areas are filled with dark gray and shown in a shared legend.
6. **Country borders:** Thin black borders clearly separate neighboring countries.
7. **Legend scales:** Legends use appropriate breaks, labels, and units for each variable.
8. **Legend placement:** Each color legend is centered above its corresponding map and oriented horizontally. The gray-category legend appears once at the bottom.
9. **Country labels:** ISO country codes are displayed for the 40 largest countries using `shadowtext` with a white halo for readability.
10. **Title:** A bold, centered title indicates the variables being compared and the year (2024).
11. **Caption:** The caption defines "Undetermined" and credits the World Bank as the data source.
12. **Background and frame:** Oceans are filled with a light blue color, the page background is white, and the globe is framed on the left and right edges only.
12. **Theme:** A minimal theme (`theme_void()`) is used to reduce non-data ink.
13. **Figure dimensions:** The figure size keeps country labels and legends legible without appearing oversized.

## 3.14 Conclusion

Stacking the Life Expectancy and GDP per Capita maps allows for a direct visual comparison of the spatial correlation between wealth and longevity. In most regions, the metrics track closely together. countries such as Australia, for example, exhibit the highest-tier color bands in both categories.

This is also supported how a reader can instantly spot that Central and Western Africa form the core of the lowest bins for both metrics (GDP under $2,000 to $5,000 and life expectancy under 60 to 65 years).

However, this comparative layout also exposes anomalies. The United States serves as an example of this divergence. Despite ranking in the highest bracket for economic wealth, its life expectancy visibly lags behind its economic peers, falling into a lower-tier bracket.

An advantage of the choropleth map is how it retains the geographical context. The Preston Curve strips away geography entirely and countries become isolated dots. The choropleth maps show where these trends live and a reader can see regional clusters, revealing patterns that a scatter plot completely hides. Additionally, the world map can be easier to interpret for the general public compared to scatter plots.

Disadvantages of the Choropleth map is how smaller countries such as Singapore is too small to really be observed. On a Preston Curve, every nation gets exactly one equal dot, regardless of geographic size. Additionally, it may not be as simple for the user to interpret the correlation as they have to keep comparing up and down between the two maps but a user is able to interpret the correlation in one glance on the Preston curve. 

## AI Liaison Report

**Tool used:** Perplexity (Gemini 3.1 Pro)

### One Useful AI Response

**In the prompt :**

```r
How do I use st_cast()?

This was how I used to it to try to cast a MULTIPOLYGON to only "POLYGON"
"no_of_polygons <- st_cast(singapore, to, 'POLYGON')"

But I ran into this error,
Error:
! object 'to' not found
Backtrace:
    ▆
 1. ├─sf::st_cast(singapore, to, "POLYGON")
 2. └─sf:::st_cast.sf(singapore, to, "POLYGON")
 3.   ├─sf::st_cast(st_geometry(x), to, group_or_split = do_split)
 4.   └─sf:::st_cast.sfc(st_geometry(x), to, group_or_split = do_split)       
 5.     └─sf:::which_sfc_col(to)

Quitting from week-10-lab.qmd:139-150 [count-points-singapore]
Execution halted
WARN: Error encountered when rendering files
```

The AI was able to assist me in getting the understanding why the error happened, how to fix it and informed me in a way that is relevant to the code I had despite not giving the full entire context of what the labsheet is about or the specifics of the question.

### Suboptimal AI Response

**Prompt that seem to be incomplete/outdated:**

From the prompt:

```r
map_gdp <- map_gdp + 
    geom_sf(data = countries, fill = "lightblue1") + 
    theme(
        legend.position = "top"
    )

Why does this not work, and is it possible to make it work without re-writing the entire plot?
```

The AI actually tried to indicate that the team was using the wrong dataset as the number 1 reason which was totally wrong. Although it is a fair concern to bring up, the reason given is suboptimal. The team to had to send the following prompt to ask for other reason which was able to help us better.


### Reflection

The team tried to use another model (Gemini 3.1) from the usual (Claude Sonnet) which we typically had used to see if the performance of it would be better. We found out how it was not as good at keeping context however, the structure of informative reasoning is arguably slightly better than that of Claude's. 

Furthermore, from today's lab, the team felt the necessity to ensure that programmer should have understanding of context and goal to ensure that AI's output can be verified before being used. Working together to complete the lab resulted in better result to achieve intended results rather than solely relying on AI.