---
title: "Week 10 Lab: Choropleth Maps of Country-Level Statistics"
author: "Team Orchid"
date: "2026-07-07"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

## 3.1 Identify Suitable Fill Variables for Choropleth Maps
| Variable | Variable Type | Suitable as fill color? Why or why not?  | Would the result be a choropleth map? Why or why not?  |
| ------ | ---- | -----------------------| -----------------------|
| GDP per capita |   |   |   |
| Life expectancy |   |   |   |
| Region |   |   |   |
| Population |   |   |   |

## 3.2 Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(mapview)
library(patchwork)
library(readxl)
library(rmapshaper)
library(scales)
library(sf)
library(shadowtext)
```

## 3.3 Import Statistics
```{r}
#| label: import-data

gdp <-
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
  read_xls(skip = 3)

life_exp <-
  "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls" |>
  read_xls(skip = 3)

gdp
life_exp
```


```{r}
#| label: selecting-and-merging-data

gdp <- gdp |>
  select(name = "Country Name", code = "Country Code", gdp_per_cap = "2024")

life_exp <- life_exp |>
  select(name = "Country Name", code = "Country Code", life_exp = "2024")

country_stats <- gdp |>
  left_join(life_exp, join_by(name, code))

country_stats
```

```{r}
#| label: import-metadata

metadata <-
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls" |>
  read_xls(sheet = "Metadata - Countries") |>
  select(code = "Country Code", region = "Region")
```

```{r}
#| label: join-metadata

country_stats <- country_stats |>
  left_join(metadata, join_by(code))

country_stats
```

```{r}
#| label: handle-missing-data

country_stats <- country_stats |>
  drop_na(region) |>
  select(name, code, gdp_per_cap, life_exp)

country_stats
```

```{r}
#| label: verifier-task-3.3

```

## 3.4 Import Boundaries

```{r}
#| label: import-boundaries
#| cache: true

country_boundaries <-
  "WB_GAD_ADM0_complete.geojson" |>
  read_sf() |>
  select(code = ISO_A3, status = WB_STATUS)

country_boundaries
```

## 3.5 Validate and Repair Geometry

```{r}
#| label: counting-geometry

sum(!st_is_valid(country_boundaries))
```

There are 5 invalid geometries within country_boundaries

```{r}
#| label: preparing-geometry

country_boundaries <- st_make_valid(country_boundaries)
sum(!st_is_valid(country_boundaries))
```

`st_make_valid()` managed to repair the invalid geometries and this is confirmed by counting the number of invalid geometries with `st_is_valid()` again, which returns 0.

## 3.6 Simplify Polygons

### 3.6a

```{r}
#| label: count-points

npts(country_boundaries)
```

### 3.6b
```{r}
#| label: count-polygons-singapore

singapore <- country_boundaries |>
  filter(code == "SGP")

singapore

no_of_polygons <- st_cast(singapore, to = 'POLYGON')
no_of_polygons
```

There are a total of 20 polygons in Singapore

### 3.6c

```{r}
#| label: simplify-boundaries

country_boundaries <- ms_simplify(
  country_boundaries,
  keep_shapes = TRUE,
  keep = 0.0015
)

npts(country_boundaries)
country_boundaries
```

Using `keep_shapes = TRUE` allows us to retain at least one polygon for every country when using `ms_simplify`.

To retain approximately 10,000 boundary points, we reduce the argument `keep` from a default value of `0.05` to `0.0015` 


### 3.6d

```{r}
#| label: simplified-polygons-singapore

singapore <- country_boundaries |>
  filter(code == "SGP")

no_of_polygons <- st_cast(singapore, to = 'POLYGON')
no_of_polygons
```

After the simplification, Singapore is left with only one polygon.

## 3.7 Cure Antimeridian

```{r}
#| label: plot-sf

ggplot(country_boundaries) + geom_sf()
```

### 3.7b

```{r}
#| label: cure-antimeridian

country_boundaries <- st_wrap_dateline(country_boundaries)
country_boundaries <- country_boundaries |>
  st_cast(country_boundaries, to = 'MULTIPOLYGON')

ggplot(country_boundaries) + geom_sf()
```

```{r}
#| label: verifier-task-3.7

sum(!st_is_valid(country_boundaries))
```

## 3.8 Merge features

### 3.8a
```{r}
#| label: find-shared-code

count_code <- count(country_boundaries, code) |>
  filter(n > 1)

count_code
```

### 3.8b

The code FRA is shared by France and Clipperton Island, an uninhabited coral atoll in the eastern Pacific Ocean, roughly 1,300 miles (2,090 km) southwest of Mexico. It carries France's country code because Clipperton is a French dependency administered directly by the Minister of Overseas France, rather than a separate territory with its own ISO code. Background information: [https://www.britannica.com/place/Clipperton-Island](https://www.britannica.com/place/Clipperton-Island)

### 3.8c

```{r}
#| label: merge-features

duplicate_countries <- count_code |>
  select(code) |>
  filter(code != "NA") |>
  st_drop_geometry()

dup_codes <- duplicate_countries$code

duplicates_only <- country_boundaries |>
  filter(code %in% dup_codes)

non_duplicates <- country_boundaries |>
  filter(!code %in% dup_codes)

merged_features <- duplicates_only |>
  group_by(code) |>
  summarize(
    status = ifelse(
      any(status == "Member State"),
      "Contains a Member State",
      "Does Not Contain a Member State"
    )
  ) |>
  st_cast("MULTIPOLYGON")

country_boundaries <- bind_rows(non_duplicates, merged_features)

country_boundaries
```


```{r}
#| label: verifier-task-3.8

country_boundaries |> st_is("MULTIPOLYGON") |> all()
```

## 3.9 Join Statistics

```{r}
#| label: join-statistics

countries <- country_boundaries |>
  left_join(country_stats, join_by(code))

countries
```

## 3.10 Compare Projections

### 3.10a

```{r}
#| label: projection-name

st_crs(country_boundaries, parameters = TRUE)[c("Name", "srid")]
```

The name is "WGS 84", which stands for World Geodetic System 84. This is the default projection which maps the longitude to the x-axis and latitude to the y-axis. However, a trait of this projection is that it is a "equirectangular" which do not plot the areas in the correct proportion.

### 3.10b

```{r}
#| label: plot-equal-earth

ggplot(country_boundaries) +
  geom_sf() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45)) +
  scale_y_continuous(
    breaks = c(-89.9, -45, 0, 45, 89.9)
  ) +
  coord_sf(crs = "EPSG:8857")
```

### 3.10c
```{r}
#| label: plot-other-projection

ggplot(country_boundaries) +
  geom_sf() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45)) +
  scale_y_continuous(
    breaks = c(-89.9, -45, 0, 45, 89.9)
  ) +
  coord_sf(crs = "ESRI:54009")
```

The Equirectangular projection while straightforward, it significantly distorts areas and shapes. It drastically stretches areas as you move away from the equator toward the poles. As an example, Greenland appears to be the same size as Australia but is only 28% of Australia's land. However the graticule interpretability is high as The grid lines form straight, evenly spaced horizontal and vertical lines, making it the easiest projection for reading raw latitude and longitude coordinates.

The Mollweide projection is an equal-area projection that depicts the pole as single points and the antimeridians merge into each other without any sharp corners at the poles. The area distortion is zero as it preserves relative land areas across the entire globe, correctly depicting e.g. the area of Greenland to Australia. This projection undergoes some shape distortion as continents near the boundary appear slanted or stretched. Lastly, the graticule interpretability isn't as high as the latitude are straight horizontal lines but the longitude curve significantly outwards towards the edge.

The Equal Earth projection like the MollWeide projection preserves the relative area well, leading it to have zero distortion in terms of Area. The shape distortion is relatively low as the map has a flat top and bottom, rather than a cylindrical shape like the Mollweide projection. The graticule interpretability is high as the latitude are straight and equally spaced, and the curvature of the meridians are predictable and balanced.

Overall, the Equal Earth projection is the most suitable for a global choropleth map as while the Mollweide projection also achieves zero distortion in terms of area (Equirectangular completely fails in this regard), Mollweide's shape distortion is relatively worse to the Equal Earth projection, and the graticule interpretability of the Equal Earth projection is better than the Mollweide projection. 

## 3.11 Choropleth Draft

```{r}
#| label: basic-choropleth-map

choropleth_map <- ggplot(countries) +
  geom_sf(aes(fill = gdp_per_cap)) +
  coord_sf(crs = "EPSG:8857")

choropleth_map
```

While there are gray patches to indicate missing values, it is unclear whether it is because of missing gdp_per_cap values or undetermined status areas. The default continuous scale is also unsuitable as due to the outlier values, the midpoint is around ~60,000. This causes a lot of countries to be in the dark blue part of the scale and most of the plot looks like the same dark blue shade. Additionally, a darker color is usually used to represent a higher value, which is not represented by default.

## 3.12 Fixing Flaws

### 3.12a

```{r}
#| label: fix-scale-and-colour

choropleth_map <- choropleth_map +
  scale_fill_fermenter(
    palette = "Reds",
    direction = -1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale())
  )

choropleth_map
```

### 3.12b

```{r}
#| label: fill-missing-countries

ggplot(countries) +
  geom_sf(aes(fill = gdp_per_cap)) +
  scale_fill_fermenter(
    palette = "Reds",
    direction = 1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  geom_sf(
    data = filter(countries, is.na(gdp_per_cap)),
    aes(color = "Missing"),
    fill = "lightgray"
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    aes(color = "Undetermined"),
    fill = "darkgray"
  ) +
  scale_color_manual(
    values = c("Missing" = "black", "Undetermined" = "black")
  ) +
  guides(
    color = guide_legend(override.aes = list(fill = c("lightgray", "darkgray")))
  ) +
  coord_sf(crs = "EPSG:8857")
```

## 3.13 Two-Panel Choropleth Map
```{r}
#| label: two-panel-choropleth-prep

earth <- st_polygon(
  x = list(
    cbind(c(rep(-180, 181), rep(180, 181), -180), c(-90:90, 90:-90, -90))
  )
) |>
  st_sfc() |>
  st_set_crs(4326) |>
  st_as_sf()

top_40_largest <- countries |>
  mutate(area = as.numeric(st_area(geometry))) |>
  slice_max(order_by = area, n = 40)
```

```{r}
#| label: life-exp-plot

map_life_exp <- ggplot(countries) +
  geom_sf(data = earth, fill = "aliceblue") +
  geom_sf(
    aes(fill = life_exp),
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_fermenter(
    name = "Life expectancy at birth \n(years)",
    palette = "Greens",
    direction = 1,
    breaks = c(60, 65, 70, 75, 80),
    guide = guide_colorsteps(
      position = "top",
      barwidth = unit(15, "lines"),
      barheight = unit(0.8, "lines"),
      title.hjust = 0.5
    )
  ) +
  geom_sf(
    data = filter(countries, is.na(gdp_per_cap)),
    fill = "lightgray",
    linewidth = 0.1
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    fill = "darkgray",
    linewidth = 0.1
  ) +
  scale_color_manual(
    values = c("Missing" = "black", "Undetermined" = "black"),
  ) +
  geom_shadowtext(
    data = top_40_largest,
    aes(label = code, geometry = geometry),
    stat = "sf_coordinates",
    bg.color = "white",
    color = "black",
    check_overlap = TRUE
  ) +
  theme_void() +
  theme(
    legend.margin = margin(5, 0, 0, 0),
    legend.frame = element_rect(color = "black"),
    legend.position = "top"
  ) +
  coord_sf(
    crs = "EPSG:8857",
    ylim = c(-56, 84),
    xlim = c(-179.99, 179.99),
    default_crs = 4326,
  )
```

```{r}
#| label: gdp-plot

map_gdp <- ggplot(countries) +
  geom_sf(data = earth, fill = "aliceblue") +
  geom_sf(
    aes(fill = gdp_per_cap),
    color = "black",
    linewidth = 0.1
  ) +
  scale_fill_fermenter(
    name = "GDP per capita \n($ Purchasing Power Parity 2021)",
    palette = "Reds",
    direction = 1,
    breaks = c(2000, 5000, 10000, 20000, 50000, 100000),
    labels = label_number(scale_cut = cut_short_scale()),
    guide = guide_colorsteps(
      position = "top",
      barwidth = unit(15, "lines"),
      barheight = unit(0.8, "lines"),
      title.hjust = 0.5
    )
  ) +
  geom_sf(
    data = filter(countries, is.na(gdp_per_cap)),
    aes(color = "Missing"),
    fill = "lightgray",
    linewidth = 0.1
  ) +
  geom_sf(
    data = filter(countries, status == "Non-determined legal status area"),
    aes(color = "Undetermined"),
    fill = "darkgray",
    linewidth = 0.1
  ) +
  scale_color_manual(
    name = "",
    values = c("Missing" = "black", "Undetermined" = "black"),
    guide = guide_legend(
      position = "bottom",
      override.aes = list(
        fill = c("Missing" = "lightgray", "Undetermined" = "darkgray"),
        color = "black",
        linewidth = 0.1
      )
    )
  ) +
  geom_shadowtext(
    data = top_40_largest,
    aes(label = code, geometry = geometry),
    stat = "sf_coordinates",
    bg.color = "white",
    color = "black",
    check_overlap = TRUE
  ) +
  theme_void() +
  theme(
    legend.margin = margin(5, 0, 0, 0),
    legend.frame = element_rect(color = "black"),
  ) +
  coord_sf(
    crs = "EPSG:8857",
    ylim = c(-56, 84),
    xlim = c(-179.99, 179.99),
    default_crs = 4326,
  )
```

```{r}
#| label: patchwork-plot
#| fig-width: 10
#| fig-height: 12

map_life_exp /
  map_gdp +
  plot_annotation(
    title = "Comparing wealth and longevity across nations (2024)",
    caption = "\"Undetermined\" refers to an area's legal source \nSource: World Bank",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.caption = element_text(hjust = 0.5),
    )
  )
```


## 3.14 Conclusion

Stacking the Life Expectancy and GDP per Capita maps allows for a direct visual comparison of the spatial correlation between wealth and longevity. In most regions, the metrics track closely together. countries such as Australia, for example, exhibit the highest-tier color bands in both categories.

This is also supported how a reader can instantly spot that Central and Western Africa form the core of the lowest bins for both metrics (GDP under $2,000 to $5,000 and life expectancy under 60 to 65 years).

However, this comparative layout also exposes anomalies. The United States serves as an example of this divergence. Despite ranking in the highest bracket for economic wealth, its life expectancy visibly lags behind its economic peers, falling into a lower-tier bracket.

An advantage of the choropleth map is how it retains the geographical context. The Preston Curve strips away geography entirely and countries become isolated dots. The choropleth maps show where these trends live and a reader can see regional clusters, revealing patterns that a scatter plot completely hides. Additionally, the world map can be easier to interpret for the general public compared to scatter plots.

Disadvantages of the Choropleth map is how smaller countries such as Singapore is too small to really be observed. On a Preston Curve, every nation gets exactly one equal dot, regardless of geographic size. Additionally, it may not be as simple for the user to interpret the correlation as they have to keep comparing up and down between the two maps but a user is able to interpret the correlation in one glance on the Preston curve. 