---
title: "Tidyverse-Lab"
author: "Team Orchid"
date: "2026-05-19"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(waldo)
```

## 3.1 Member's Analysis of Dataset


**a. What should be the columns and their classes (character, numeric, integer, etc.) in the tidy data frame?
**

| Member | country | class |
|---|---|---|

Quan Yi:


Animal Type: character

Brunei Darussalam: numeric

Cambodia: numeric

Indonesia: numeric

Lao PDR: numeric

Malaysia: numeric

Myanmar: numeric

Philippines: numeric

Singapore: numeric

Thailand: numeric

Viet Nam: numeric

ASEAN: numeric

b. How many rows should the tidy data frame have? Why?

It should have 16 rows as some rows are labels such as 'Livestock production' and year and should be omitted 

c. Why is the current layout of sheet X.14 not tidy? List at least three specific reasons, identifying rows or columns that violate tidy-data principles.

Tidy data requires 

1. Each variable forms a column. [This is violated as the meat should be the column and the country should be the rows]
2. Each observation forms a row. [This is violated as livestock production is a row but there are no values]
3. Each type of observational unit forms a table. [This is violated as each table should be its own year, e.g. we have 4 tables, and each year has its own livestock and poultry production table]

## 3.2 

```{r}
#| label: loading-data

meat_raw <- read_xlsx("ASEAN-Statistical-Yearbook-2023.xlsx", sheet="X.14", range="A4:L36")
meat_raw
```

## Task 3.3 Tidying the data

```{r}
#| label: tidying-meat
meat_tidy <- meat_raw |>
  mutate(year = recode_values(
    `Animal Type`,
    "2021" ~ 2021,
    "2020" ~ 2020,
    "2019" ~ 2019,
    "2018" ~ 2018
  ))|>
  fill(year, .direction = "down") |>
  mutate(`Animal Type` = str_to_lower(`Animal Type`)) |>
  filter(str_detect(`Animal Type`, "meat")) |>
  mutate(
    `Animal Type` = recode_values(
      `Animal Type`,
      "beef and buffalo meat" ~ "beef and buffalo",
      "mutton and goat meat"  ~ "mutton and goat",
      "pig meat"              ~ "pig",
      "poultry meat"          ~ "poultry"
  )) |>
  drop_na(`Animal Type`) |>
  pivot_longer(
    cols = -c(`Animal Type`, year),
    names_to = "country",
    values_to = "production"
  ) |>
  filter(country != "ASEAN") |>
  select(country, animal = `Animal Type`,year, production) |>
  arrange(desc(year), animal, country)
  
```

```{r}
#| label: tidy-check

head(meat_tidy)
tail(meat_tidy)

```


```{r}
#| label: verifier-3.3
```

## Task 3.4 Correcting Cambodia's Mutton-and-goat Entries

```{r}
#| label: correcting-cambodia-values

meat_tidy <- mutate(
  meat_tidy,
  production = if_else(
    country == "Cambodia" & animal == "mutton and goat" & production == 0,
    NA_real_,
    production
  )
)

```

```{r}
#| label: check-cambodia-values

check_rows <- meat_tidy |>
  filter(country == "Cambodia" & animal == "mutton and goat")

check_rows
```

We use `NA_real_` as the column contains decimal numbers/is numeric and standard NA can cause a mismatch error. `NA_real_` tells R to insert a missing value that is safe for numeric columns.


```{r}
#| label: verifier-3.4


```


## 3.5 Checkpoint: Compare meat_tidy to Reference Data

```{r}
#| label: checkpoint-1
#| message: false

meat_ref <- read_csv("meat_tidy_reference.csv")
compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

## 3.6 Tidy of Population Data

```{r}
#| label: tidy-population
pop_raw <- read_xlsx("ASEAN-Statistical-Yearbook-2023.xlsx", sheet="I.1", range="A5:K15")

pop_tidy <- pop_raw |>
  pivot_longer(
    cols = -Country,
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(year = recode_values(
    year,
    "2022" ~ 2022,
    "2021" ~ 2021,
    "2020" ~ 2020,
    "2019" ~ 2019,
    "2018" ~ 2018,
    "2017" ~ 2017,
    "2016" ~ 2016,
    "2015" ~ 2015,
    "2014" ~ 2014,
    "2013" ~ 2013
  )) |>
  select(country = "Country", year, pop = "population")|>
  arrange(desc(year), country)

pop_tidy
```

```{r}
#| label: verifier-3.6

```

## 3.7 Checkpoint: Compare pop_tidy to Reference Data

```{r}
#| label: checkpoint-2
#| message: false

pop_ref <- read_csv("population_tidy_reference.csv")
compare(pop_tidy, pop_ref, tolerance = 1e-6)
```

## 3.8 Join Meat Production to Population

```{r}
#| label: join-meat-and-pop

merged <- left_join(meat_tidy, pop_tidy, join_by(country, year))
merged
```

```{r}
#| label: verifier-3.8

meat_tidy
```

## 3.9 Derive Per-Capita Production

```{r}
#| label: add-meat-per-cap

merged <- mutate(
  merged,
  meat_per_cap = production/pop * (10^3)
)

merged

```

```{r}
#| label: check-merged-cambodia-values

check_rows2 <- merged |>
  filter(country == "Cambodia" & animal == "mutton and goat")

check_rows2
```


```{r}
#| label: verifier-3.9


```

## 3.10 Checkpoint: Compare meat_per_cap to Reference Data

```{r}
#| label: checkpoint-3
#| message: false

merged_ref <- read_csv("merged_reference.csv")
compare(merged, merged_ref, tolerance = 1e-6)
```

## 3.11 Visualization of Result

```{r}
#| label: visualize-result

ggplot(
    merged,
    aes(x = year, y = meat_per_cap, color = animal)
  ) +
  geom_line() +
  geom_point() +
  labs(
    x = NULL,
    y = "Production (kg per person)",
    color = "Animal",
    title = "Per-Capita Meat Production in ASEAN, 2018–2021",
    caption = "Source: ASEAN Statistical Yearbook 2023"
  ) +
  xlim(2017.9, 2021.1) +
  scale_y_log10(
  breaks = c(0.01, 0.1, 1, 10, 100),
  labels = c("0.01", "0.1", "1", "10", "100")
  ) +
  facet_wrap(~country) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification.inside = c(1, 0.05)
)
```

## 3.12 Comments on Plot

It was observed that from 2020 to 2021, there was a sharp decline for the production of mutton and goat in Myanmar. 

This pattern is supported by a report from Myanmar Digital News on October 6, 2020, which noted that neighboring countries like China and Thailand completely suspended live goat purchases amid the COVID-19 crisis. This trade suspension left local breeders with no export market for months, causing prices to plummet. The news also mentioned that the other meat markets were not doing well which explains the decline across the other meat production in the dataset during this period.

*Reference: https://www.mdn.gov.mm/en/domestic-goat-market-experiencing-downward-trend-due-covid-19-crisis*



