---
title: "Tidyverse-Lab"
author: "Team Orchid"
date: "2026-05-19"
format:
  html:
    toc: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(waldo)
```

## 3.1 Member's Analysis of Dataset


**a. What should be the columns and their classes (character, numeric, integer, etc.) in the tidy data frame?
**

| Member | country | class |
|---|---|---|

Quan Yi:


Animal Type: character

Brunei Darussalam: numeric

Cambodia: numeric

Indonesia: numeric

Lao PDR: numeric

Malaysia: numeric

Myanmar: numeric

Philippines: numeric

Singapore: numeric

Thailand: numeric

Viet Nam: numeric

ASEAN: numeric

b. How many rows should the tidy data frame have? Why?

It should have 16 rows as some rows are labels such as 'Livestock production' and year and should be omitted 

c. Why is the current layout of sheet X.14 not tidy? List at least three specific reasons, identifying rows or columns that violate tidy-data principles.

Tidy data requires 

1. Each variable forms a column. [This is violated as the meat should be the column and the country should be the rows]
2. Each observation forms a row. [This is violated as livestock production is a row but there are no values]
3. Each type of observational unit forms a table. [This is violated as each table should be its own year, e.g. we have 4 tables, and each year has its own livestock and poultry production table]

## 3.2 

```{r}
#| label: loading-data

meat_raw <- read_xlsx("ASEAN-Statistical-Yearbook-2023.xlsx", sheet="X.14", range="A4:L36")
meat_raw
```

## Task 3.3 Tidying the data

```{r}
#| label: tidying-meat
meat_tidy <- meat_raw |>
  mutate(year = recode_values(
    `Animal Type`,
    "2021" ~ 2021,
    "2020" ~ 2020,
    "2019" ~ 2019,
    "2018" ~ 2018
  ))|>
  fill(year, .direction = "down") |>
  mutate(`Animal Type` = str_to_lower(`Animal Type`)) |>
  filter(str_detect(`Animal Type`, "meat")) |>
  mutate(
    `Animal Type` = recode_values(
      `Animal Type`,
      "beef and buffalo meat" ~ "beef and buffalo",
      "mutton and goat meat"  ~ "mutton and goat",
      "pig meat"              ~ "pig",
      "poultry meat"          ~ "poultry"
  )) |>
  drop_na(`Animal Type`) |>
  pivot_longer(
    cols = -c(`Animal Type`, year),
    names_to = "country",
    values_to = "production"
  ) |>
  filter(country != "ASEAN") |>
  select(country, animal = `Animal Type`,year, production) |>
  arrange(desc(year), animal, country)
  
```

```{r}
#| label: tidy-check

head(meat_tidy)
tail(meat_tidy)

```


```{r}
#| label: verifier-3.3
```

## Task 3.4 Correcting Cambodia's Mutton-and-goat Entries


