---
title: "Tidyverse-Lab"
author: "Team Orchid"
date: "2026-05-19"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | 
|---|---|
| Quan Yi | Coder |
| Shi Min | Navigator |
| Aiman | AI Liaison |
| Subhashini | Verifier |
| Jabar | Reporter |
| Min Xuan | Constructive Critic |

## Setup
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(waldo)
```

## 3.1 Member's Analysis of Dataset

In this section, we first examine the structure of sheet X.14 dataset before running deeper analysis.  
Each member makes prediction about the column classes, distinct number of rows and list at least three specific reasons with regards to the current layout not being tidy.

Each member predicts:

1. The class of each column (`Animal Type`, `Brunei Darussalam`, `Cambodia`, `Indonesia`, `Lao PDR`, `Malaysia`, `Philippines`, `Singapore`, `Thailand`, `Viet nam`, `ASEAN`)
2. The number of distinct row values
3. List at least three specific reasons identifying rows or columns that violate tidy-data principles

**Class Prediction**

| Member | Character Column | Numeric Class |
|---|---|---|
| Quan Yi | Animal Type | Brunei Darussalam, Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN |
| Shi Min | Animal Type | Brunei Darussalam, Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN |
| Aiman | Animal Type, Brunei Darussalam, Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN | Production, Year | 
| Subhashini | Brunei Darussalam | Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN |
| Jabar | Animal Type | Brunei Darussalam, Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN |
| Min Xuan | Animal Type | Brunei Darussalam, Cambodia, Indonesia, Lao PDR, Malaysia, Phillippines, Singapore, Thailand, Viet nam, ASEAN |

**Number of Distinct Row Values**

| Member | Row | Reason |
|---|---|---|
| Quan Yi | 16 | It should have 16 rows as some rows are labels such as 'Livestock production' and year and should be omitted |
| Shi Min | 33 | Following the excel layout, there should be roughly 33 rows without white spaces |
| Aiman | 176 | There should be about 176 rows (countries x animal types x years = 11 x 4 x 4) |
| Subhashini |  33 | 33 rows when imported without manipulation |
| Jabar | 23 | Based on the excel format, the number of rows should be approximately 23 without the visual separation and missing values |
| Min Xuan | 33 | When imported into a tiny data frame without manipulation, it should have 33 rows |

**List 3 Reasons**

| Member | Reason 1 | Reason 2 | Reason 3 |
|---|---|---|---|
| Quan Yi | Country names are stored as columns instead of one `country` column. | Some rows, such as “Livestock production”, are labels rather than observations. | Year values are used as section headers instead of being stored in a `year` column. |
| Shi Min | The table uses repeated year sections. | Category header rows are mixed with actual data rows. | Country names are spread across multiple columns instead of one variable column. |
| Aiman | Column headers are values, not variable names. | Year is stored as a grouped row label instead of a variable. | Production values are spread across multiple rows and columns. |
| Subhashini | Year of production lacks its own column. | Column headers are values, not variable names. Country names are spread across columns instead of being values under a single country column. | There are empty separator rows, repeated group headers under "Animal Type". |
| Jabar | Empty visual separator rows create rows with missing values. | Year should be a variable instead of appearing as an observation/header. | Aggregate rows, such as ASEAN totals, are not single country observations. |
| Min Xuan | Year of production should be in its own column. | Country names are stored as column headers instead of under `country`. | Empty separator rows and aggregate rows make the data untidy. |

## 3.2 

```{r}
#| label: loading-data
meat_raw <- read_xlsx(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "X.14",
  range = "A4:L36"
)
meat_raw
```

`read_xlsx()` loads data from an Excel file specifically sheet "X.14" and grabbing a specific range.

## Task 3.3 Tidying the data

```{r}
#| label: tidying-meat
meat_tidy <- meat_raw |>
  mutate(
    year = recode_values(
      `Animal Type`,
      "2021" ~ 2021,
      "2020" ~ 2020,
      "2019" ~ 2019,
      "2018" ~ 2018
    )
  ) |>
  fill(year, .direction = "down") |>
  mutate(`Animal Type` = str_to_lower(`Animal Type`)) |>
  filter(str_detect(`Animal Type`, "meat")) |>
  mutate(
    `Animal Type` = recode_values(
      `Animal Type`,
      "beef and buffalo meat" ~ "beef and buffalo",
      "mutton and goat meat" ~ "mutton and goat",
      "pig meat" ~ "pig",
      "poultry meat" ~ "poultry"
    )
  ) |>
  drop_na(`Animal Type`) |>
  pivot_longer(
    cols = -c(`Animal Type`, year),
    names_to = "country",
    values_to = "production"
  ) |>
  filter(country != "ASEAN") |>
  select(country, animal = `Animal Type`, year, production) |>
  arrange(desc(year), animal, country)
```

`meat_tidy` is assigned from raw data of `meat_raw` which goes thru an operator `|>` that take the outcome and passes it as first argument to the next function `mutate()`.
`fill()` fills missing values by copying the above value of year column downward.
`mutate(`` = str_lower(``))` converts all text into lowercase.
`filter(str_detect(``, ""))` keeps rows that contains the word meat from "Animal Type".
`mutate(`` = recode_values())` renames the end names "pig meat" to "pig" from specific animal type labels.
`drop_na(``)` removes NA rows under "Animal Type".
`pivot_longer()` reshapes the data from wide to long format.
`filter(!= "")` removes aggregate "ASEAN" row keeping only individual countries.
`select()` picks and reorders the four intended columns.
`arrange()` sorts the table by most recent year then animal then country.

```{r}
#| label: tidy-check
head(meat_tidy)
tail(meat_tidy)
```

`head()` shows the first six rows.
`tail()` shows the last six rows.

```{r}
#| label: verifier-3.3
```

## Task 3.4 Correcting Cambodia's Mutton-and-goat Entries

```{r}
#| label: correcting-cambodia-values
meat_tidy <- mutate(
  meat_tidy,
  production = if_else(
    country == "Cambodia" & animal == "mutton and goat" & production == 0,
    NA_real_,
    production
  )
)
```

`mutate(, = if_else())` updates the "production" column with "NA_real_" if the condition equals to 0. 


```{r}
#| label: check-cambodia-values
check_rows <- meat_tidy |>
  filter(country == "Cambodia" & animal == "mutton and goat")

check_rows
```

We use `NA_real_` as the column contains decimal numbers/is numeric and standard NA can cause a mismatch error. `NA_real_` tells R to insert a missing value that is safe for numeric columns.


```{r}
#| label: verifier-3.4


```


## 3.5 Checkpoint: Compare meat_tidy to Reference Data

```{r}
#| label: checkpoint-1
#| message: false
meat_ref <- read_csv("meat_tidy_reference.csv")
compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

`read_csv()` loads an existing data.
`compare()` checks whether the tidied data corresponds to reference exactly, allowing tiny floating point.

## 3.6 Tidy of Population Data

```{r}
#| label: tidy-population
pop_raw <- read_xlsx(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  range = "A5:K15"
)

pop_tidy <- pop_raw |>
  pivot_longer(
    cols = -Country,
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = recode_values(
      year,
      "2022" ~ 2022,
      "2021" ~ 2021,
      "2020" ~ 2020,
      "2019" ~ 2019,
      "2018" ~ 2018,
      "2017" ~ 2017,
      "2016" ~ 2016,
      "2015" ~ 2015,
      "2014" ~ 2014,
      "2013" ~ 2013
    )
  ) |>
  select(country = "Country", year, pop = "population") |>
  arrange(desc(year), country)

pop_tidy
```

`read_csv()` loads data from an Excel file specifically sheet "I.1" and grabbing a specific range.
`pivot_longer()` reshapes from wide to long
`mutate(`` = recode_values())` converts the year values from from character strings to integers.
`select()` keeps the three intended columns and renames them.
`arrange()` sorts the table by most recent year then alphabetical by country.

```{r}
#| label: verifier-3.6

```

## 3.7 Checkpoint: Compare pop_tidy to Reference Data

```{r}
#| label: checkpoint-2
#| message: false
pop_ref <- read_csv("population_tidy_reference.csv")
compare(pop_tidy, pop_ref, tolerance = 1e-6)
```

`compare()` compares against the tidied data.

## 3.8 Join Meat Production to Population

```{r}
#| label: join-meat-and-pop
merged <- left_join(meat_tidy, pop_tidy, join_by(country, year))
merged
```

`left_join()` merges both by matching country and year. 

```{r}
#| label: verifier-3.8
meat_tidy
```

## 3.9 Derive Per-Capita Production

```{r}
#| label: add-meat-per-cap
merged <- mutate(
  merged,
  meat_per_cap = production / pop * (10^3)
)

merged
```

`mutate(, = / * (10^3))` calculates the meat in kilograms produced per person.

```{r}
#| label: check-merged-cambodia-values
check_rows2 <- merged |>
  filter(country == "Cambodia" & animal == "mutton and goat")

check_rows2
```

`filter()` spot-checks if Cambodia's mutton and goat carried NA.

```{r}
#| label: verifier-3.9


```

## 3.10 Checkpoint: Compare meat_per_cap to Reference Data

```{r}
#| label: checkpoint-3
#| message: false
merged_ref <- read_csv("merged_reference.csv")
compare(merged, merged_ref, tolerance = 1e-6)
```

`read_csv()` loads an existing data.
`compare()` is to check whether the tidied data corresponds to reference exactly, allowing tiny floating point.

## 3.11 Visualization of Result

```{r}
#| label: visualize-result
ggplot(
  merged,
  aes(x = year, y = meat_per_cap, color = animal)
) +
  geom_line() +
  geom_point() +
  labs(
    x = NULL,
    y = "Production (kg per person)",
    color = "Animal",
    title = "Per-Capita Meat Production in ASEAN, 2018–2021",
    caption = "Source: ASEAN Statistical Yearbook 2023"
  ) +
  xlim(2017.9, 2021.1) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("0.01", "0.1", "1", "10", "100")
  ) +
  facet_wrap(~country) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(1, 0),
    legend.justification.inside = c(1, 0.05)
  )
```

`ggplot(, aes()) + geom_line() + geom_point() + labs()..` sets up the scatter plot.
`xlim()` adds tolerance to the data by extending the x-axis limits.
`scale_y_log10()` transforms y-axis into a log scale.
`facet_wrap()` splits the chart one panel per country.
`theme()` just inputs the legend beside the plot area.

## 3.12 Comments on Plot

It was observed that from 2020 to 2021, there was a sharp decline for the production of mutton and goat in Myanmar. 

This pattern is supported by a report from Myanmar Digital News on October 6, 2020, which noted that neighboring countries like China and Thailand completely suspended live goat purchases amid the COVID-19 crisis. This trade suspension left local breeders with no export market for months, causing prices to plummet. The news also mentioned that the other meat markets were not doing well which explains the decline across the other meat production in the dataset during this period.

*Reference: https://www.mdn.gov.mm/en/domestic-goat-market-experiencing-downward-trend-due-covid-19-crisis*



