---
title: "Week 12 Lab: Raster Maps of Sea Surface Temperature Anomalies"
author: "Team Orchid"
date: "2026-07-21"
format:
  html:
    toc: true
execute:
  cache: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Specific Contribution  |
| ------ | ---- | -----------------------|
| Quan Yi | Coder | Developed the main QMD implementation by coding the data processing steps, creating the required visualisations, and ensuring the code chunks rendered correctly. Also checked that the final QMD could be executed from start to finish without rendering errors. |
| Shi Min | Navigator  | Read and interpreted the assignment requirements during the lab, kept the group aligned with the required tasks, and ensured that each section of the QMD addressed the relevant questions. Also helped maintain the flow, readability, and organisation of the QMD document. |
| Aiman | Reporter | Prepared the written Markdown explanations for the code and visualisations, including descriptions of what each code section does and how the resulting plots should be interpreted. Helped ensure that the report explanations were clear and connected to the assignment objectives. |
| Subhashini | Verifier  | Validated the accuracy of the generated plots, calculations, and interpretations by cross-checking outputs against the dataset and task requirements. Identified possible inconsistencies and ensured that the final answers were supported by the visual evidence.  |
| Jabar  | AI Liaison | Managed the team’s AI-assisted workflow by prompting AI tools together with the coder, documenting relevant AI interactions, and reviewing AI-generated suggestions before they were accepted, modified, or rejected. Ensured that AI use supported the assignment without replacing team judgement. |
| Min Xuan | Constructive Critic | Reviewed the combined work after merging individual contributions, suggested improvements to wording, structure, and visual clarity, and checked that the final submission package was complete. Also helped ensure that the final ZIP contained the required files and maintained submission integrity. |

## 3.1 Plan the Map Encodings

Q1: Why should `anom`, rather than `sst`, determine the principal fill colors?

A1: There would be a distinct gradient color difference between anom and not so much of sst. In addition, sst would just show tropics are warm and the poles are cold which fails to remove the baseline geography which is intended to show how much conditions differ from normal.

Q2: Should the anomaly scale be sequential or diverging? What does its central value represent?

A2: It should be diverging. The central value represents the observed sst that precisely tallies with its climate average.

Q3: Why should both panels use identical anomaly-scale limits?

A3: Using identical anomaly-scale gives us the leverage to compare colours across the two Decembers for instance when a patch of same colour gets encoded in the same magnitude, it is easy to distinguish the differences of intensity.


## 3.2 Start Coding
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(stars)
library(scales)
library(ggnewscale)
library(units)
```

Utlized new R packages `stars` for handling raster data and `ggnewscale` for plotting a workaround of stacking two different fill encodings.

## 3.3 Import the Daily Grids as `stars` Data Cubes
```{r}
#| label: import-stars
files_2009 <- list.files("C:/Users/mjaba/Documents/GITHUB/CSC3107-InformationVisualisation-Orchid/team-orchid-raster-lab/oisst_daily/el_nino_2009-12", pattern = "\\.nc$", full.names = TRUE)
files_2010 <- list.files("C:/Users/mjaba/Documents/GITHUB/CSC3107-InformationVisualisation-Orchid/team-orchid-raster-lab/oisst_daily/la_nina_2010-12", pattern = "\\.nc$", full.names = TRUE)

dec_2009 <- read_stars(files_2009, along = "time")
dec_2010 <- read_stars(files_2010, along = "time")

dec_2009

```

`list.files(...)` list all .nc files for each December, sorted chronologically for both `files_2009` and `files_2010`.

`read_stars()` reads multiple files and stacks them along a new dimension labelled `time`.

### 3.3 Verifier Task
```{r}
#| label: verify-3-3
#| eval: true

# Attributes
names(dec_2009)
names(dec_2010)

# Dimensions
st_dimensions(dec_2009)
st_dimensions(dec_2010)
st_get_dimension_values(dec_2009, "time")
st_get_dimension_values(dec_2010, "time")
```

`names()` confirms both objects consists of exactly two attributes: `anom` and `ice`.

`st_dimensions()` verifies that each object has dimensions.

`st_get_dimension_values(..., "time")` returns 31 chronological dates from December the first to December thirtieth of the respective year.

## 3.4 Establish the Coordinate Reference System
```{r}
#| label: establish-crs

dec_2009 <- dec_2009 |>
    st_set_crs(4326)

dec_2010 <- dec_2010 |>
    st_set_crs(4326)

```

`st_set_crs(4326)` assigns WGS 84 geographic CRS (EPSG:4326) so R interprets x and y as longitude and latitude. In addition, this does not alter the grid values or bounding box.

### 3.4 Verifier Task
```{r}
#| label: verify-3-4
#| eval: true

st_crs(dec_2009) == st_crs(4326)
st_crs(dec_2010) == st_crs(4326)
st_bbox(dec_2009)
st_bbox(dec_2010)

```

Both `st_crs()` comparisons should return `TRUE`  deducing that WGS 84 was assigned.

`st_bbox()` establishes that the grid geometry is unchanged.

## 3.5 Calculate Monthly Cell Summaries
```{r}
#| label: calculate-monthly-summaries

mean_anom_2009 <- st_apply(dec_2009["anom"], c("x", "y"), mean, na.rm = TRUE)
mean_ice_2009 <- st_apply(dec_2009["ice"], c("x", "y"), function(x) {
  mean(ifelse(is.na(x), 0, x))
})

mean_anom_2010 <- st_apply(dec_2010["anom"], c("x", "y"), mean, na.rm = TRUE)
mean_ice_2010 <- st_apply(dec_2010["ice"], c("x", "y"), function(x) {
  mean(ifelse(is.na(x), 0, x))
})

summary_2009 <- c(mean_anom = mean_anom_2009, mean_ice = mean_ice_2009) |>
  mutate(
    land = is.nan(mean_anom) & mean_ice == 0,
    mean_ice = if_else(land, NA_real_, mean_ice)
  ) |>
  select(land, mean_anom, mean_ice)

summary_2010 <- c(mean_anom = mean_anom_2010, mean_ice = mean_ice_2010) |>
  mutate(
    land = is.nan(mean_anom) & mean_ice == 0,
    mean_ice = if_else(land, NA_real_, mean_ice)
  ) |>
  select(land, mean_anom, mean_ice)

names(summary_2009)
dim(summary_2009)

```

`st_apply(...)` produces a single grid per attribute from collapsing the 31-day `time` dimension. 

`mean_anom` and `na.rm = TRUE` interprets that missing `anom` across 31 days will return `NaN`.

`mean_ice` missing values are replaced with 0 before averaging so all missing cells return 0.

`mutate(...)` flags `land` as `TRUE` when `mean_atom` is `NaN` and `mean_ice` is 0 and sets `mean_ice` to `NA`.

### 3.5 Verifier Task
```{r}
#| label: verify-3-5
#| eval: true

st_dimensions(summary_2009)
st_dimensions(summary_2010)

names(summary_2009)
names(summary_2010)

sum(is.na(summary_2009$land))
sum(is.na(summary_2010$land))

all(is.na(summary_2009$mean_ice) == summary_2009$land)
range(summary_2009$mean_ice[!summary_2009$land], na.rm = TRUE) # within [0,1]

all(is.na(summary_2010$mean_ice) == summary_2010$land)
range(summary_2010$mean_ice[!summary_2010$land], na.rm = TRUE)

st_bbox(summary_2009) == st_bbox(dec_2009)
st_res(summary_2009)
st_crs(summary_2009) == st_crs(dec_2009)

st_bbox(summary_2010) == st_bbox(dec_2010)
st_res(summary_2010)
st_crs(summary_2010) == st_crs(dec_2010)
```

`st_dimensions()` confirms both summaries are 1440 by 720 with no `time` dimension - the 31 day collapse was successful.

`names()` confirms exactly the three required attributes: `land`, `mean_anom` and `mean_ice`.

`sum(is.na(...$land))` should return 0 for both interpreting that `land` has no missing values.

`all(is.na(...$mean_ice) == ...$land)` should return `TRUE` confirming `mean_ice` is `NA` perculiarly at land cells and no other places.

`range()` on non-land `mean_ice` values should fall within `[0, 1]`.

`st_bbox()`, `st_res()` and `st_crs()` comparisons confirm the spatial metadata was preserved.

## 3.6 Derive Surface Classes
```{r}
#| label: derive-surface

summary_2009 <- summary_2009 |>
  mutate(
    surface = case_when(
      land ~ "land",
      mean_ice >= 0.15 ~ "sea ice",
      is.na(mean_anom) ~ "sea ice",
      TRUE ~ "open ocean"
    )
  )

summary_2010 <- summary_2010 |>
  mutate(
    surface = case_when(
      land ~ "land",
      mean_ice >= 0.15 ~ "sea ice",
      is.na(mean_anom) ~ "sea ice",
      TRUE ~ "open ocean"
    )
  )

table(summary_2009$surface, useNA = "ifany")
table(summary_2010$surface, useNA = "ifany")

```

`case_when(...)` introduces the classification rules in priority order.

`is.na(mean_atom)` fallback catches non-land cells at the polar edge lacking an anomaly value - without it, visible holes would appear in the open-ocean layer.

### 3.6 Verifier Task
```{r}
#| label: verify-3-6
#| eval: true

table(summary_2009$surface, useNA = "always")
table(summary_2010$surface, useNA = "always")

all(summary_2009$surface[which(summary_2009$mean_ice >= 0.15)] == "sea ice")
all(summary_2010$surface[which(summary_2010$mean_ice >= 0.15)] == "sea ice")

non_land_2009 <-!summary_2009$land
all(summary_2009$surface[non_land_2009 & is.na(summary_2009$mean_anom)] == "sea ice")

non_land_2010 <-!summary_2010$land
all(summary_2010$surface[non_land_2010 & is.na(summary_2010$mean_anom)] == "sea ice")

oo_2009 <-summary_2009$surface == "open ocean"
all(!is.na(summary_2009$mean_anom[oo_2009]))
all(summary_2009$mean_ice[oo_2009] < 0.15)

oo_2010 <-summary_2010$surface == "open ocean"
all(!is.na(summary_2010$mean_anom[oo_2010]))
all(summary_2010$mean_ice[oo_2010] < 0.15)
```

`table(..., useNA = "always")` accounts for `"land"`, `"open_ocean"` and `"sea_ice"` only with `NA` summing at 0.

The first pair of `all()` verifies the `mean_ice >= 0.15` rule every cell meeting that threshold is classified as `"sea_ice"`.

The second pair confirms for the fallback rule.

The final four checks for open-ocean integrity meaning every `open ocean` cell has a valid `mean_anom` and `mean_ice` below 0.15.

## 3.7 Combine the Monthly Summaries Along a Year Dimension
```{r}
#| label: combine-summaries

dec_summaries <- c(
  summary_2009,
  summary_2010,
  along = "year"
) |>
  select(mean_anom, surface)

names(dec_summaries)
st_dimensions(dec_summaries)
dim(dec_summaries)

```

`c(..., along = "year")` produces a single `stars` object when stacking the two monthly summaries with a new `year` dimension.

`select()` drops `land` and `mean_ice` which are redundant.

### 3.7 Verifier Task
```{r}
#| label: verify-3-7
#| eval: true

st_dimensions(dec_summaries)

st_get_dimension_values(dec_summaries, "year")
names(dec_summaries)

st_bbox(dec_summaries) == st_bbox(summary_2009)
st_bbox(dec_summaries) == st_bbox(summary_2010)
st_res(dec_summaries)
st_crs(dec_summaries) == st_crs(summary_2009)
st_crs(dec_summaries) == st_crs(summary_2010)
```

`st_dimensions()` confirms the added up object has `x`, `y` and `year`.

`st_get_dimensions_values(..., "year")` should return 2009 and 2010.

`names()` confirms only `mean_anom` and `surface` remain.

`st_bbox()`, `st_res()` and `st_crs()` comparisons deduce that spatial metadata is consistent with both input summaries.

## 3.8 Draft a First Faceted Raster Map
```{r}
#| label: draft-raster

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = mean_anom)) +
  facet_wrap(vars(year))

```

`geom_stars(...)` renders the raster grid and `facet_wrap(vars(year))` separates the two Decembers into separate panels.

1. The default colors are unsuitable for mapping the anomaly. In Figure 1.1, it uses a diverging palette to help map the positive and negative colors

2. The land masses are squished, and the area is uninterpretable as a data value. The map exaggerates polar regions because it is drawn in longitude and latitude rather than an equal-area projection.

3. There is no mapping for the surface/land, it is only patches of gray

4. Land and sea ice are not separated from open ocean, so the surface pattern is ambiguous.

## 3.9 Design a Diverging Fill Scale
```{r}
#| label: diverging-fill-scale

dec_summaries <- dec_summaries |>
  mutate(
    open_ocean_anom = if_else(surface == "open ocean", mean_anom, NA_real_)
  )

limit <- dec_summaries |>
  pull(open_ocean_anom) |>
  abs() |>
  quantile(0.99, na.rm = TRUE) |>
  ceiling()

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = mean_anom)) +
  facet_wrap(vars(year)) +
  scale_fill_fermenter(
    palette = "RdYlBu",
    direction = -1,
    limits = c(-limit, limit),
    oob = scales::squish
  )
```

`mutate()` produces `open_ocean_anom` which equals `mean_anom` at open-ocean cells and `NA` elsewhere - this prevents the colour scale range to be influenced by under-ice anomalies.

`quantile(0.99)` on the absolute values, rounded up using `ceiling()` in giving the symmetric `limit` tibble.

`scale_fill_fermenter(...)` incorporates the reversed RdYlBu diverging palette (blue = cold, red = warm) and `oob = scales::squish` clamps extreme values to the endpoint colours rather than voiding them.

## 3.10 Add Land and Sea Ice as Surface Layers
```{r}
#| label: add-land-and-seaice

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = open_ocean_anom)) +
  facet_wrap(vars(year)) +
  scale_fill_fermenter(
    palette = "RdYlBu",
    direction = -1,
    limits = c(-limit, limit),
    oob = scales::squish
  ) +
  ggnewscale::new_scale_fill() +
  geom_stars(data = dec_summaries, aes(fill = surface)) +
  scale_fill_manual(
    values = c(
      land = "grey70",
      `sea ice` = "white",
      `open ocean` = "transparent"
    ),
    breaks = c("land", "sea ice"),
    labels = c("Land", "Sea-ice concentration >= 15%"),
    name = NULL
  ) 

```

`new_scale_fill()` begins a second fill so both continuous(anomaly) and categorical(surface) scale can coexist.

The surface layer has to sit above the anomaly layer because it acts asa m ask that controls what hte viewer can see. Land and sea ice should cover the anomlay colors wherever they occur, while opean ocean must stay transparent so the anomaly remains visible underneath. If the surface layer were drawn first, the anomaly layer would paint over it and the categorical distinction would be lost.

### 3.10 Verifier Task
```{r}
#| label: verify-3-10
#| eval: true

oo <-dec_summaries$surface == "open ocean"

all(dec_summaries$open_ocean_anom[oo] == dec_summaries$mean_anom[oo], na.rm = TRUE)
all(is.na(dec_summaries$open_ocean_anom[!oo]))
```

The first `all()` verifies for the mask preserved anomaly values abided at open-ocean cells.

The second `all()` verifies for the mask being correctly blanked out anomaly values at land and sea ice cells.

## 3.11 Compare Equal-Area World Projections

```{r}
#| label: compare-equal-area

equal_earth_crs <- st_crs("+proj=eqearth +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs")
mollweide_crs <- st_crs("+proj=moll +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs")

warp_and_plot <- function(crs) {
  keep <- which(abs(st_get_dimension_values(dec_summaries, "y")) <= 89.5)
  trimmed <- dec_summaries[, , keep, ]
  
  target <- st_as_stars(
    st_bbox(st_transform(trimmed, crs)),
    dx = 50000, # 50 km cells (CRS units are meters)
    dy = 50000
  )
  
  warped <- st_warp(trimmed, target, method = "near")
  
  bbox_latlon <- st_bbox(c(xmin = 0.001, xmax = 359.999, ymin = -89.5, ymax = 89.5), crs = 4326)
  world_boundary <- st_as_sfc(bbox_latlon) |>
    st_segmentize(dfMaxLength = 1) |>
    st_transform(crs)
  
  warped_cropped <- st_crop(warped, world_boundary)
  
  ggplot() +
    geom_stars(data = warped_cropped, aes(fill = open_ocean_anom)) +
    scale_fill_fermenter(
      palette = "RdYlBu",
      direction = -1,
      limits = c(-limit, limit),
      oob = scales::squish
    ) +
    ggnewscale::new_scale_fill() +
    geom_stars(data = warped_cropped, aes(fill = surface)) +
    facet_wrap(vars(year), ncol = 1) +
    scale_fill_manual(
      values = c(
        land = "grey70",
        `sea ice` = "white",
        `open ocean` = "transparent"
      ),
      breaks = c("land", "sea ice"),
      labels = c("Land", "Sea-ice concentration >= 15%"),
      name = NULL
    ) +
    coord_equal()
}

map_equal_earth <- warp_and_plot(equal_earth_crs)
map_equal_earth

map_mollweide <- warp_and_plot(mollweide_crs)
map_mollweide

```

`warp_and_plot` produces a reusable function by wrapping the reprojection pipeline. It trims poles to +-89.5 degrees, builds a 50 km target grid in projected CRS, preserving `surface` as a factor and deters blurring at coastlines by warpping via resampling then draws the two-layer map with `coord_equal()`.

a. How does each projection represent the polar regions?

The Equal Earth projection represents the poles as straight, horizontal lines across the top and bottom of the map. This stretches high-latitude landmasses horizontally but maintains a familiar overall world shape.

The Mollweide projection represents the poles as single points, this pinches the high-latitude landmasses much more sharply and produces a narrower, more pointed silhouette at the top and bottom.

b. How clearly does each projection show the equatorial Pacific?

Both projections show the equatorial Pacific very clearly because they are centered on the 180 degrees meridian, placing the El Nino and La Nina temperature anomalies dead center. As they are both equal-area projections, the physical size of the anomaly regions relative to the rest of the globe is accurately preserved in both.

c. Which projection uses the rectangular panel space more effectively?

The Equal Earth projection uses the rectangular panel space more effectively. As its poles are stretched into lines rather than points, the map takes up more of the rectangular panel, leaving less empty gray space in the four corners compared to the curving Mollweide projection.

d. Which projection makes the contrast between the two Decembers more visually prominent?

While Equal Earth fills more of the panel, Mollweide makes the contrast between the two Decembers more visually prominent by concentrating the viewer's focus. As Mollweide's oval converges at the poles, it naturally draws the eye straight to the center of the frame, where the equatorial Pacific anomaly resides.

By minimizing peripheral distractions at high latitudes, Mollweide forces a direct, centered vertical comparison between the red El Nino plume in 2009 and the blue La Nina wave in 2010, making the core signal easier to isolate.

I would choose Mollweide for the final map. It sacrifices some rectangular panel efficiency compared to Equal Earth, and compresses the poles more aggressively, but since the anomalies of interest are concentrated in the equatorial and mid-latitude Pacific, that polar distortion has little practical cost here. The compact elliptical form also keeps the data-rich areas closer to the visual center of the panel, which we found to make the two December panels easier to compare side by side.

## 3.12 Build the Two-Panel Anomaly Map {#build-the-two-panel-anomaly-map}

```{r}
#| label: data-preparation

keep <- which(abs(st_get_dimension_values(dec_summaries, "y")) <= 89.5)
trimmed <- dec_summaries[, , keep, ]
  
target <- st_as_stars(
  st_bbox(st_transform(trimmed, mollweide_crs)),
  dx = 50000, # 50 km cells (CRS units are meters)
  dy = 50000
)

warped <- st_warp(trimmed, target, method = "near")
  
bbox_latlon <- st_bbox(c(xmin = 0.001, xmax = 359.999, ymin = -89.5, ymax = 89.5), crs = 4326)
world_boundary <- st_as_sfc(bbox_latlon) |>
  st_segmentize(dfMaxLength = 3) |>
  st_transform(mollweide_crs)

warped_cropped <- st_crop(warped, world_boundary)

equator_line <- st_linestring(matrix(c(seq(0.001, 359.999, length.out = 100), rep(0, 100)), ncol = 2)) |>
  st_sfc(crs = 4326) |>
  st_segmentize(dfMaxLength = 3) |>
  st_transform(mollweide_crs)

ocean_labels <- st_sf(
    name = c("Pacific\nOcean", "Atlantic\nOcean", "Indian\nOcean"),
    geometry = st_sfc(
      st_point(c(190, 20)),   
      st_point(c(315, 25)),
      st_point(c(75, -15)),
      crs = 4326
    )
  ) |> 
  st_transform(mollweide_crs)

ocean_coords <- as.data.frame(st_coordinates(ocean_labels))
ocean_coords$name <- ocean_labels$name

equator_df <- as.data.frame(st_coordinates(equator_line))

world_boundary_df <- as.data.frame(st_coordinates(world_boundary))

```

`keep` tibble chooses the absolute latitude is at most 89.5 degrees row indices, trimming the extreme poles falling outside Mollweide's domain. 

`trimmed` subsets `dec_summaries` of those along the `y` dimension.

`st_transform(trimmed, mollweide_crs)` reprojects the trimmed raster's metadata into Mollweide CRS. `st_bbox()` then extracts the bounding box in metres. `st_as_stars()` builds an empty target grid over the bounding box with 50km by 50km which determines the resolution of the final map.

`st_warp(...)` resamples the raster onto the target grid using the interpolation.

`bbox_latlon` defines a bounding box in WGS 84 spanning 0.001-359.999 degrees longitude and +-89.5 degrees latitude. 

`st_as_sfc()` coverts it to a polygon, `st_segmentize(dfMaxLength = 3)` densifies its edges with points every 3 degrees, reason being when `st_transform(mollweide_crs)` projects it, the straight edges will curve to follow Mollweide's elliptical outline instead of cutting across it.

`st_crop(warped, world_boundary)` clips the warped raster to projected boundary, masking out the rectangular corner artifacts falling outside of the ellipse.

`st_set_limitations("year", values = ...)` relabels the `year` dimension from `2009` / `2010` to `December 2009 (El Nino)` / `December 2010 (La Nina)` so `facet_wrap()` uses this as panel strip titles.

`equator_line` forms a line at latitude 0 degrees from 100 evenly spaced points spanning the full longitude range, densifies and transforms it with `st_segmentize()` to Mollweide so it curves nicely on the projected map.

`ocean_labels` produces three labelled points at roughly the centres of the Pacific (180 degrees, 0 degrees), Atlantic (330 degrees, 0 degrees) and Indian (80 degrees, -20 degrees) Ocean in WGS 84 then to Mollweide. `st_coordinates()` extracts projected X/Y values into a data frame so `geom_label()` can place text at those positions.

`equator_df` and `world_boundary_df` also extract coordinates into data frames for use with `geom_path()` and `geom_polygon()` respectively.


```{r}
#| label: continuous-scale-plot
#| warning: false

make_sst_map <- function(anomaly_scale, anomaly_mapping = aes(fill = open_ocean_anom)) {
  ggplot() +
    geom_stars(data = warped_cropped, anomaly_mapping) +
    anomaly_scale +
    ggnewscale::new_scale_fill() +
    geom_stars(data = warped_cropped, aes(fill = surface)) +
    facet_wrap(
      vars(year), 
      ncol = 1,
      labeller = labeller(year = as_labeller(c(
        "1" = "December 2009 (El Niño)",
        "2" = "December 2010 (La Niña)"
      )))
    ) +
    scale_fill_manual(
      values = c(
        land = "grey70",
        `sea ice` = "white",
        `open ocean` = "transparent"
      ),
      breaks = c("land", "sea ice"),
      labels = c("Land", "Sea-ice concentration >= 15%"),
      name = NULL,
      na.value = NA,
      guide = guide_legend(
        order = 2,
        override.aes = list(colour = "black", linewidth = 0.3)
      )
    ) +
    geom_path(
      data = equator_df, aes(x = X, y = Y),
      colour = "grey60", linetype = "dashed", linewidth = 0.35
    ) +
    geom_polygon(
      data = world_boundary_df, aes(x = X, y = Y, group = L1),
      fill = NA, colour = "black", linewidth = 0.4
    ) +
    geom_label(
      data = ocean_coords,
      aes(x = X, y = Y, label = name),
      fill = alpha("white", 0.6),
      color = "grey20",
      size = 2.8,
      fontface = "plain",
      linewidth = 0,
      label.padding = unit(0.15, "lines")
    ) +
    coord_equal(clip = "off") +
    labs(
      title = "Sea Surface Temperature Anomalies During El Nino and La Nina",
      subtitle = "Monthly mean departure from 1971–2000 climatology",
      caption = "Source: NOAA Optimum Interpolation Sea Surface Temperature (OISST) v2.1"
    ) +
    theme_void() +
    theme(
      plot.title.position = "plot",
      strip.text = element_text(face = "bold", size = 11, margin = margin(b = 5, t = 5)),
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 10, color = "grey30", hjust = 0, margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "grey40", hjust = 1, margin = margin(t = 10)),
      legend.position = "right",
    )
}

continuous_scale_obj <- scale_fill_fermenter(
  palette = "RdYlBu", 
  direction = -1,
  limits = c(-limit, limit), 
  oob = scales::squish, 
  na.value = NA,
  name = "Anomaly (°C)",
  guide = guide_colorsteps(
    order = 1,
    frame.colour = "black", 
    frame.linewidth = 0.3, 
    ticks.colour = "black"
  )
)

map_continuous <- make_sst_map(continuous_scale_obj)
map_continuous
```

`make_sst_map()` is a prototype function that takes an `anomaly_scale`(the ggplot2 scale object for the anomaly fill) and an optional `anomaly_mapping`(defaulting to `aes(fill = open_ocean_anom)`).

`geom_stars(data = warped_cropped, anomaly_mapping)` plots the first fill layer, the open-ocean anomaly raster using whichever aesthetic mapping that was passed in. `anomaly_scale` is added immediately after to control how those fill values are coloured.

`new_scale_fill()` is deployed so a second `geom_stars()` can map `surface` to its own independent fill scale via resetting the fill aesthetic. Without this call, ggplot2 would reject the use of two fill scales on the same plot.

`facet_wrap(vars(year), ncol = 1, labeller = ...)` separates the two year slices into vertically stacked panels. The `labeller` maps the internal dimension values `1` and `2` to the descriptive titles.

`scale_fill_manual()` delegates `"grey70"` to land, `"white"` to sea ice and `"transparent"` to open ocean so the anomaly layer shows through. `breaks` limits the legend to land and sea ice only. `guide = guide_legend(order = 2)` positions the surface legend below the anomaly legend and `override.aes` inputs a black border around each legend key for clarity.

`geom_path(data = equator_df, ...)` plots the equator as a dashed grey line using the afore-extracted Mollweide projected coordinates. `geom_polygon(data = world_boundary_df, ...)` traces the Mollweide elliptical boundary with a thin black frame and transparent fill. `geom_label(data = ocean_coords, ...)` positions the three ocean name labels via a semitransparent white background so text remains readable over any fill colour.

`coord_equal(clip = "off")` guarantees the projected coordinates are drawn with a 1:1 aspect ratio allowing labels near the edge to extend slightly without getting clipped.

`theme_void()` discards all axes, ticks and gridlines. The next `theme()` overrides restore only the elements needed.

`continuous_scale_obj` defines the diverging fill via `scale_fill_fermenter()` with the reversed RdYlBu palette, symmetric limits at `tlimit` and `squish` for out-of-bounds values. `guide_coloursteps(order = 1)` plots the anomaly legend as a stepped color bar placed above the surface legend.

`make_sst_map(continuous_scale_obj)` invokes the function with the continuous scale and default mapping to output the final continuous-scale-two-panel map.  

## 3.13 Bin the Anomaly Scale {#bin-the-anomaly-scale}

```{r}
#| label: binned-scale
#| warning: false

breaks <- seq(-limit, limit, by = 1)
fmt_breaks <- sprintf("%+g", breaks)
fmt_breaks <- sub("^\\+0$", "0", fmt_breaks)
bin_labels <- c(
  paste0("\u2264 ", fmt_breaks[1]),
  paste0(head(fmt_breaks, -1), " to ", tail(fmt_breaks, -1)),
  paste0("> ", fmt_breaks[length(fmt_breaks)])
)

binned_mapping <- aes(
  fill = cut(open_ocean_anom, breaks = c(-Inf, breaks, Inf), labels = bin_labels)
)

binned_scale_obj <- scale_fill_brewer(
  palette = "RdYlBu", 
  direction = -1,
  name = "Anomaly (°C)",
  na.translate = FALSE,
  guide = guide_legend(
    order = 1,
    reverse = TRUE,
    override.aes = list(color = "black", linewidth = 0.3)
  )
)

map_binned <- make_sst_map(binned_scale_obj, binned_mapping)
map_binned
```

`breaks <- seq(-limit, limit, by = 1)` produces whole-degree bin boundaries from `-limit` to `+limit`. `fmt_breaks <- sprintf("%g", breaks)` formats them with explicit `+`/`-` polarity for ease of read. `fmt_breaks <- sub(...)` helps to remove the `"+"` from 0.

`bin_labels` constructs the legend text for each class. The first element is the lower class(e.g. `"<= 3"`), the middle elements (e.g. `"-3 to -2, .."`) and the last element is the upper class (e.g. `"> +3"`).

`binned_mapping` substitutes the default `aes(fill = open_ocean_anom)` with `aes(fill = cut(...))`. `cut()` delegates each cell's anomaly into one of the labelled bins with `Inf` to be the outermost boundaries to catch values beyond `+-limit` classes.

`binned_scale_obj` uses `scale_fill_brewer()` instead of `scale_fill_fermenter()` as the fill is now a discrete factor. The same reversed RdYlBu palette is incorporated. `na.translate = FALSE` masks `NA` values from the legend. `guide_legend(order = 1, reverse = TRUE)` showcases the bins from highest(red, top) to lowest(blue, bottom) and `override.aes` adds black borders around each legend.

`make_sst_map()` invokes the shared function with the binned scale and the `cut()`. 

## 3.14 Reflect on the Visualization

Q1: What does the two-panel map reveal about the difference between the December 2009 El Niño and the December 2010 La Niña? Point to specific regions.

A1: The two-panel map reveals a contrast across several key regions, driven by the shift between El Nino (2009) and La Nina (2010):

- Pacific Ocean: In December 2009, a strong, continuous "warm tongue" of positive anomalies (red) spans the central and eastern equatorial Pacific, with sea surface temperatures exceeding 2 to 3 degree celsius above average. By December 2010, this region completely reverses into a broad band of cold negative anomalies (blue), marking the La Nina cooling pattern.
- North Pacific & Mid-Latitudes: In 2009, cooler anomalies (blue) surround the warm equatorial core. In 2010, this pattern flips, warm anomalies (red) flank the central cold La Nina pattern to the north and south.
- Indian and Atlantic Oceans: The warmth of the 2009 El Niño extends into the Indian Ocean, which shows widespread positive anomalies (yellow/orange). In contrast, the 2010 panel shows a cooling trend across the Indian Ocean alongside strong regional warming in the North Atlantic.

Overall, the map illustrates that the 2009–2010 shift was not just localized to the equator, but more of a global-scale sea surface temperature adjustment.

Q2: The map draws the anomaly for open ocean only and collapses sea ice into a single class. What does the reader gain and lose from these two choices?

A2: By drawing temperature anomalies only for the open ocean, the map makes it much easier to identify the characteristic warming during El Nino and cooling during La Nina. Since sea surface temperature anomalies are only meaningful where there is exposed ocean water, masking out sea ice avoids displaying misleading or unavailable data and produces a cleaner, more interpretable map.

However, this choice also means the reader cannot see temperature variations beneath or within sea-ice-covered regions, making it impossible to examine how polar ocean temperatures may differ or change near the ice edge.

Collapsing sea ice into a single class also improves readability by clearly distinguishing sea ice from both land and open ocean without adding another variable to interpret. This keeps the focus on the sea surface temperature anomalies.

The downside is that the reader loses information about differences in sea-ice concentration. Areas with thin, partial ice and areas with nearly 100% ice cover appear identical, so changes in ice extent or concentration cannot be assessed from the map.

Q3: You now have two versions of the same map, one with a continuous scale ([Section 3.12](#build-the-two-panel-anomaly-map)) and one with a binned scale ([Section 3.13](#bin-the-anomaly-scale)). The maps are intended to compare the extent of warm and cool anomalies between the two Decembers. Which scale serves that goal better, and what does each sacrifice?

A3: The binned scale serves the goal better. By grouping continuous temperature values into discrete, bounded classes (e.g., +1 to +2 °C), it creates clear visual categories that make it easier to compare the spatial extent of anomaly thresholds across the two Decembers. The reader can more readily assess how far the El Niño warm tongue or La Niña cold plume extends without being distracted by minor colour variations.

However, the binned scale sacrifices precision and internal variation. Cells with distinct anomaly values (such as +1.1 °C and +1.9 °C) are rendered with the same fill colour. This flattens fine-grained temperature gradients and obscures subtle regional differences within each class.

In contrast, the continuous scale preserves the full spectrum of temperature anomalies and reveals smooth spatial transitions. The trade-off is that its continuous range of colour shades makes it more difficult to judge the overall extent of warm and cool anomalies or identify where specific anomaly thresholds begin and end.