---
title: "NYC-Delays-Lab"
author: "Team Orchid"
date: "2026-06-09"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | Contribution |
|---|---|---|
| Quan Yi | Coder | Coded out the tasks in the QMD and ensure code rendering |
| Shi Min | Navigator | Read task requirements, ensured all questions were addressed, assisted when needed, and maintained QMD readability. |
| Aiman | Reporter | Detailed the Markdown explanation for the code snippets | 
| Subhashini | Verifier | Validated plots/calculations and completed the verifier tasks. | 
| Jabar | AI Liaison | Queried the AI along the Coder, log AI interaction and ensured AI output quality |
| Min Xuan | Constructive Critic | Reviewed merged content, suggests refinements, and ensures final ZIP integrity |

## Setup

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(nycflights13)
library(ggrepel)
```

## 3.1 Data Preparation

### 3.1a Creating Tibble

```{r}
#| label: data-preparation

departures <- flights |>
    drop_na(dep_delay)

departures
```

We start to create a new tibble named `departures` from the `flights` tibble in the `nycflights13` package 

- `drop_na()` is to remove any row where `dep_delay` is `NA`

### 3.1b Select and rename columns, create new variables
```{r}
#| label: select-rename-create

departures <- departures |>
  mutate(
    airport = replace_values(
      origin,
      "EWR" ~ "Newark",
      "LGA" ~ "LaGuardia",
    )) |>
    mutate(on_time = dep_delay < 15) |>
    select(airline_code = carrier, airport_code = origin, airport, hour, on_time)

departures
```

We now create two new variables and reduce `departures` to only the columns needed

- The first `mutate()` is to convert the three letter IATA airport codes in `origin` into human-readable names
- The second `mutate()` creates a logical (`TRUE`/`FALSE`) column
- `select()` keeps only the five required columns and renames two of them 

### 3.1c Join tables

```{r}
#| label: joining-with-carrier

carriers <- read_csv("carriers.csv")

departures <- departures |>
  left_join(carriers, by = "airline_code") |>
  select(airline, airport, hour, on_time)

departures

```

The `airline_code` column uses two letter IATA codes which are not self-explanatory to public. We replaced them using a lookup table

- `read_csv()` imports the relevant csv dataset
- `left_join()` matches each row in `departures` to its subsequent airline name in `carriers` using `airline_code` as the referral
- `select()` keeps only the four required columns

### 3.1 Verifier Task

```{r}
#| label: verifier-task-3.1

# Record row count before the join (after filtering NAs)
rows_before_join <- flights |>
  drop_na(dep_delay) |>
  nrow()

# Row count after the join and column selection
rows_after_join <- nrow(departures)

# Check: row counts must match (join should not add or drop rows)
stopifnot(rows_before_join == rows_after_join)
cat("Row count before join:", rows_before_join, "\n")
cat("Row count after join: ", rows_after_join, "\n")
cat("Row counts match:", rows_before_join == rows_after_join, "\n")

# Check: no missing values in final departures tibble
missing_vals <- sum(is.na(departures))
cat("Missing values in departures:", missing_vals, "\n")
stopifnot(missing_vals == 0)
```

A join can silently add rows if the lookup table contains duplicate airline codes

- `flights |> drop_na(dep_delay) |> nrow()` recomputes by applying only the cancellation filter to the expected row count of raw `flights` tibble, giving a clean reference value that is independent of the join 
- `nrow(departures)` counts rows in the final `departures` tibble after the join and column selection
- `stopifnot(rows_before_join == rows_after_join)` halts execution with an error if the counts differ, preventing silently incorrect data
- `cat()` print both counts and their comparison to the rendered output 
- `sum(is.na(departures))` counts every `NA` across all columns and rows of `departures`
- `stopifnot(missing_vals == 0)` — halts if any missing values remain, catching unmatched airline codes that the join would have left as `NA`

## 3.2 Airlines Ranked by Departures
### 3.2a Creating Tibble `departures_by_airline`

```{r}
#| label: ranked-by-departure

departures_by_airline <- departures |>
  summarize(
    n = n(),
    on_time_pct = mean(on_time) * 100,
    .by = airline
  ) |>
  mutate(airline = fct_reorder(airline, n)) |>
  arrange(n)

departures_by_airline
levels(departures_by_airline$airline)

departure_labs <- function (...) {
    ggplot2::labs(
        caption = 'Source: R package "nycflights13"',
        ...
    )
}

```

We first determine the number of departures each airline made from airports in the New York City area in 2013 before comparing carriers based on their on-time records by creating a summary tibble `departures_by_airline` with one row per airline consisting of its total departure count and on-time percentage

- `summarize()` collapses rows by each airline into a single summary row
- `n = n()` counts the total number of departures for each airline
- `on_time_pct` returns the proportion of on-time departures by multiplying hundred which converts it to a percentage
- `fct_reorder()` converts `airline` in ascending order of `n`
- `arrange()` sorts the rows of tibble itself
-  `levels()` confirms whether the factor level matches the ascending sort of `n`
- `departure_labs()` is a reusable function which appends the data source caption to any plot 

### 3.2b Horizontal Lollipop Chart

```{r}
#| label: lollipop-departures
#| fig-width: 7
#| fig-height: 6

ggplot(departures_by_airline, aes(x = n, y = airline)) +
    geom_segment(aes(x = 0, xend = n, yend = airline), color = "grey") +
    geom_point(size = 3, color = "blue") +
    departure_labs(
        title = "Number of Departures by Airline from NYC-Area Airports", 
        y = NULL
    ) 

```

We then draw a horizontal lollipop chart to rank airlines based on their overall number of departures

- `ggplot()` initialises the plot with x-axis the departure count and y-axis the airline name
- since `airline` is an ordered factor, ggplot2 automatically ranks the airlines in the right order
- `geom_segment()` draws a horizontal stem where starts at zero for each airline and ends at its departure count; `color = "grey"` keeps stem noticeable so the dots stand out
`geom_point()` draws a dot at the end of each stem at position `(n, airline)`; `size = 3` refers to the dot dimension enough to see clearly
`departure_labs()` adds the data source caption and plot title

### 3.2c Advantage of Placing the Airline Names on the y-axis Rather than the x-axis

The advantage of placing the airline names on the y-axis rather than the x-axis in
this plot is that there are 16 airlines and the airline names are relatively long. If it was to be plotted on the x-axis, it may have to be tilted to fit all the airline names along the x-axis, making it harder to read

### 3.2d Labelling Lollipop

```{r}
#| label: labelled-lollipop

ggplot(departures_by_airline, aes(x = n, y = airline)) +
    geom_segment(aes(x = 0, xend = n, yend = airline), color = "grey") +
    geom_point(size = 3, color = "blue") +
    geom_text(aes(label = n), hjust = 0, nudge_x = 500) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    departure_labs(
        title = "Number of Departures by Airline from NYC-Area Airports", 
        x = NULL,
        y = NULL
    ) +
    guides(x = "none")
```

We incorporate count labels directly to the chart to discard the redundant x-axis eliminating duplication

- `geom_text(aes(label = n))` places the departure count as a text label at each dot position
- `hjust = 0` left-aligns the label 
- `nudge_x = 500` shifts the label 500 units to the right so it does not overlap with the dot
- `expansion(mult = c(0, 0.15))` adds no padding on the left side and adds 15% extra space on the right
- `x = NULL` discards the x-axis title
- `guides(x = "none")` discards the x-axis ticks and text wholely.  

### 3.2e One Advantage and Disadvantage of a Lollipop Chart compared to a Bar Chart

One advantage of a lollipop chart is that there is less visual clutter compared to a bar chart as it swaps the thick bar for a thin line.

One disadvantage is that in datasets with few categories, e.g. 2 or 3 categories, it can cause the visualization to look sparse and a bar chart would be more suitable.

### 3.2 Verifier Task

```{r}
#| label: verifier-task-3.2

# sum of all airline departure counts must equal total rows in departures
total_from_summary <- sum(departures_by_airline$n)
total_from_departures <- nrow(departures)

cat("Sum of counts in departures_by_airline:", total_from_summary, "\n")
cat("nrow(departures):                       ", total_from_departures, "\n")
cat("Counts match:", total_from_summary == total_from_departures, "\n")
stopifnot(total_from_summary == total_from_departures)
```

The `summarize()` must account for every row in `departures` as in no flight should be lost or double-counted during aggregation
- `sum(departures_by_airline$n)` adds up the departure counts across all airlines in the summary tibble and the `$n` extracts only the numeric count column 
- `nrow(departures)` the number of non-canceled flights in the cleaned `departures` tibble 
- `stopifnot(total_from_summary == total_from_departures)` confirms the two totals are identical
- `cat()` display both values side by side so the match is immediately readable in the rendered output

## 3.3 On-Time Departures by Airline
### 3.3a Remove airlines that departed fewer than 1000 flights

```{r}
#| label: removal-of-airlines

filtered_departures <- departures_by_airline |>
    filter(n >= 1000)

filtered_departures
```

Airlines with a low number of departures may report an exceptionally high or low on-time due to the sample size being insufficient for statistical inference. Hence, we limit the research to airlines which have at least 1000 departures.

- `departures_by_airlines` is similar to the summary tibble from section 3.2
- `filter(n >= 1000)` keeps only rows where airline departure count is 1000 or more
- The result is saved as `filtered_departures`

### 3.3b Sort factor levels of `airline` in ascending order

```{r}
#| label: reorder-of-factor

filtered_departures <- filtered_departures |>
    mutate(airline = fct_drop(airline)) |>
    mutate(airline = fct_reorder(airline, on_time_pct)) 

levels(filtered_departures$airline)
```

We amend the factor order of `airline` so the y-axis reflects the punctuality order rather than volume ranking.

- `fct_drop(airline)` discard unused factor levels 
- `fct_reorder(airline, on_time_pct)` reorders the factor levels in ascending order of `on_time_pct`
`levels()` makes sure the reordering was applied correctly

### 3.3c Cleveland Dot Plot

```{r}
#| label: cleaveland-dot-plot

ggplot(filtered_departures, aes(x = on_time_pct, y = airline)) +
    geom_point(size = 3, color = "blue") +
    geom_text(
        aes(label = round(on_time_pct, 1)),
        hjust = 0,
        nudge_x = 0.3,
        size = 3
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.05, 0.1))) +
    departure_labs(
        title = "On-Time Departure Rate by Airline from NYC-Area Airports",
        x = "On-Time Departure",
        y = NULL
    )

```

We then draw a Cleveland dot plot ranking airlines based on their on-time departure percentage. Unlike the lollipop chart in Section 3.2, this plot uses no stem and only dots as the x-axis is not starting from zero

- `fct_rev(airline)` reverses the factor order with the highest on-time percentage appearing at the top of y-axis
- `geom_point()` draws one dot per airline
- `geom_text(aes(label = round(on_time_pct, 1)))` inputs the on-time percentage as a text label next to each dot rounding off to one decimal place
- `hjust = 0` left-aligns label from its anchor point
- `nudge_x = 0.3` shifts the label 0.3% to the right avoiding overlapping with the dot
- `size = 3` config the text size to be small enough to not crowd each other
- `expansion(mult = c(0.05, 0.1))` adds 5% padding on the left and 10% on the right
- `departure_labs()` labels x and y-axis with the y-axis removed

### 3.3d Cleveland Dot Plot vs. Bar Chart Baselines

Bar and lollipop charts use length to show value, meaning they must start at zero. If we cut the baseline, it warps the visual scale a bar that looks twice as large should actually be twice the value. 

Cleveland dot plots are different because they show value through position. Since readers just compare where the dots sit relative to each other, we don't need a zero baseline. Therefore, we can safely zoom the axis into the data's actual range, maximizing visual detail to show clear differences between the airlines without being misleading.

### 3.3e Busiest Airline

The busiest airline is the United airline and it ranks 6 from the top of the punctuality ranking.

### 3.3 Verifier Task

```{r}
#| label: verifier-task-3.3

# Identify the busiest airline directly from departures_by_airline
busiest_airline <- departures_by_airline |>
  slice_max(n, n = 1) |>
  pull(airline) |>
  as.character()

cat("Busiest airline:", busiest_airline, "\n")

# Compute on-time % directly from departures tibble
direct_pct <- departures |>
  filter(airline == busiest_airline) |>
  summarize(on_time_pct = mean(on_time) * 100) |>
  pull(on_time_pct)

# Get the value from departures_by_airline
summary_pct <- departures_by_airline |>
  filter(as.character(airline) == busiest_airline) |>
  pull(on_time_pct)

cat("On-time % from departures_by_airline:", round(summary_pct, 4), "\n")
cat("On-time % computed directly:         ", round(direct_pct, 4), "\n")
cat("Values match:", isTRUE(all.equal(direct_pct, summary_pct)), "\n")
stopifnot(isTRUE(all.equal(direct_pct, summary_pct)))
```

- `slice_max(n, n = 1)` selects the single row with the largest departure count from `departures_by_airline`
- `pull(airline) |> as.character()` extracts the airline name as a plain string, stripping the factor class so it can be used in `filter()`
- `filter(airline == busiest_airline)` subsets `departures` to only the rows belonging to the busiest airline 
- `summarize(on_time_pct = mean(on_time) * 100)` recomputes the on-time percentage using the raw `on_time` logical values
- `all.equal(direct_pct, summary_pct)` compares the two values with a small numerical tolerance to account for floating-point arithmetic
- `isTRUE()` wraps `all.equal()` because it returns a character string describing differences when values do not match

## 3.4 Airport Mix by Airline
### 3.4a Airlines with at least 1,000 departures

```{r}
#| label: airport-mix-by-airline

airline_levels <- filtered_departures |>
    arrange(on_time_pct) |>
    pull(airline) |>
    as.character()

airport_mix <- departures |>
    filter(airline %in% airline_levels) |>
    summarize(n = n(), .by = c(airline, airport)) |>
    mutate(
        pct = n / sum(n) * 100,
        .by = airline
    ) |>
    mutate(airline = factor(airline, levels = airline_levels)) |>
    arrange(airline, pct)

airport_mix
```

We compute each of the airline its departure percentage that came from each of the three NYC-area airports

- `arrange(on_time_pct)` sorts `filtered_departures` in ascending order of on-time percentage to match the punctuality ranking in Section 3.3
- `pull(airline)` extracts the `airline` column as plain character vector
- `as.character()` strips factor class for reassigning later without inheriting any prior ordering
- `airline_levels` stores the correct airline ordering as a character vector and to be used in `airport_mix`
- `filter(airline %in% airline_levels)` maintains only airlines that passed the 1000 threshold 
- `summarize(n = n(), .by = c(airline, airport))` counts the number of departures for every airline-airport
- `mutate(pct = n/sum(n) * 100, .by = airline)` computes each airline total departure
- `factor(airline, levels = airline_levels)` converts `airline` back to a factor using the punctuality ordered levels `airline_levels`
- `arrange(airline, pct)` sorts the rows by airline and then percentage   

### 3.4b Horizontal, Stacked Bar Chart

```{r}
#| label: stacked-bar
#| fig-width: 8
#| fig-height: 5

ggplot(airport_mix, aes(x = pct, y = airline, fill = airport)) +
  geom_col() +
  departure_labs(
    title = "Share of Departures by Airport for Each Airline",
    x = "Percentage of Departures",
    y = NULL,
    fill = "Airport"
  )

```

We draw a horizontal stacked bar chart showing how each airline departures are distributed across the three airports

- `aes(x = pct, y = airline, fill = airport)` maps the percentage to x-axis, the airline to y-axis and fills each segment by airport
- `geom_col()` draws stacked bars as `x = pct` is already computed, `geom_col()` is used instead of `geom_bar()`
- `departure_labs()` labels the respective x and y-axis with `fill = Airport` and `y = NULL` discarding y-axis title

### 3.4c Labelled Segments

```{r}
#| label: labelled-stack-bar

ggplot(airport_mix, aes(x = pct, y = airline, fill = airport)) +
    geom_col() +
    geom_text(
        aes(label = ifelse(pct >= 8, paste0(round(pct, 1), "%"), "")),
        position = position_stack(vjust = 0.5),
        size = 3,
        color = "white"
    ) +
    departure_labs(
        title = "Share of Departures by Airport for Each Airline",
        x = "Percentage of Departures",
        y = NULL,
        fill = "Airport"
    )


```

We add percentage labels inside every segment which its exact values could be read directly without needing to estimate from axis

- `geom_text(aes(label = ifelse(pct >= 8, ..)))` conditionally displays label if segment is wide enough to fit text
- `paste0(round(pct, 1), "%")` formats label by rounding to one decimal place and appending a symbol `%`
- `position_stack(vjust = 0.5)` places each label at the horizontal midpoint so the text sits cleanly 
`size = 3` keeps the text small enough to fit within the narrow segments
- `color = white` sets the color to white 


### 3.4d Which airport does the busiest airline depart from most, and how lopsided is that share?

In United airline, the most dominant airport is Newark. The split is highly lopsided with 78.7% of the flights originating from Newark, with LaGuardia being in second with 13.5%, and JFK being 7.8%. 

### 3.4 Verifier Task

```{r}
#| label: verifier-task-3.4

# Each airline's airport percentages must sum to 100% (within rounding tolerance)
airline_pct_sums <- airport_mix |>
  summarize(total_pct = sum(pct), .by = airline)

cat("Per-airline percentage sums:\n")
print(airline_pct_sums)

# all.equal checks they are all within tolerance of 100
check <- all.equal(
  airline_pct_sums$total_pct,
  rep(100, nrow(airline_pct_sums)),
  tolerance = 0.01
)
cat("All airline percentages sum to 100%:", isTRUE(check), "\n")
stopifnot(isTRUE(check))
```

- `summarize(total_pct = sum(pct), .by = airline)` adds up the three airport percentages for each airline
- `print(airline_pct_sums)` displays the full table so every airline's total is visible in the rendered submission 
- `rep(100, nrow(airline_pct_sums))` generates a vector of 100 with one entry per airline, forming the expected target for `all.equal()` 
- `all.equal(.., tolerance = 0.01)` allows a tolerance of 0.01 percentage points to account for floating-point rounding during division
- `isTRUE(check)` — safely converts `all.equal()`'s output to a logical for use in `stopifnot()`

## 3.5 On-Time Departures by Airport
### 3.5a Airport Number of Departures and the On-Time Percentage

```{r}
#| label: on-time-departures

departures_on_time <- departures |>
  summarize(
    n = n(),
    on_time_pct = mean(on_time) * 100,
    .by = airport
  ) |>
  arrange(desc(on_time_pct))

departures_on_time

```

We compute for each of the three NYC-area airports its total number of departures and the on-time departure percentage 

- `summarize()` similar to the previous sections collapses rows sharing same airport into a single summary row
- `arrange(desc(on_time_pct))` sorts from highest to lowest on-time percentage so the best performance airport is at the top

### 3.5b Dominant Departure Airport of the Busiest Airline also the Airport with the Lowest On-Time Percentage?

Yes, the Newark airport has the lowest on-time percentage of 74.7%. 

This suggests that United's rank of 6th in punctuality is likely dragged down by its heavy reliance on Newark. An airline that flies 78.7% of its flights from the worst-performing NYC airport will be disadvantaged in the punctuality comparison.

This implies that the airline’s raw rank partly reflect where it flies rather than its schedule-adjusted punctuality as looking at US Airways, which has a 87.4% punctuality rate, flies 63.3% of their flights from LaGuardia, which is the most punctual airport. Additionally, ExpressJet which has a 68.9% punctuality flies 81.3% of their flights from Newark. Therefore, the difference in ranking may say more about airport choice than airline quality.

### 3.5 Verifier Task

```{r}
#| label: verifier-task-3.5

# Overall on-time % across all departures
overall_pct <- mean(departures$on_time) * 100
cat("Overall on-time %:", round(overall_pct, 4), "\n")

# Departure-weighted mean of the three airport on-time percentages
weighted_pct <- with(departures_on_time, weighted.mean(on_time_pct, n))
cat("Weighted mean of airport on-time %:", round(weighted_pct, 4), "\n")

# Check they match
cat("Weighted mean matches overall:", isTRUE(all.equal(overall_pct, weighted_pct)), "\n")
stopifnot(isTRUE(all.equal(overall_pct, weighted_pct)))

# For comparison: unweighted (incorrect) average
unweighted_pct <- mean(departures_on_time$on_time_pct)
cat("Unweighted mean of airport on-time % (incorrect):", round(unweighted_pct, 4), "\n")
cat("Difference from true overall:", round(abs(unweighted_pct - overall_pct), 4), "pp\n")
```

- `mean(departures$on_time) * 100` computes the overall on-time percentage directly from every row of `departures`
- `weighted.mean(on_time_pct, n)` computes the departure-weighted average of the three airport on-time percentages, giving larger airports proportionally more influence
- `all.equal(overall_pct, weighted_pct)` confirms the weighted mean reproduces the overall percentage within floating-point tolerance
- `mean(departures_on_time$on_time_pct)` the plain unweighted average, which treats all three airports equally regardless of size
- `"pp\n"` — the unit `pp` stands for percentage points, the correct unit for differences between percentages

## 3.6 On-Time Departures by Time of Day and Airport
### 3.6a On-Time Percentage Computation

```{r}
#| label: on-time-percentage-hour

ontime_by_hour_airport <- departures |>
  summarize(
    on_time_pct = mean(on_time) * 100,
    .by = c(airport, hour)
  )

ontime_by_hour_airport
```

We compute the on-time departure percentage for every combination of airport and scheduled hour of departure

- `summarize()` similar to the previous sections but this time it collapses rows sharing same airport and hour into a single summary row 

### 3.6b Line-and-point plot

```{r}
#| label: line-and-point-plot

ggplot(
    ontime_by_hour_airport,
    aes(x = hour, y = on_time_pct, color = airport, linetype = airport)
    ) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(5, 23, by = 2)) +
    departure_labs(
        title = "On-Time Departure Rate by Hour and Airport",
        x = "Scheduled Departure Hour",
        y = "On-Time Departure",
        color = "Airport",
        linetype = "Airport"
    )

```

We draw a line and point plot showing how the changes of the on-time percentage across the hours of the day

- `aes(x = hour, y = on_time_pct)` inputs the x-axis to be of the scheduled departure hour and y-axis the on-time percentage
- `color = airport` assigns a distinct color to each airport line 
- `linetype = airport` assigns a distinct line type (solid, dashed, dotted) to each airport 
- `geom-line()` connects with a line for the hourly points of each airport showing a trend across the day
- `geom_point()` adds a dot at each hourly data point
- `scale_x_continuous(breaks = seq(5, 23, by =2))` sets the x-axis tick every two hours from 5 to 23, maintaining readibility without crowding every single hour
- `color = Airport` and `linetype = Airport` merges them into a single legend 


### 3.6c Two Visual Properties for One Variable Redundancy

Relying strictly on color to distinguish between the different data groups can make the visualization inaccessible to those with color deficiencies. With the line plots, the readers are able to tell the lines apart because one is solid and the others are dashed. 

### 3.6 Verifier Task

```{r}
#| label: verifier-task-3.6

# Confirm the declining trend: on-time % should generally fall across hours
# Check that the mean on-time % for hours <= 10 is higher than for hours >= 18
early_avg <- ontime_by_hour_airport |>
  filter(hour <= 10) |>
  summarize(mean_pct = mean(on_time_pct)) |>
  pull(mean_pct)

late_avg <- ontime_by_hour_airport |>
  filter(hour >= 18) |>
  summarize(mean_pct = mean(on_time_pct)) |>
  pull(mean_pct)

cat("Mean on-time % for hours <= 10:", round(early_avg, 2), "\n")
cat("Mean on-time % for hours >= 18:", round(late_avg, 2), "\n")
cat("Early hours more punctual than late hours:", early_avg > late_avg, "\n")
```

- `filter(hour <= 10)` selects the early morning slots from 5 to 10, where on-time rates are expected to be highest
- `filter(hour >= 18)` — selects the late evening slots from 18 to 23, where delay propagation from earlier flights is expected to have depressed
- `summarize(mean_pct = mean(on_time_pct))` averages the on-time percentages across all airports within each time window
- `pull(mean_pct)` — extracts the single computed value as a plain number for use in `cat()` and the comparison 
- `early_avg > late_avg` — the key logical check and if it should return `TRUE`, the data supports the declining trend visible in the plot but if it returns `FALSE`, the team should re-examine the line plot and the underlying data

**Verifier Task 3.6b — Evidence that On-Time Rate Declines Through the Day**

The line-and-point plot in Section 3.6b shows a clear downward trend in on-time departure rates as the day progresses across all three airports. This pattern is consistent with the known phenomenon of delay propagation described in airline operations research:

> "The average delays are larger in the evening. The reason for this is delay propagation. Delay propagation is a phenomenon in which delays caused by earlier flights using the same aircraft lead to later flights delayed because of the short maintenance times."

— Kawasaki et al. (2016). *Effect of Burstiness on the Air Transportation System*. arXiv:1608.01786. Available at: <https://arxiv.org/pdf/1608.01786>

## 3.7 Schedule Mix by Airline
### 3.7a Airline’s Own Departures Falling

```{r}
#| label: airline-hour-pct

airline_hourly_pct <- departures |>
  count(airline, hour, name = "n_flights") |>
  complete(airline, hour = 5:23, fill = list(n_flights = 0)) |>
  mutate(
    pct = (n_flights / sum(n_flights)) * 100,
    .by = airline
  )

airline_hourly_pct
```

We compute the percentage for each airline its own departures that fall in each scheduled hour of the day

- `count(airline, hour, name = "n_flights")` counts the number of departures for every airline-hour and names the count column `n_flights`
- `complete(airline, hour = 5:23, fill = list(n_flights = 0))` makes sure every airline has a row for every hour from 5 to 23. Even missing combinations are set to `n_flights = 0` rather than `NA` to avoid gaps in the heatmap
- `mutate(pct = n_flights/sum(n_flights) * 100, .by = airline)` ensures independent(adds up only that airline flights hourly percentages) sum to 100%  

### 3.7b Heatmap

```{r}
#| label: order-airline-rows

airline_performance <- departures |>
  summarize(on_time_pct = mean(on_time) * 100, .by = airline)

airline_performance

airline_hourly_pct <- airline_hourly_pct |>
  left_join(airline_performance, by = "airline") |>
  mutate(airline = fct_reorder(airline, on_time_pct))

airline_hourly_pct

```

We fixated each airline on-time percentage to match the punctuality ranking to the order of heatmap rows

- `summarize()` similar to the previous sections collapses rows sharing same airport into a single summary row

- `left_join(airline_performance, by = "airline")` brings the `on_time_pct` column into `airline_hourly_pct` for factor reordering; `left_join` preserves all rows in `airline_hourly_pct`
- `fct_reorder(airline, on_time_pct)` reorders the factor levels in ascending order of `on_time_pct` so the heatmap follows the same top-to-bottom ordering

```{r}
#| label: heatmap-plot

ggplot(airline_hourly_pct, aes(x = hour, y = airline, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.2) + 
  scale_x_continuous(breaks = seq(5, 23, by = 2)) +
  departure_labs(
    title = "Hourly Flight Distribution by Airline",
    x = "Scheduled Departure Hour",
    y = "Airline",
    fill = "Share of Daily Flights"
  )
```

We draw a heatmap where the fill color encodes what share of that airline flights depart in that hour with one tile per airline-hour

`aes(x = hour, y = fct_rev(airline), fill = pct)` inputs the x-axis to be of the hour, y-axis the airlines and maps the hourly percentage to the fill color; `fct_rev()` flips the factor order so the most punctual airline sits on top 

### 3.7c Heatmap Interpretation

Looking at the heatmap, United airlines seem to have consistent flights from the 6th hour all the way to the 20th hour. The early morning hours tend to be favourable times while the afternoon/evening hours tend to be more unfavourable times. However, this heatmap does not explain the punctuality rank of United airlines as US Airways also has a similar flight schedule but achieves a 87.4% on-time departure rate to United's 78.2% on-time departure rate

## 3.8 Observed versus Expected On-Time Departure Percentage

### Preparation of Data for Bubble Plot

```{r}
#| label: observed-and-expected-departure

slots <- departures |>
    summarize(
        n = n(),
        baseline_on_time_pct = mean(on_time) * 100,
        .by = c(airport, hour)
    )
slots

airline_performance <- departures |>
    left_join(slots, by = c("airport", "hour")) |>
    summarize(
        n = n(),
        on_time_pct = mean(on_time) * 100,
        expected_on_time_pct = mean(baseline_on_time_pct),
        .by = airline
    ) |>
    filter(n >= 1000) |>
    left_join(
        departures |>
        count(airline, airport) |>
        slice_max(n, n = 1, by = airline, with_ties = FALSE) |>
        select(airline, dominant_airport = airport),
        by = "airline"
    )

airline_performance

```

We first compute a baseline on-time rate for every airport-hour slot and then with those baselines to calculate each airline expected on-time percentage given its own schedule

- `summarize(.by = c(airport, hour))` groups every departures by airport and hour simultaneously
- `baseline_on_time_pct = mean(on_time) * 100` this is the circumstancial baseline the on-time percentage across all flights in that airport-hour slot

- The first `left_join(slots, by = c("airport", "hour"))` attaches the slot baseline to every row of `departures` so each flight carries both its own `on_time` and baseline rate value for its airport-hour slot
- `expected_on_time_pct = mean(baseline_on_time_plot)` the average of slot baselines across all airline departures
-  The second `left_join()` identifies each airline dominant airport by counting airline-airport with `count()` and taking the maximum with `slice_max()`
- `with_ties = FALSE` ensures exactly one dominant airport is returned per airline


### Bubble Plot

```{r}
#| label: bubble-plot

ggplot(
  airline_performance,
  aes(
    x = expected_on_time_pct,
    y = on_time_pct,
    size = n,
    color = dominant_airport,
    label = airline
  )
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    size = 3.5, 
    show.legend = FALSE,
  ) +
  scale_size_area(max_size = 15, guide = "none") +
  departure_labs(
    title = "Observed vs Expected On-time Percentages",
    x = "Expected On-Time",
    y = "Observed On-Time",
    color = "Dominant Airport"
  ) 

```

We draw a bubble plot to compare the rate expected from the airport-hour circumstances to each airline observed on-time percentage

- `aes(x = expected_on_time_pct, y = on_time_pct)` places the circumstantial baseline on the x-axis and the airline's actual performance on the y-axis
- `size = n` — scales each bubble's area by the airline's departure count so larger airlines are visually more prominent
- `color = dominant_airport` colours each bubble by the airport from which that airline operates the most flights
- `geom_abline(slope = 1, intercept = 0)` — draws the line of equality where observed equals expected
- `geom_point(alpha = 0.7)` draws the bubbles with slight transparency so overlapping points remain visible 
- `geom_text_repel()` — labels each bubble with its airline name, automatically repositioning labels to avoid overlap with other bubbles 
- `scale_size_area(max_size = 15, guide = "none")` maps `n` to bubble area rather than radius, capping the largest bubble at size 15; `guide = "none"` suppresses the size legend since the exact departure counts are less important than the relative sizes 
- `show.legend = FALSE` in `geom_text_repel()` — prevents the text labels from generating their own legend entry

## 3.9 Summary and Interpretation

United has an expected on-time percentage of roughly ~75.8%, and an observed on-time percentage of about 78.2%. This means that United airlines is actually more on-time than expected.

Looking at the line of equality, United should at least be ranked 5th instead of 6th in the raw punctuality rank. The 5th ranked airline in terms of punctuality was Envoy, but Envoy falls below the line of equality. Additionally, United’s dominant airport is Newark. Looking at the x-axis, the airlines that are dominantly Newark generally cluster on the far left, meaning Newark has the lowest "airport-hour baseline" expectations in the dataset.

Looking at ExpressJet which was at the bottom of the punctuality ranking, ExpressJet has an expected on-time percentage of roughly ~75.2%, but its observed percentage is far lower, at 68.9%. Like United, ExpressJet’s dominant airport is also Newark. Both airlines are situated on the far left of the x-axis, meaning they both face the same difficult airport-hour circumstances.

However, United manages to overperform despite the difficult Newark conditions, being above the line of equality and ExpressJet falls dramatically below the line. Since we know they are flying out of the same dominant airport, the difference suggests that ExpressJet's bottom-tier punctuality ranking is a result of genuine operational inefficiency. Their poor punctuality cannot be fully blamed on the airport or schedule and that it is an internal performance issue.

### 3.9a Raw on-time departure ranking and airport-hour circumstances

The raw on-time departure ranking reveals the absolute punctuality of each airline, showing passengers exactly who departs on schedule most frequently. 

However, it completely obscures the difficulty of the environments they fly in, unfairly penalizing airlines that operate heavily out of delay-prone airports. 

By accounting for these airport-hour circumstances, we can separate operational efficiency from schedule disadvantages, revealing that a middle-ranked airline like United is actually overperforming its difficult Newark schedule, while a bottom-ranked carrier like ExpressJet genuinely underperforms despite facing the same constraints.

### 3.9b Personal Reflections

#### Quan Yi

The hour column in the dataset does not seem the most impactful, and it could be debatable as of whether there is a difference for a morning flight vs an afternoon flight. To me, it is usually a preference thing as of whether you would want to start the trip at night to book in at a hotel, so you take the afternoon flight or take the morning/night flight so you get to start the trip once you land. However, after plotting the line-and-point-plot in `3.6b`, I noticed that all the airports tend to have a downdward trend of on-time departure from morning till evening, which could not be observed until plotted. This shows that morning flights tend to be on-time and that afternoon/evening flights tend to not be on-time.

#### Shi Min
Over the course of this lab, I was able to visually see the data, which allowed me to gain a deeper understanding compared to just reading it in a table. This helped me understand how powerful visualisation can be in revealing patterns that may not be obvious from raw values alone. Personally, I feel that the line-and-point plot in [Section 3.6b](#b-line-and-point-plot) revealed something that a column of numbers would have hidden, as it clearly shows the trend of on-time departures by airport and scheduled departure hour. If I only looked at the table, I might have noticed some individual values, but the overall downward trend from morning to evening would not have stood out as clearly. 

#### Aiman
When I first started doing the tasks, I was able to understand what the on-time percentages meant but did not see how it tied with airlines and airports. When we starting on tasks 3.7 to 3.8 is when I started seeing how United airline sat above the equality line that showed its actual performance relative to the difficult Newark heavy schedule. I was surprised by how ExpressJet and United share similar airport-hour circumstance despite having observed the rates diverging by nearly 10%. This experience made me appreciate that visualization is very much part of the analysis and can contribute to its effectiveness in realising the data.

#### Subhashini

Using a chart instead of a table for the `3.8` bubble plot showcased the accuracy of the expected on-time departure percentages for each airline, and how they compare to the observed on-time departure percentages. The bubble plot made it easy to identify which airlines were overperforming or underperforming relative to their expected values, and also highlighted the relationship between the airport and punctuality. A table with the same data would have been less clear and would not have provided the same visual insights into the performance of each airline in relation to their expected on-time departure rates.

#### Jabar

Approaching this lab, I thought that a simple on-time ranking would be a straighforward measure of airline quality. Section 3.4 stacked bar chart was the turning point in my perspective. The ranking was clearly influenced by airport effects that I had not taken into account when I saw how much United relied on Newark.

#### Min Xuan

Before this lab, I thought a simple ranking was enough to judge an airline. Making the charts showed me why that is wrong. The bubble plot in Section 3.8 changed my mind completely. It put United above the expected line, showing the airline actually performs better than normal for its tough Newark schedule. Meanwhile, ExpressJet falls far below the line even though it faces the exact same airport problems. A normal table of percentages hides this entirely. Seeing the dots mapped out proved you have to visualise the bad schedules to see how well an airline really runs.

## AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Useful AI Response

**In the response :**
```r
#| eval: false
And what is fct_reorder()?
fct_reorder() is a function from the forcats package (part of the tidyverse) that reorders the levels of a factor based on the values of another variable. This is especially useful in ggplot2 charts — without it, categorical variables like airline codes or airport names appear in arbitrary alphabetical order instead of a meaningful order.
```
In multiple occasions where the team was uncertain of how a specific function should be used and formatted, AI was able to help understand with much each and even clearing our doubts regarding multiple questions that the team had. This greatly helped the team complete each tasks and clear our doubts.


### Suboptimal AI Response


It is difficult to point out a sub-optimal response this week as the group took a more manual approach of referring to the lab sheet and tackling one task at a time together instead of prompting the AI at each stage of each task. 

However, if there was a need to force out a sub-optimal response of the team's choice, it would be the fact that in most cases, AI would assume fully that it already knows what the full dataset and result would look like within it's answers which may vary from the actual results that the team has. This could be seen from page 11 through 14 (of the AI_log.pdf) when the team needed help in approaching a question followed by asking if answer that we finally derived is on the right track.


### Reflection

The team found AI broadly useful for explaining unfamiliar functions, theory questions, and formatting areas where the AI's general knowledge is much better than ours. However, blindly copy pasting the questions, even with guidelines on the limitations of packages, is not always reliable and in-line with what is being taught to us, thus making the team refer to notes to do it in a manual way whilst using AI to check for formatting and correctness. As always, AI proves itself as a tool that can get the job done, but not always in the way one would intend for it to be.
