---
title: "ASEAN-Lab"
format: html
---

Assigned Role 
1. Coder - Quan Yi
2. Navigator - Shi Min
3. Verifier - Subhashini
4. Constructive Critic + Reporter - Jabar
5. AI Liaison - Aiman

```{r}
#| label: setup-meat
#| message: false
library(tidyverse)
meat <- read_csv("meat.csv")
```



Prediction:

a.
country: Brunei Darussalam, Malaysia, Singapore, China, Japan, Korea

animal: beef and buffalo, chicken, pig, cow

year: 2021, 2022, 2023

production: 4.26741, 4.33234, 4.68384

b.
country: 6
animal: 4
year: 3

c.
nrow(meat) -> 160
ncol(meat) -> 4

## 3.1 Inspecting the tibble

### Task 3.1.2
```{r}
#| label: meat-reveal

class(meat$country)
class(meat$animal)
class(meat$year)
class(meat$production)

nrow(meat)
ncol(meat)

length(unique(meat$country))
length(unique(meat$animal))
length(unique(meat$year))

head(meat)
tail(meat)
```

#### VERIFIER CHECK 3.1

Independently re-ran the same checks to confirm coder's output.

```{r}
#| label: verifier-classes

class(meat$country)    # Expected: "character" ✅
class(meat$animal)     # Expected: "character" ✅
class(meat$year)       # Expected: "numeric" ✅
class(meat$production) # Expected: "numeric" ✅

nrow(meat)  # Expected: 160 ✅
ncol(meat)  # Expected: 4 ✅

length(unique(meat$country))  # Expected: 10 ✅
length(unique(meat$animal))   # Expected: 4 ✅
length(unique(meat$year))     # Expected: 4 ✅

# head() and tail() outputs match coder's rendered results ✅
```

## 3.2 Handle Missing Values

```{r}
#| label: handle-missing-values

any(is.na(meat))
mask <- is.na(meat$production)

meat[mask, c("country", "animal", "year")]
```

In the rows where production is `NA`, the countries are always **Cambodia**, for **mutton and goat** and for the years 2018 to 2021



```{r}
#| label: total-production

sum(meat$production)
sum(meat$production, na.rm = TRUE)

```

As there is a `NA` value in the production column, it will return `NA`. 

By adding `na.rm = TRUE`, it will remove the rows with NA values and add the sum, therefore returning a value of 83937.52

When we use `na.rm = TRUE`, we assume that there are no production for that country for the animal product in that year and is thus `NA`. 

This assumption is valid, as since Cambodia did not have any production for mutton and goat through the years 2018 - 2021. Therefore we can assume that they do not have such produce in Cambodia

#### VERIFIER CHECK 3.2

Used `meat$production` specifically (coder used `any(is.na(meat))` on the whole tibble). Both return `TRUE` since the only NAs are in the production column.

```{r}
#| label: verifier-missing-values

any(is.na(meat$production))  # Expected: TRUE ✅

# Confirmed the same 4 NA rows: Cambodia, mutton and goat, years 2018–2021
meat[is.na(meat$production), c("country", "animal", "year")]  # ✅

sum(meat$production)              # Expected: NA ✅
sum(meat$production, na.rm = TRUE)  # Expected: 83937.52 ✅
```

## 3.3 Country Comparison

### Task 3.3.1
```{r}
#| label: compare-countries

poultry_2020 <- meat[meat$animal == "poultry" & meat$year == 2020, ]

max_poultry <- max(poultry_2020$production, na.rm = TRUE)

poultry_2020$country[poultry_2020$production == max_poultry]

```

### Task 3.3.2

```{r}
#| label: find-second-highest

ranks <- min_rank(desc(poultry_2020$production))
poultry_2020$country[ranks == 2]
```

### Task 3.3.3

```{r}
#| label: poultry-to-pig

pig_2020 <- meat[meat$animal == "pig" & meat$year == 2020, ]

identical(poultry_2020$country, pig_2020$country)

sum(poultry_2020$production > pig_2020$production, na.rm = TRUE)

```

#### VERIFIER CHECK 3.3

Used alternative approaches to independently confirm coder's results.

```{r}
#| label: verifier-country-comparison

# Task 3.3.1 – used which.max() instead of max(); same answer
poultry_2020$country[which.max(poultry_2020$production)]  # Expected: "Indonesia" ✅

# Task 3.3.2 – used order() instead of min_rank(); same answer
ranked <- poultry_2020[order(-poultry_2020$production), ]
ranked$country[2]  # Expected: "Thailand" ✅

# Task 3.3.3 – confirmed row order match and count
identical(poultry_2020$country, pig_2020$country)           # Expected: TRUE ✅
sum(poultry_2020$production > pig_2020$production)          # Expected: 6 ✅
```

## 3.4 Construct Factors

### Task 3.4.1

```{r}
#| label: assigning-animal-factor
meat$animal_factor <- factor(meat$animal)

is.factor(meat$animal_factor)

meat
```

### Task 3.4.2
```{r}
#| label: looking-at-factor

levels(meat$animal_factor)

identical(min_rank(levels(meat$animal_factor)), 1:length(levels(meat$animal_factor)))
```

### Task 3.4.3
It is not necessary for the meat to be an ordered factor, as there is no need for a comparison of e.g. poultry > pig.

### Task 3.4.4
```{r}
meat$color <- fct_collapse(
  meat$animal_factor,
  red   = c("beef and buffalo", "mutton and goat", "pig"),
  white = "poultry"
)
levels(meat$color)
```

#### VERIFIER CHECK 3.4

```{r}
#| label: verifier-factors

# Task 3.4.1 – confirmed animal_factor is a factor
is.factor(meat$animal_factor)  # Expected: TRUE ✅

# Task 3.4.2 – confirmed levels are in alphabetical order using two approaches
levels(meat$animal_factor)
# Expected: "beef and buffalo" "mutton and goat" "pig" "poultry" ✅

identical(
  levels(meat$animal_factor),
  levels(meat$animal_factor)[order(levels(meat$animal_factor))]
)  # Expected: TRUE — confirms alphabetical ordering ✅

identical(
  min_rank(levels(meat$animal_factor)),
  seq_along(levels(meat$animal_factor))
)  # Expected: TRUE ✅

# Task 3.4.4 – confirmed color collapse levels
levels(meat$color)  # Expected: "red" "white" ✅
```

## 3.5 Adding a Country-Code Column

### Task 3.5.1
```{r}
meat$iso3 <- recode_values(
  meat$country,
  "Brunei Darussalam" ~ "BRN",
  "Cambodia"          ~ "KHM",
  "Indonesia"         ~ "IDN",
  "Lao PDR"           ~ "LAO",
  "Malaysia"          ~ "MYS",
  "Myanmar"           ~ "MMR",
  "Philippines"       ~ "PHL",
  "Singapore"         ~ "SGP",
  "Thailand"          ~ "THA",
  "Viet Nam"          ~ "VNM"
)
```

```{r}
#| label: confirmation

any(is.na(meat$iso3))       

length(unique(meat$country))
length(unique(meat$iso3))
```

#### VERIFIER CHECK 3.5

```{r}
#| label: verifier-iso3

# Confirmed no NAs in iso3 column after assignment
any(is.na(meat$iso3))  # Expected: FALSE ✅

# Confirmed 1-to-1 mapping: 10 unique countries, 10 unique iso3 codes
length(unique(meat$country))  # Expected: 10 ✅
length(unique(meat$iso3))     # Expected: 10 ✅

# Cross-verified each country maps to the correct iso3 code
expected <- c(
  "Brunei Darussalam" = "BRN", "Cambodia" = "KHM", "Indonesia" = "IDN",
  "Lao PDR" = "LAO", "Malaysia" = "MYS", "Myanmar" = "MMR",
  "Philippines" = "PHL", "Singapore" = "SGP", "Thailand" = "THA",
  "Viet Nam" = "VNM"
)
all(meat$iso3 == expected[meat$country])  # Expected: TRUE ✅
```


## AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Suboptimal AI Response

During Section 3.5, AI suggested the following line to verify the iso3 mapping:

```r
length(unique(paste(meat$country, meat$iso3)))
```

This response was suboptimal because `paste()` was not within the list of permitted functions specified for this task. Despite being clearly informed of the constraints upfront, the AI defaulted to its own preferred method rather than working within the given boundaries. This highlights how AI tools can fail to fully adhere to specified restrictions even when they have been explicitly stated.

The issue was caught when the coder was typing out the code and the team subsequently prompted the AI to do a full check on all of its output. The AI was able to identify the violation itself and provide a corrected alternative that used only permitted functions — which is what the team adopted as the final solution.

### Reflection

When using AI tools, it is important to **clearly specify the parameters of the task**, including any constraints on which functions or methods are permitted. Being explicit about what is and is not allowed helps the AI give more accurate and appropriate responses. Additionally, performing a final full-output review by prompting the AI to self-check its own suggestions is a useful practice, as demonstrated here — the AI was able to catch and correct its own mistake when prompted to do so.
