---
title: "ASEAN-Lab"
author: "Team Orchid"
date: "2026-05-13"
format:
  html:
    toc: true
---

## Assigned Roles

The table below lists the assigned roles for each team member in this lab:

| Member | Role | 
|---|---|
| Quan Yi | Coder |
| Shi Min | Navigator |
| Aiman | AI Liaison |
| Subhashini | Verifier |
| Jabar | Reporter |
| Min Xuan | Constructive Critic |

## Setup

```{r}
#| label: setup-meat
#| message: false
library(tidyverse)
meat <- read_csv("meat.csv")
```

The above code is used to import the `meat.csv` dataset into R as a tibble for analysis. The `tidyverse` package is loaded to provide access to functions for data manipulation and analysis.

## Task 3.1 Inspecting meat tibble

In this section, we first examine the structure of the meat dataset before running deeper analysis. Each member makes predictions about the column classes, distinct values, and dataset dimensions, then we compare those predictions against the actual results from R.

Each member predicts:

1. The class of each column (`country`, `animal`, `year`, `production`)
2. The number of distinct values in `country`, `animal`, and `year`
3. The dataset dimensions: `nrow(meat)` and `ncol(meat)`

---

### Task 3.1.1 - Member's predictions for the `meat` dataset:

**Class Prediction**

| Member | country | animal | year | production |
|---|---|---|---|---|
| Quan Yi | character (e.g., "Brunei Darussalam", "Malaysia", "Singapore", "China", "Japan", "Korea") | character (e.g., beef and buffalo, chicken, pig, cow) | numeric (e.g., 2021, 2022, 2023) | numeric (e.g., 4.26741, 4.33234, 4.68384) |
| Shi Min | character | character | date | double |
| Aiman | character | character | integer | numeric |
| Subhashini | character (e.g., Brunei Darussalam, Indonesia, Cambodia, Singapore, Malaysia, Vietnam, Thailand) | character (e.g., pig, chicken, beef) | numeric (e.g., 2018–2021) | numeric (e.g., 60.8341) |
| Jabar | character | factor | integer | numeric |
| Min Xuan | character | character | numeric | numeric |

**Number of Distinct Values**

| Member | country | animal | year |
|---|---:|---:|---:|
| Quan Yi | 6 | 4 | 3 |
| Shi Min | 10 | 20 | 4 |
| Aiman | 3 | 4 | 4 |
| Subhashini | 10 | 7 | 4 |
| Jabar | 10 | 4 | 4 |
| Min Xuan | 10 | 4 | 4 |

**Dataset Dimensions**

| Member | `nrow(meat)` | `ncol(meat)` |
|---|---:|---:|
| Quan Yi | 160 | 4 |
| Shi Min | 1000 | 10 |
| Aiman | 20 | 15 |
| Subhashini | 100 | 10 |
| Jabar | 110 | 8 |
| Min Xuan | 160 | 4 |


### Task 3.1.2

```{r}
#| label: meat-class-reveal
class(meat$country)
class(meat$animal)
class(meat$year)
class(meat$production)
```

`class()` returns the data type of the specified column. The returned classes indicate that `year` and `production` are "numeric" but `country` and `animal` are "character". 

```{r}
#| label: meat-dimension-reveal
nrow(meat)
ncol(meat)
```

`nrow()` and `ncol()` return the number of rows and columns.

```{r}
#| label: meat-uniques-reveal
length(unique(meat$country))
length(unique(meat$animal))
length(unique(meat$year))
```

`unique()` retrieves all distinct values in a column, and then counts them with `length()` to find the number of unique countries, animals, and years in the dataset.

Majority of teammates correctly predicted the class `country` and `animal` as "character".  Most of us were surprised by the dataset's dimensions and the number of different ASEAN nations. In addition, some almost accurately anticipated the exact `nrow` and `ncol`.

### Task 3.1.3

```{r}
#| label: meat-overview-reveal
head(meat)
tail(meat)
```

`head()` and `tail()` show first and last 6 rows, which helps to get a quick overview of the data structure and content.

#### VERIFIER CHECK 3.1

Independently re-ran the same checks to confirm coder's output.

```{r}
#| label: verifier-classes
class(meat$country) # Expected: "character" ✅
class(meat$animal) # Expected: "character" ✅
class(meat$year) # Expected: "numeric" ✅
class(meat$production) # Expected: "numeric" ✅

nrow(meat) # Expected: 160 ✅
ncol(meat) # Expected: 4 ✅

length(unique(meat$country)) # Expected: 10 ✅
length(unique(meat$animal)) # Expected: 4 ✅
length(unique(meat$year)) # Expected: 4 ✅

# head() and tail() outputs match coder's rendered results ✅
```

## Task 3.2 Handle Missing Values

This section checks whether the dataset contains missing production values and identifies where they occur. We also explore how missing values affect summary calculations and discuss whether removing them with na.rm = TRUE is a reasonable assumption.

### Task 3.2.1, 3.2.2
```{r}
#| label: handle-missing-values
any(is.na(meat))
mask <- is.na(meat$production)
```

`any(is.na(meat))` checks if any value in the tibble is `NA` to return `TRUE`.

`is.na(meat$production)` returns a logical vector where each element is TRUE if the corresponding value is `NA`.

```{r}
#| label: production-logical-mask
meat[mask, c("country", "animal", "year")]
```

`meat[mask, c("country", "animal", "year")]` subsets the tibble using the mask to show only rows where production is missing.

It was observed that in the rows where production is `NA`, the countries are always **Cambodia**, for **mutton and goat** and from the year 2018 to 2021

### Task 3.2.3

```{r}
#| label: total-production
sum(meat$production)
sum(meat$production, na.rm = TRUE)
```

As there is a missing value in the production column, the sum operation cannot be completed and returns `NA`. 

By adding `na.rm = TRUE`, it will ignore the rows with missing/NA values and perform the sum operation, returning a value of `83937.52`

### Task 3.2.4

When we use `na.rm = TRUE`, we assume that there are no production for that country for the animal product in that year and is thus `NA`. 

This assumption is valid because Cambodia recorded zero production for mutton and goat from 2018 to 2021. Therefore, we can assume that they do not have such produce in Cambodia and treat these missing values as a lack of production.

#### VERIFIER CHECK 3.2

Used `meat$production` specifically (coder used `any(is.na(meat))` on the whole tibble). Both return `TRUE` since the only NAs are in the production column.

```{r}
#| label: verifier-missing-values
any(is.na(meat$production)) # Expected: TRUE ✅

# Confirmed the same 4 NA rows: Cambodia, mutton and goat, years 2018–2021
meat[is.na(meat$production), c("country", "animal", "year")] # ✅

sum(meat$production) # Expected: NA ✅
sum(meat$production, na.rm = TRUE) # Expected: 83937.52 ✅
```

## Task 3.3 Country Comparison

In this section, we compare meat production across ASEAN countries in 2020. The focus is on identifying the highest and second-highest poultry producers, then comparing poultry and pig production across countries.

### Task 3.3.1

```{r}
#| label: compare-countries
poultry_2020 <- meat[meat$animal == "poultry" & meat$year == 2020, ]

max_poultry <- max(poultry_2020$production, na.rm = TRUE)

poultry_2020$country[poultry_2020$production == max_poultry]
```

The first line keeps only the rows for poultry in 2020 and subsets the tibble using compound logical condition(both criteria must be TRUE).

The maximum production value is found with `max()`. The second line determines which nation's production is equal to that maximum using a logical index.

The final line uses a logical index `(production == max_poultry)` to find the row(s) where the production matches that maximum value, and then returns the corresponding country name

### Task 3.3.2

```{r}
#| label: find-second-highest
ranks <- min_rank(desc(poultry_2020$production))
poultry_2020$country[ranks == 2]
```

`desc()` reverses the order so the largest value ranks first.

`min_rank()` assigns integer ranks. Subsetting where rank equals 2.

This helps to find the country with the second-highest poultry production in 2020

### Task 3.3.3

Using the same logic in **Task 3.3.1**, we create `pig_2020`

```{r}
#| label: poultry-to-pig-comparison
pig_2020 <- meat[meat$animal == "pig" & meat$year == 2020, ]

identical(poultry_2020$country, pig_2020$country)
```

`identical()` confirms the country vectors are in the same order before doing an element wise comparison returning `TRUE`.

```{r}
#| label: poultry-to-pig-sum
sum(poultry_2020$production > pig_2020$production, na.rm = TRUE)
```

`poultry_2020$production > pig_2020$production` compares each country's poultry vs pig production row by row.

`sum()` counts the `TRUE` and by summing the logical vector, we calculate the total number of countries that produced more poultry than pig.

#### VERIFIER CHECK 3.3

Used alternative approaches to independently confirm coder's results.

```{r}
#| label: verifier-country-comparison
# Task 3.3.1 – used which.max() instead of max(); same answer
poultry_2020$country[which.max(poultry_2020$production)] # Expected: "Indonesia" ✅

# Task 3.3.2 – used order() instead of min_rank(); same answer
ranked <- poultry_2020[order(-poultry_2020$production), ]
ranked$country[2] # Expected: "Thailand" ✅

# Task 3.3.3 – confirmed row order match and count
identical(poultry_2020$country, pig_2020$country) # Expected: TRUE ✅
sum(poultry_2020$production > pig_2020$production) # Expected: 6 ✅
```

`poultry_2020$country[which.max(poultry_2020$production)]` directly returns the index of the maximum value. 

`ranked <- poultry_2020[order(-poultry_2020$production), ]` returns indices that would sort in descending order then takes row two giving the second highest country.

## Task 3.4 Construct Factors

This section converts the animal column into a factor so that the categorical meat types can be handled more appropriately in R. We also inspect the factor levels and group the meat categories into broader “red” and “white” meat classifications.

### Task 3.4.1

```{r}
#| label: assigning-animal-factor
meat$animal_factor <- factor(meat$animal)

is.factor(meat$animal_factor)

meat
```

`factor(meat$animal)` converts the character column animal into a factor and stores it in a new column. 

`is.factor(meat$animal_factor)` confirms it worked, returning `TRUE`.

### Task 3.4.2

```{r}
#| label: looking-at-factor-levels
levels(meat$animal_factor)
```

`levels(meat$animal_factor)` returns the factor's levels in their stored order. 

```{r}
#| label: looking-at-factor-identical
identical(
  min_rank(levels(meat$animal_factor)),
  1:length(levels(meat$animal_factor))
)
```

`min_rank(levels(meat$animal_factor)` assigns a numerical rank to the levels based on their alphabetical position

`1:length(levels(meat$animal_factor)))` generates a simple sequence of integers from 1 to the total number of levels.

This check confirms that they are alphabetical by using `identical()` as the rank matches the sequence.

### Task 3.4.3

It is not necessary for the meat to be an ordered factor, as there is no need for a comparison of e.g. poultry > pig.

### Task 3.4.4

```{r}
#| label: collapsing-color
meat$color <- fct_collapse(
  meat$animal_factor,
  red = c("beef and buffalo", "mutton and goat", "pig"),
  white = "poultry"
)
levels(meat$color)
```

`fct_collapse()` merges multiple factor levels into new ones. In our case, we mapped "beef and buffalo", "mutton and goat" and "pig" to "red" meat and the level "poultry" to "white"

We then confirm that there are only "red" and "white" meat in `animal_factor` 

#### VERIFIER CHECK 3.4

```{r}
#| label: verifier-factors
# Task 3.4.1 – confirmed animal_factor is a factor
is.factor(meat$animal_factor) # Expected: TRUE ✅

# Task 3.4.2 – confirmed levels are in alphabetical order using two approaches
levels(meat$animal_factor)
# Expected: "beef and buffalo" "mutton and goat" "pig" "poultry" ✅

identical(
  levels(meat$animal_factor),
  levels(meat$animal_factor)[order(levels(meat$animal_factor))]
) # Expected: TRUE — confirms alphabetical ordering ✅

identical(
  min_rank(levels(meat$animal_factor)),
  seq_along(levels(meat$animal_factor))
) # Expected: TRUE ✅

# Task 3.4.4 – confirmed color collapse levels
levels(meat$color) # Expected: "red" "white" ✅
```

## Task 3.5 Adding a Country-Code Column

In this section, we add ISO 3-letter country codes to the dataset by mapping each ASEAN country name to its corresponding code. We then verify that the new `iso3` column has no missing values and correctly represents all ten countries.

### Task 3.5.1

```{r}
#| label: create-iso3
meat$iso3 <- recode_values(
  meat$country,
  "Brunei Darussalam" ~ "BRN",
  "Cambodia" ~ "KHM",
  "Indonesia" ~ "IDN",
  "Lao PDR" ~ "LAO",
  "Malaysia" ~ "MYS",
  "Myanmar" ~ "MMR",
  "Philippines" ~ "PHL",
  "Singapore" ~ "SGP",
  "Thailand" ~ "THA",
  "Viet Nam" ~ "VNM"
)
```

`recode_values()` helps to create a new column (iso3) by mapping full country names to the ISO3 codes using tilde-based formulas.

### Task 3.5.2

```{r}
#| label: confirmation-iso3
any(is.na(meat$iso3))

length(unique(meat$country))
length(unique(meat$iso3))
nrow(unique(meat[, c("country", "iso3")]))
```

We verify that there are no missing values as `any(is.na(meat$iso3))` returned `FALSE` and that there are ten unique name-plus-code pairs

#### VERIFIER CHECK 3.5

```{r}
#| label: verifier-iso3
# Confirmed no NAs in iso3 column after assignment
any(is.na(meat$iso3)) # Expected: FALSE ✅

# Confirmed 1-to-1 mapping: 10 unique countries, 10 unique iso3 codes
length(unique(meat$country)) # Expected: 10 ✅
length(unique(meat$iso3)) # Expected: 10 ✅

# Cross-verified each country maps to the correct iso3 code
expected <- c(
  "Brunei Darussalam" = "BRN",
  "Cambodia" = "KHM",
  "Indonesia" = "IDN",
  "Lao PDR" = "LAO",
  "Malaysia" = "MYS",
  "Myanmar" = "MMR",
  "Philippines" = "PHL",
  "Singapore" = "SGP",
  "Thailand" = "THA",
  "Viet Nam" = "VNM"
)
all(meat$iso3 == expected[meat$country]) # Expected: TRUE ✅
```

## AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Suboptimal AI Response

During Section 3.5, AI suggested the following line to verify the iso3 mapping:

```r
length(unique(paste(meat$country, meat$iso3)))
```

This response was suboptimal because `paste()` was not within the list of permitted functions specified for this task. Despite being clearly informed of the constraints upfront, the AI defaulted to its own preferred method rather than working within the given boundaries. This highlights how AI tools can fail to fully adhere to specified restrictions even when they have been explicitly stated.

The issue was caught when the coder was typing out the code and the team subsequently prompted the AI to do a full check on all of its output. The AI was able to identify the violation itself and provide a corrected alternative that used only permitted functions — which is what the team adopted as the final solution.

### Reflection

When using AI tools, it is important to **clearly specify the parameters of the task**, including any constraints on which functions or methods are permitted. Being explicit about what is and is not allowed helps the AI give more accurate and appropriate responses. Additionally, performing a final full-output review by prompting the AI to self-check its own suggestions is a useful practice, as demonstrated here — the AI was able to catch and correct its own mistake when prompted to do so.
