---
title: "ASEAN-Lab"
format: html
---

Assigned Role 
1. Coder - Quan Yi
2. Navigator - Shi Min
3. Verifier - Subhashini
4. Constructive Critic - MinXuan
5. AI Liaison - Aiman
6. Reporter - Jabar

```{r}
#| label: setup-meat
#| message: false
library(tidyverse)
meat <- read_csv("meat.csv")
```

This document contains each group member's individual predictions for the `meat` dataset, made **before** running any code or consulting AI.

Each member predicts:

- **a.** The class of each column (`country`, `animal`, `year`, `production`)
- **b.** The number of distinct values in `country`, `animal`, and `year`
- **c.** The dataset dimensions: `nrow(meat)` and `ncol(meat)`

---

Quan Yi's Predictions

a. Class Prediction

- **country:** character (e.g., "Brunei Darussalam", "Malaysia", "Singapore", "China", "Japan", "Korea")
- **animal:** character (e.g., beef and buffalo, chicken, pig, cow)
- **year:** numeric (e.g., 2021, 2022, 2023)
- **production:** numeric (e.g., 4.26741, 4.33234, 4.68384)

b. Number of Distinct Values

- **country:** 6
- **animal:** 4
- **year:** 3

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 160
- `ncol(meat)` → Predicted: 4

---

Shi Min's Predictions

a. Class Prediction

- **country:** character
- **animal:** character
- **year:** date
- **production:** double

b. Number of Distinct Values

- **country:** 10
- **animal:** 20
- **year:** 4

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 1000
- `ncol(meat)` → Predicted: 10

---

Aiman's Predictions

a. Class Prediction

- **country:** character
- **animal:** character
- **year:** integer
- **production:** numeric

b. Number of Distinct Values

- **country:** 3
- **animal:** 4
- **year:** 4

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 20
- `ncol(meat)` → Predicted: 15

---

Subhashini's Predictions

a. Class Prediction

- **country:** character (e.g., Brunei Darussalam, Indonesia, Cambodia, Singapore, Malaysia, Vietnam, Thailand)
- **animal:** character (e.g., pig, chicken, beef)
- **year:** numeric (e.g., 2018–2021)
- **production:** numeric (e.g., 60.8341)

b. Number of Distinct Values

- **country:** 10
- **animal:** 7
- **year:** 4

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 100
- `ncol(meat)` → Predicted: 10

---

Jabar's Predictions

a. Class Prediction

- **country:** character
- **animal:** factor
- **year:** integer
- **production:** numeric

b. Number of Distinct Values

- **country:** 10
- **animal:** 4
- **year:** 4

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 110
- `ncol(meat)` → Predicted: 8

---

Min Xuan's Predictions

a. Class Prediction

- **country:** character
- **animal:** character
- **year:** numeric
- **production:** numeric

b. Number of Distinct Values

- **country:** 10
- **animal:** 4
- **year:** 4

c. Dataset Dimensions

- `nrow(meat)` → Predicted: 160
- `ncol(meat)` → Predicted: 4

## 3.1 Inspecting the tibble

### Task 3.1.2
```{r}
#| label: meat-reveal
class(meat$country)
class(meat$animal)
class(meat$year)
class(meat$production)

nrow(meat)
ncol(meat)

length(unique(meat$country))
length(unique(meat$animal))
length(unique(meat$year))

head(meat)
tail(meat)
```

data type of every column is returned by class(). 
`year` and `production` are "numeric" but nation and animal are "character".

`nrow()` and `ncol()` return the number of rows and columns.

`unique()` retrieves all distinct values in a column.
wrapping it in `length()` and counts them.

`head()` and `tail()` show first and last 6 rows.

Majority of teammates correctly predicted the class `country` and `animal` as "character".  Most of us were surprised by the dataset's dimensions and the number of different ASEAN nations. In addition, some almost accurately anticipated the exact `nrow` and `ncol`.



#### VERIFIER CHECK 3.1

Independently re-ran the same checks to confirm coder's output.

```{r}
#| label: verifier-classes
class(meat$country) # Expected: "character" ✅
class(meat$animal) # Expected: "character" ✅
class(meat$year) # Expected: "numeric" ✅
class(meat$production) # Expected: "numeric" ✅

nrow(meat) # Expected: 160 ✅
ncol(meat) # Expected: 4 ✅

length(unique(meat$country)) # Expected: 10 ✅
length(unique(meat$animal)) # Expected: 4 ✅
length(unique(meat$year)) # Expected: 4 ✅

# head() and tail() outputs match coder's rendered results ✅
```


## 3.2 Handle Missing Values

```{r}
#| label: handle-missing-values
any(is.na(meat))
mask <- is.na(meat$production)

meat[mask, c("country", "animal", "year")]
```

`is.na(meat$production)` creates a logical TRUE.
`any(is.na(meat))` checks if any value in the tibble is `NA` to return `TRUE`.
`meat[mask, c("country", "animal", "year")]` subsets the tibble using the mask to show only rows where production is missing.

In the rows where production is `NA`, the countries are always **Cambodia**, for **mutton and goat** and for the years 2018 to 2021



```{r}
#| label: total-production
sum(meat$production)
sum(meat$production, na.rm = TRUE)
```

As there is a `NA` value in the production column, it will return `NA`. 

By adding `na.rm = TRUE`, it will remove the rows with NA values and add the sum, therefore returning a value of 83937.52

When we use `na.rm = TRUE`, we assume that there are no production for that country for the animal product in that year and is thus `NA`. 

This assumption is valid, as since Cambodia did not have any production for mutton and goat through the years 2018 - 2021. Therefore we can assume that they do not have such produce in Cambodia

#### VERIFIER CHECK 3.2

Used `meat$production` specifically (coder used `any(is.na(meat))` on the whole tibble). Both return `TRUE` since the only NAs are in the production column.

```{r}
#| label: verifier-missing-values
any(is.na(meat$production)) # Expected: TRUE ✅

# Confirmed the same 4 NA rows: Cambodia, mutton and goat, years 2018–2021
meat[is.na(meat$production), c("country", "animal", "year")] # ✅

sum(meat$production) # Expected: NA ✅
sum(meat$production, na.rm = TRUE) # Expected: 83937.52 ✅
```

## 3.3 Country Comparison

### Task 3.3.1
```{r}
#| label: compare-countries
poultry_2020 <- meat[meat$animal == "poultry" & meat$year == 2020, ]

max_poultry <- max(poultry_2020$production, na.rm = TRUE)

poultry_2020$country[poultry_2020$production == max_poultry]
```

This keeps only the rows for poultry in 2020 and subsets the tibble using compound logical condition(both criteria must be TRUE).

The maximum production value is found with `max()`. The second line determines which nation's production is equal to that maximum using a logical index.



### Task 3.3.2

```{r}
#| label: find-second-highest
ranks <- min_rank(desc(poultry_2020$production))
poultry_2020$country[ranks == 2]
```

`desc()` reverses the order so the largest value ranks first.
`min_rank()` assigns integer ranks. Subsetting where rank equals 2.



### Task 3.3.3

```{r}
#| label: poultry-to-pig
pig_2020 <- meat[meat$animal == "pig" & meat$year == 2020, ]

identical(poultry_2020$country, pig_2020$country)

sum(poultry_2020$production > pig_2020$production, na.rm = TRUE)
```

`identical()` confirms the country vectors are in the same order before doing an element wise comparison returning `TRUE`.

`poultry_202$production > pig_202$production` compares each country's poultry vs pig production row by row.
`sum()` counts the `TRUE`



#### VERIFIER CHECK 3.3

Used alternative approaches to independently confirm coder's results.

```{r}
#| label: verifier-country-comparison
# Task 3.3.1 – used which.max() instead of max(); same answer
poultry_2020$country[which.max(poultry_2020$production)] # Expected: "Indonesia" ✅

# Task 3.3.2 – used order() instead of min_rank(); same answer
ranked <- poultry_2020[order(-poultry_2020$production), ]
ranked$country[2] # Expected: "Thailand" ✅

# Task 3.3.3 – confirmed row order match and count
identical(poultry_2020$country, pig_2020$country) # Expected: TRUE ✅
sum(poultry_2020$production > pig_2020$production) # Expected: 6 ✅
```

`poultry_2020$country[which.max(poultry_2020$production)]` directly returns the index of the maximum value. 

`ranked <- poultry_2020[order(-poultry_2020$production), ]` returns indices that would sort in descending order then takes row two giving the second highest country.



## 3.4 Construct Factors

### Task 3.4.1

```{r}
#| label: assigning-animal-factor
meat$animal_factor <- factor(meat$animal)

is.factor(meat$animal_factor)

meat
```

`factor(meat$animal)` converts the character column animal into a factor and stores it in a new column. 

`is.factor(meat$animal_factor)` confirms it worked, returning `TRUE`.

### Task 3.4.2
```{r}
#| label: looking-at-factor
levels(meat$animal_factor)

identical(
  min_rank(levels(meat$animal_factor)),
  1:length(levels(meat$animal_factor))
)
```

`levels(meat$animal_factor)` returns the factor's levels in their stored order. 

`identical(min_rank(levels(meat$animal_factor)), 1:length(levels(meat$animal_factor)))` check confirms they are alphabetical.

### Task 3.4.3
It is not necessary for the meat to be an ordered factor, as there is no need for a comparison of e.g. poultry > pig.

### Task 3.4.4
```{r} 
#| label: collapsing-color
meat$color <- fct_collapse(
  meat$animal_factor,
  red = c("beef and buffalo", "mutton and goat", "pig"),
  white = "poultry"
)
levels(meat$color)
```

`fct_collapse()` merges multiple factor levels into new ones.

#### VERIFIER CHECK 3.4

```{r}
#| label: verifier-factors
# Task 3.4.1 – confirmed animal_factor is a factor
is.factor(meat$animal_factor) # Expected: TRUE ✅

# Task 3.4.2 – confirmed levels are in alphabetical order using two approaches
levels(meat$animal_factor)
# Expected: "beef and buffalo" "mutton and goat" "pig" "poultry" ✅

identical(
  levels(meat$animal_factor),
  levels(meat$animal_factor)[order(levels(meat$animal_factor))]
) # Expected: TRUE — confirms alphabetical ordering ✅

identical(
  min_rank(levels(meat$animal_factor)),
  seq_along(levels(meat$animal_factor))
) # Expected: TRUE ✅

# Task 3.4.4 – confirmed color collapse levels
levels(meat$color) # Expected: "red" "white" ✅
```



## 3.5 Adding a Country-Code Column

### Task 3.5.1
```{r}
#| label: assign-iso3
meat$iso3 <- recode_values(
  meat$country,
  "Brunei Darussalam" ~ "BRN",
  "Cambodia" ~ "KHM",
  "Indonesia" ~ "IDN",
  "Lao PDR" ~ "LAO",
  "Malaysia" ~ "MYS",
  "Myanmar" ~ "MMR",
  "Philippines" ~ "PHL",
  "Singapore" ~ "SGP",
  "Thailand" ~ "THA",
  "Viet Nam" ~ "VNM"
)
```

```{r}
#| label: confirmation
any(is.na(meat$iso3))

length(unique(meat$country))
length(unique(meat$iso3))
```

#### VERIFIER CHECK 3.5

```{r}
#| label: verifier-iso3
# Confirmed no NAs in iso3 column after assignment
any(is.na(meat$iso3)) # Expected: FALSE ✅

# Confirmed 1-to-1 mapping: 10 unique countries, 10 unique iso3 codes
length(unique(meat$country)) # Expected: 10 ✅
length(unique(meat$iso3)) # Expected: 10 ✅

# Cross-verified each country maps to the correct iso3 code
expected <- c(
  "Brunei Darussalam" = "BRN",
  "Cambodia" = "KHM",
  "Indonesia" = "IDN",
  "Lao PDR" = "LAO",
  "Malaysia" = "MYS",
  "Myanmar" = "MMR",
  "Philippines" = "PHL",
  "Singapore" = "SGP",
  "Thailand" = "THA",
  "Viet Nam" = "VNM"
)
all(meat$iso3 == expected[meat$country]) # Expected: TRUE ✅
```


## AI Liaison Report

**Tool used:** Perplexity (Claude Sonnet 4.6)

### One Suboptimal AI Response

During Section 3.5, AI suggested the following line to verify the iso3 mapping:

```r
length(unique(paste(meat$country, meat$iso3)))
```

This response was suboptimal because `paste()` was not within the list of permitted functions specified for this task. Despite being clearly informed of the constraints upfront, the AI defaulted to its own preferred method rather than working within the given boundaries. This highlights how AI tools can fail to fully adhere to specified restrictions even when they have been explicitly stated.

The issue was caught when the coder was typing out the code and the team subsequently prompted the AI to do a full check on all of its output. The AI was able to identify the violation itself and provide a corrected alternative that used only permitted functions — which is what the team adopted as the final solution.

### Reflection

When using AI tools, it is important to **clearly specify the parameters of the task**, including any constraints on which functions or methods are permitted. Being explicit about what is and is not allowed helps the AI give more accurate and appropriate responses. Additionally, performing a final full-output review by prompting the AI to self-check its own suggestions is a useful practice, as demonstrated here — the AI was able to catch and correct its own mistake when prompted to do so.
